### 2022-12-06 experiments/superg/doc/paper/paper.tex
Start writing the paper.

\documentclass[conference]{IEEEtran}

\usepackage{cite}
\usepackage[pdftex]{graphicx}
% declare the path(s) where your graphic files are
\graphicspath{{images/}}

% Packages that are sometimes useful
%\usepackage{amsmath,amssymb,amsfonts}
%\usepackage{algorithmic}
%\usepackage{float}
%\usepackage{url}

\begin{document}

\title{Reviving the Lost Art of Combinator Graph Reduction}

\author{\IEEEauthorblockN{Daan de Graaf}
    \IEEEauthorblockA{\textit{Master's student Computer Engineering} \\
        \textit{TU Delft / Eindhoven University of Technology}\\
        Eindhoven, Netherlands \\
        d.j.a.degraaf@student.tudelft.nl / d.j.a.d.graaf@student.tue.nl}
}

\maketitle

\begin{abstract}
    % Rules of thumb
    % - 1 sentence background
    % - 1 sentence motive/problem
    % - 1 sentence objective
    % - 1 sentence approach/method
    % - 2 sentences support
    % - 1 sentence conclusion
    % - 1 sentence implication
    % - 1 sentence limitation
    This is the abstract.
\end{abstract}

\begin{IEEEkeywords}
    combinatory logic, graph reduction, lazy functional language, lazy evalution
\end{IEEEkeywords}

\section{Introduction}
% Why did I research this. Give the motive, usually a knowledge gap or problem
Lazy evaluation of programs is a powerful idea.
It allows a program to operate on infinitely large data structures, and if part of the computation turns out to be unnecessary, it can be much more efficient than a strict evaluation.
In practice however, delaying computation often introduces overhead, making languages based on lazy evaluation slower than their strict counterparts.
To compete with a strict implementation, a lazy runtime must maximize lazyness while minizing overhead.
It is a difficult balance, and over the years many different \textit{abstract machines} have been proposed\cite{kieburtz_g-machine_1985,fairbairn_tim_1987,burn_spineless_1988,koopman_fresh_1989,jones_implementing_nodate}.
One of the early approaches to implementating lazy functional languages is combinator graph reduction\cite{turner_new_1979}.

\textbf{TODO:} what is combinator reduction?

Combinator reduction was the runtime for David Turner's Miranda language\cite{turner_miranda_1985}, the popular choice for pure functional langauges at the time, and one of the few to be commercially supported.
Eventually the interest for Miranda and combinator graph reduction with it waned, and today Haskell is the dominant language.
Part of the reason may be that Miranda was closed source for 35 years, its source code was released only in 2020\cite{noauthor_open_2021}, but combinator reduction also deserves part of the blame.
Expressions are reduced in many small steps, so the per-step overhead greatly impacts runtime performance.
Every reduction step modifies the stack and usually also the graph, resulting in high memory traffic.
Another important issue is that the translation to combinators may result in much larger program.
For a lambda calculus expression of size $N$, the traditional bracket abstraction algorithm produces an output of $O(n^2)$ in the worst case, were other approaches such as supercombinators are linear.

These disadvantages have led to combinator graph reduction to be all but abandoned.
Recently however, a new algorithm for translation of lambda calculus expressions into combinatory logic has been proposed that is linear in the size of the input\cite{kiselyov_lambda_2018}.
While a clear improvement over previous translation strategies, it remains to be seen whether it can make the combinator reduction approach competitive with contemporary reduction engines.
This leads us to our first research question:

\textbf{RQ1:} How does Kiselyov's semantic translation compare to the classical bracket abstraction algorithm in terms of runtime performance of compiled programs?

We can take a set of benchmark programs and compile them with both algorithms, but to compare real-world execution time we also need a fast runtime to execute the compiled programs.

Alas, there is no performant combinator graph reduction engine on which we can execute programs compiled with Kiselyov's.
The now open sourced Miranda is a good starting point, but it is showing its age.
The 2020 port compiles cleanly on a recent GCC, but enabling optimizations is known to break the garbage collector.
Miranda's architecture is based on cells with a tag, head and tail as described in\cite{turner_new_1979}, but later work has shown that a tagless representation with direct pointers to code leads to better performance\cite{koopman_fresh_1989}.
In absence of an up to date reference on building a practical combinator reduction engine, we formlate the second research question:

\textbf{RQ2:} How is combinator graph reduction implemented efficiently on conventional hardware?

We evaluate the state of the art in combinator graph reduction, and build a fast implementation for a language that is simple but powerful enough to express our benchmark programs.
Using this engine, we can then answer RQ1.

% What did I research?
% - Research problem / question
%   * Definitions
%   * Choices
%   * Presuppositions
% - Why specifically this research question?
%   * Theoretical or societal relevance

% How did I research this?
% - Methods
% - Sub-questions

% How did I organize the report?
% - Structure overview

%NOTES:
%While initially a popular approach to lazy evaluation, combinator graph reduction fell out of fashion with the advent of new abstract machines based on supercombinators (G-machine, TIM).
%
%In the 1980s David Turner developed the language Miranda\cite{turner_miranda_1985}.
%At the time it was the popular choice for lazy functional programming, and one of few to be commercially supported.
%
%Lazy evaluation of programs is a powerful idea.
%It allows a program to operate on infinitely large data structures, and if part of the computation turns out to be unnecessary, it can be much more efficient than a strict evaluation.
%In practice however, delaying computation often introduces overhead, making languages based on lazy evaluation slower than their strict counterparts.
%To compete with a strict implementation, a lazy runtime must maximize lazyness while minizing overhead.
%
%Some of the early lazy functional languages were built on combinatory logic\cite{turner_new_1979,turner_miranda_1985}.
%Where the well-known lambda calculus has a concept of named variables that an evaluator must track and substitute values for in function application,
%in combinatory logic there are no bound variables, only a small .
%
%- Most important lazy functional language today is Haskell
%- It is built on top of the spineless tagless G-machine model
%- STG combines ideas from G-machine and TIM, both in the class of supercombinators
%- *What are supercombinators?*
%- Haskell inherits from Miranda, which was built on a very different model, combinator graph reduction
%- *What is Combinator Graph Reduction?*
%- Advantages of Combinators:
%* Small and fixed set of combinators. Reduction machine is simple, can be done in hardware.
%* Lambda bodies instantiated lazily, lazier than supercombinators
%- Disadvantages:
%* Reduction steps are small. Lot of transient memory traffic, high strain on garbage collector. Caching is also very coarse-grained.
%* Translation to combinators results in a larger program. (attack this with Kiselyov)
%- At the turn of the century, combinator graph reduction was all but abandoned.
%- Recently, a new translation strategy has been proposed for translating to combinators (cue Kiselyov).
%- After 35 years under a commercial license, Miranda was released under the open BSD license.
%- In this work we build on Miranda's design, incorporating the proposed semantic combinator translation algorithm and changes to the execution engine proposed in later papers.
%
%Research Questions:
%1. How is combinator graph reduction implemented efficiently on conventional hardware?
%2. How does Kiselyov's semantic translation compare to the classical bracket abstraction algorithm in terms of runtime performance of compiled programs?

\section*{Kiselyov's algorithms}
Describes the strict and lazy translation algorithms in a practical manner.

\section*{Engine implementation}
\subsection*{Miranda}
\subsection*{TIGRE}

\section*{Conclusion}
% Repeat the main question / objective of your research

% Give the answer to the question

% Why is this the answer? Present the main arguments

% What are the implications of this answer?

% Suggestions for future research

% FINAL CHECK: Can the conclusion be read independently?

\bibliographystyle{IEEEtran}
\bibliography{zotero}

\end{document}

### 2022-12-06 experiments/superg/doc/paper/paper.tex
More on introduction.

@@
 To compete with a strict implementation, a lazy runtime must maximize lazyness while minizing overhead.
-It is a difficult balance, and over the years many different \textit{abstract machines} have been proposed\cite{kieburtz_g-machine_1985,fairbairn_tim_1987,burn_spineless_1988,koopman_fresh_1989,jones_implementing_nodate}.
+It is a difficult balance, and over the years many different evaluator designs, or \textit{abstract machines}, have been proposed\cite{kieburtz_g-machine_1985,fairbairn_tim_1987,burn_spineless_1988,koopman_fresh_1989,jones_implementing_nodate}.
 One of the early approaches to implementating lazy functional languages is combinator graph reduction\cite{turner_new_1979}.
@@
 Combinator reduction was the runtime for David Turner's Miranda language\cite{turner_miranda_1985}, the popular choice for pure functional langauges at the time, and one of the few to be commercially supported.
-Eventually the interest for Miranda and combinator graph reduction with it waned, and today Haskell is the dominant language.
-Part of the reason may be that Miranda was closed source for 35 years, its source code was released only in 2020\cite{noauthor_open_2021}, but combinator reduction also deserves part of the blame.
-Expressions are reduced in many small steps, so the per-step overhead greatly impacts runtime performance.
+Eventually the interest for Miranda and combinator graph reduction with it waned.
+Part of the reason may be that Miranda was until very recently a closed source, licensed software (its source code was released only in 2020\cite{noauthor_open_2021}), leading to the creation of the open source Haskell language.
+But combinator reduction also deserves part of the blame.
+Firstly, expressions are reduced in many small steps, so the per-step overhead greatly impacts runtime performance.
 Every reduction step modifies the stack and usually also the graph, resulting in high memory traffic.
-Another important issue is that the translation to combinators may result in much larger program.
-For a lambda calculus expression of size $N$, the traditional bracket abstraction algorithm produces an output of $O(n^2)$ in the worst case, were other approaches such as supercombinators are linear.
+Another important issue is that the translation to combinators may result in much larger programs.
+For a lambda calculus expression of size $N$, the traditional bracket abstraction algorithm produces an output of $O(N^2)$ in the worst case, where other approaches such as supercombinators are linear\cite{spj_impl}.
 
@@
 Recently however, a new algorithm for translation of lambda calculus expressions into combinatory logic has been proposed that is linear in the size of the input\cite{kiselyov_lambda_2018}.
-While a clear improvement over previous translation strategies, it remains to be seen whether it can make the combinator reduction approach competitive with contemporary reduction engines.
+While a clear improvement over previous translation strategies, it remains to be seen how big the impact is on real-world performance.
+Is it enough to bridge the performance gap with newer reduction engines?
 This leads us to our first research question:
 
-\textbf{RQ1:} How does Kiselyov's semantic translation compare to the classical bracket abstraction algorithm in terms of runtime performance of compiled programs?
+\textbf{RQ1:} Does Kiselyov's semantic translation make combinator graph reduction competitive with contemporary lazy functional evaluators?
 
-We can take a set of benchmark programs and compile them with both algorithms, but to compare real-world execution time we also need a fast runtime to execute the compiled programs.
-
-Alas, there is no performant combinator graph reduction engine on which we can execute programs compiled with Kiselyov's.
+We use GHC as the reference for contemporary evaluator, and then measure the execution time of a set of benchmark programs.
+We also need a fast reduction engine that can run the program compiled with Kiselyov's algorithm.
+Alas, there is no performant combinator graph reduction engine tailored to executing such programs.
 The now open sourced Miranda is a good starting point, but it is showing its age.
@@
 Using this engine, we can then answer RQ1.
+
+Concretely, our contributions are the following:
+\begin{itemize}
+    \item Kiselyov describes not one but two algorithms. The first does not have linear complexity, but tends to produce very compact code.
+          The second is guaranteed to be of linear complexity, at the cost of being less compact in certain cases.
+          We describe both algorithms in a way that is straightforward to implement.
+    \item We implement a fast combinator graph reduction engine based on the ideas in Miranda~\cite{turner_miranda_1985} and later improvements~\cite{koopman_architecture_1992}.
+    \item We evaluate the performance of Miranda, Haskell and our engine on a set of benchmark programs.
+\end{itemize}
 
@@
 \subsection*{Miranda}
+Describe implementation of SASL/Miranda.
+
+Compare performance with Open Source Miranda.
+
 \subsection*{TIGRE}
+Describe changes made to align with TIGRE.
+
+Compare performance with SASL-style engine.
+
+\section*{Performance evaluation}
+Programs:
+\begin{itemize}
+    \item ackermann
+    \item digits of e
+    \item linfib
+    \item lsort
+    \item nfib
+    \item primes
+    \item queens
+    \item tak
+    \item towers of hanoi
+    \item treesort
+\end{itemize}
+
+Evaluation performance of:
+
+\begin{itemize}
+    \item Bracket abstraction
+    \item Kiselyov
+    \item GHC
+\end{itemize}
 
@@
 \bibliographystyle{IEEEtran}
-\bibliography{zotero}
+\bibliography{zotero, references}
 

### 2022-12-07 experiments/superg/doc/paper/paper.tex
Grammar.

@@
 \begin{IEEEkeywords}
-    combinatory logic, graph reduction, lazy functional language, lazy evalution
+    combinatory logic, graph reduction, lazy functional language, lazy evaluation
 \end{IEEEkeywords}
@@
 \section{Introduction}
-% Why did I research this. Give the motive, usually a knowledge gap or problem
+% Why did I research this? Give the motive, usually a knowledge gap or problem
 Lazy evaluation of programs is a powerful idea.
-It allows a program to operate on infinitely large data structures, and if part of the computation turns out to be unnecessary, it can be much more efficient than a strict evaluation.
-In practice however, delaying computation often introduces overhead, making languages based on lazy evaluation slower than their strict counterparts.
-To compete with a strict implementation, a lazy runtime must maximize lazyness while minizing overhead.
+It allows a program to operate on infinitely large data structures, and if part of the computation turns out to be unnecessary, it is more efficient than a strict evaluation.
+In practice, delaying computation often introduces overhead, making languages based on lazy evaluation slower than their strict counterparts.
+To compete with a strict implementation, a lazy runtime must maximize laziness while minimizing overhead.
 It is a difficult balance, and over the years many different evaluator designs, or \textit{abstract machines}, have been proposed\cite{kieburtz_g-machine_1985,fairbairn_tim_1987,burn_spineless_1988,koopman_fresh_1989,jones_implementing_nodate}.
-One of the early approaches to implementating lazy functional languages is combinator graph reduction\cite{turner_new_1979}.
+One of the early approaches to implementing lazy functional languages is combinator graph reduction\cite{turner_new_1979}.
 
@@
 
-Combinator reduction was the runtime for David Turner's Miranda language\cite{turner_miranda_1985}, the popular choice for pure functional langauges at the time, and one of the few to be commercially supported.
-Eventually the interest for Miranda and combinator graph reduction with it waned.
-Part of the reason may be that Miranda was until very recently a closed source, licensed software (its source code was released only in 2020\cite{noauthor_open_2021}), leading to the creation of the open source Haskell language.
+Combinator reduction was the runtime for David Turner's Miranda language\cite{turner_miranda_1985}, the popular choice for pure functional languages at the time, and one of the few to be commercially supported.
+Eventually, the interest in Miranda and combinator graph reduction with it waned.
+Part of the reason may be that Miranda was until very recently a closed-source, licensed software (its source code was released only in 2020\cite{noauthor_open_2021}), leading to the creation of the open-source Haskell language.
 But combinator reduction also deserves part of the blame.
@@
 For a lambda calculus expression of size $N$, the traditional bracket abstraction algorithm produces an output of $O(N^2)$ in the worst case, where other approaches such as supercombinators are linear\cite{spj_impl}.
-
-These disadvantages have led to combinator graph reduction to be all but abandoned.
-Recently however, a new algorithm for translation of lambda calculus expressions into combinatory logic has been proposed that is linear in the size of the input\cite{kiselyov_lambda_2018}.
-While a clear improvement over previous translation strategies, it remains to be seen how big the impact is on real-world performance.
-Is it enough to bridge the performance gap with newer reduction engines?
-This leads us to our first research question:
-
-\textbf{RQ1:} Does Kiselyov's semantic translation make combinator graph reduction competitive with contemporary lazy functional evaluators?
-
-We use GHC as the reference for contemporary evaluator, and then measure the execution time of a set of benchmark programs.
-We also need a fast reduction engine that can run the program compiled with Kiselyov's algorithm.
-Alas, there is no performant combinator graph reduction engine tailored to executing such programs.
-The now open sourced Miranda is a good starting point, but it is showing its age.
-The 2020 port compiles cleanly on a recent GCC, but enabling optimizations is known to break the garbage collector.
-Miranda's architecture is based on cells with a tag, head and tail as described in\cite{turner_new_1979}, but later work has shown that a tagless representation with direct pointers to code leads to better performance\cite{koopman_fresh_1989}.
-In absence of an up to date reference on building a practical combinator reduction engine, we formlate the second research question:
-
-\textbf{RQ2:} How is combinator graph reduction implemented efficiently on conventional hardware?
-
-We evaluate the state of the art in combinator graph reduction, and build a fast implementation for a language that is simple but powerful enough to express our benchmark programs.
-Using this engine, we can then answer RQ1.
-
-Concretely, our contributions are the following:
-\begin{itemize}
-    \item Kiselyov describes not one but two algorithms. The first does not have linear complexity, but tends to produce very compact code.
-          The second is guaranteed to be of linear complexity, at the cost of being less compact in certain cases.
-          We describe both algorithms in a way that is straightforward to implement.
-    \item We implement a fast combinator graph reduction engine based on the ideas in Miranda~\cite{turner_miranda_1985} and later improvements~\cite{koopman_architecture_1992}.
-    \item We evaluate the performance of Miranda, Haskell and our engine on a set of benchmark programs.
-\end{itemize}
+These disadvantages have led to combinator graph reduction being all but abandoned.
 
@@
 %   * Theoretical or societal relevance
+Recently, however, a new algorithm for the translation of lambda calculus expressions into combinatory logic has been proposed that is linear in the size of the input\cite{kiselyov_lambda_2018}.
+While a clear improvement over previous translation strategies, it remains to be seen how big the impact is on real-world performance.
+Is it enough to bridge the performance gap with newer reduction engines based on different paradigms?
+This leads us to our first research question:
+
+\textbf{RQ1:} Does Kiselyov's semantic translation make combinator graph reduction competitive with contemporary lazy functional evaluators?
 
@@
 % - Sub-questions
+We use GHC as the reference for a contemporary evaluator, and then measure the execution time of a set of benchmark programs.
+We also need a fast reduction engine that can run the program compiled with Kiselyov's algorithm.
+Alas, there is no performant combinator graph reduction engine tailored to executing such programs.
+The now open-sourced Miranda is a good starting point, but it is showing its age.
+The 2020 port compiles cleanly on a recent GCC, but enabling optimizations is known to break the garbage collector.
+Miranda's architecture is based on cells with a tag, head and tail as described in\cite{turner_new_1979}, but later work has shown that a tagless representation with direct pointers to code leads to better performance\cite{koopman_fresh_1989}.
+In absence of an up-to-date reference on building a practical combinator reduction engine, we formulate the second research question:
+
+\textbf{RQ2:} How is combinator graph reduction implemented efficiently on conventional hardware?
+
+We evaluate the state of the art in combinator graph reduction and build a fast implementation for a language that is simple but powerful enough to express our benchmark programs.
+Using this engine, we can then answer RQ1.
 
 % How did I organize the report?
-% - Structure overview
+% - Structure Overview
+Concretely, our contributions are the following:
+\begin{itemize}
+    \item Kiselyov describes not one but two algorithms. The first does not have linear complexity, but tends to produce very compact code.
+          The second is guaranteed to be of linear complexity, at the cost of being less compact in certain cases.
+          In section~\ref{sec:kiselyov} we describe both algorithms in a way that is straightforward to implement.
+    \item We implement a fast combinator graph reduction engine based on the ideas in Miranda~\cite{turner_miranda_1985} and later improvements~\cite{koopman_architecture_1992}.
+          We elaborate on its design in section~\ref{sec:engine}.
+    \item We evaluate the performance of Miranda, Haskell and our engine on a set of benchmark programs in section~\ref{sec:eval}.
+\end{itemize}
 
-%NOTES:
-%While initially a popular approach to lazy evaluation, combinator graph reduction fell out of fashion with the advent of new abstract machines based on supercombinators (G-machine, TIM).
-%
-%In the 1980s David Turner developed the language Miranda\cite{turner_miranda_1985}.
-%At the time it was the popular choice for lazy functional programming, and one of few to be commercially supported.
-%
-%Lazy evaluation of programs is a powerful idea.
-%It allows a program to operate on infinitely large data structures, and if part of the computation turns out to be unnecessary, it can be much more efficient than a strict evaluation.
-%In practice however, delaying computation often introduces overhead, making languages based on lazy evaluation slower than their strict counterparts.
-%To compete with a strict implementation, a lazy runtime must maximize lazyness while minizing overhead.
-%
-%Some of the early lazy functional languages were built on combinatory logic\cite{turner_new_1979,turner_miranda_1985}.
-%Where the well-known lambda calculus has a concept of named variables that an evaluator must track and substitute values for in function application,
-%in combinatory logic there are no bound variables, only a small .
-%
-%- Most important lazy functional language today is Haskell
-%- It is built on top of the spineless tagless G-machine model
-%- STG combines ideas from G-machine and TIM, both in the class of supercombinators
-%- *What are supercombinators?*
-%- Haskell inherits from Miranda, which was built on a very different model, combinator graph reduction
-%- *What is Combinator Graph Reduction?*
-%- Advantages of Combinators:
-%* Small and fixed set of combinators. Reduction machine is simple, can be done in hardware.
-%* Lambda bodies instantiated lazily, lazier than supercombinators
-%- Disadvantages:
-%* Reduction steps are small. Lot of transient memory traffic, high strain on garbage collector. Caching is also very coarse-grained.
-%* Translation to combinators results in a larger program. (attack this with Kiselyov)
-%- At the turn of the century, combinator graph reduction was all but abandoned.
-%- Recently, a new translation strategy has been proposed for translating to combinators (cue Kiselyov).
-%- After 35 years under a commercial license, Miranda was released under the open BSD license.
-%- In this work we build on Miranda's design, incorporating the proposed semantic combinator translation algorithm and changes to the execution engine proposed in later papers.
-%
-%Research Questions:
-%1. How is combinator graph reduction implemented efficiently on conventional hardware?
-%2. How does Kiselyov's semantic translation compare to the classical bracket abstraction algorithm in terms of runtime performance of compiled programs?
+\section{Kiselyov's algorithms}
+\label{sec:kiselyov}
+Describes the strict and lazy translation algorithms in practical terms.
 
-\section*{Kiselyov's algorithms}
-Describes the strict and lazy translation algorithms in a practical manner.
-
-\section*{Engine implementation}
+\section{Engine implementation}
+\label{sec:engine}
 \subsection*{Miranda}
-Describe implementation of SASL/Miranda.
+Describe the implementation of SASL/Miranda.
 
@@
 
-\section*{Performance evaluation}
+\section{Performance evaluation}
+\label{sec:eval}
 Programs:
@@
 
-\section*{Conclusion}
+\section{Conclusion}
 % Repeat the main question / objective of your research

### 2022-12-07 experiments/superg/doc/paper/paper.tex
Explain combinatory logic.

@@
 To compete with a strict implementation, a lazy runtime must maximize laziness while minimizing overhead.
-It is a difficult balance, and over the years many different evaluator designs, or \textit{abstract machines}, have been proposed\cite{kieburtz_g-machine_1985,fairbairn_tim_1987,burn_spineless_1988,koopman_fresh_1989,jones_implementing_nodate}.
-One of the early approaches to implementing lazy functional languages is combinator graph reduction\cite{turner_new_1979}.
+It is a difficult balance, and over the years many different evaluator designs, or \textit{abstract machines}, have been proposed~\cite{turner_new_1979,kieburtz_g-machine_1985,fairbairn_tim_1987,burn_spineless_1988,koopman_fresh_1989,jones_implementing_nodate}.
+One of the early approaches to implementing lazy functional languages is combinator graph reduction~\cite{turner_new_1979}.
 
-\textbf{TODO:} what is combinator reduction?
+Starting from a lambda calculus expression, we first translate it into an expression consisting exclusively of the application of a small set of predefined functions, or \textit{combinators}.
+It turns out that any lambda calculus expression can be expressed using just three combinators:
+\begin{itemize}
+    \item $S \ f \ g \ x = f \ x \ (g \ x)$
+    \item $K \ x \ y = x$
+    \item $I \ x = x$\footnote{It turns out that $SKK = I$, so even this combinator is not strictly necessary, but all reasonable implementations do include $I$, and it simplifies the example.}
+\end{itemize}
 
-Combinator reduction was the runtime for David Turner's Miranda language\cite{turner_miranda_1985}, the popular choice for pure functional languages at the time, and one of the few to be commercially supported.
+For example, $\lambda x.x+1$ can be converted into the combinator expression:
+
+\[
+    S \ (S \  (K \  +) \  I) \ (K \ 1)
+\]
+
+We invite the reader to verify this by applying $x$ to this expression and reducing it to $x+1$ with the usual $\beta$-reduction rule.
+While the generated expression is not as easy to read, the simplicity of its components is appealing: to execute a program compiled to combinators, an evaluator only needs to support four operations: the built-in $S$, $K$ and $I$ combinators, and function application\footnote{And any builtins of the language, like the plus operator in the example.}.
+Better still, conversion to combinators removes all bound variables.
+The runtime need not concern itself with passing arguments, this is all handled by the combinators.
+The compiled program is executed using graph reduction, a technique that is also common in other reduction engines\cite{kieburtz_g-machine_1985,fairbairn_tim_1987}.
+\textbf{TODO: } Should we also cover graph reduction?
+% In the reduction graph, vertices are combinators, constants or function applications, and edges connect the function and argument for application (combinators and constants have no outgoing edges, they are leaf nodes).
+
+Combinator reduction was the runtime for David Turner's Miranda language~\cite{turner_miranda_1985}, the popular choice for pure functional languages at the time, and one of the few to be commercially supported.
 Eventually, the interest in Miranda and combinator graph reduction with it waned.
-Part of the reason may be that Miranda was until very recently a closed-source, licensed software (its source code was released only in 2020\cite{noauthor_open_2021}), leading to the creation of the open-source Haskell language.
+Part of the reason may be that Miranda was until very recently a closed-source, licensed software (its source code was released only in 2020~\cite{noauthor_open_2021}), leading to the creation of the open-source Haskell language.
 But combinator reduction also deserves part of the blame.
-Firstly, expressions are reduced in many small steps, so the per-step overhead greatly impacts runtime performance.
+Firstly, owing to the simplicity of individual combinators, expressions are reduced in many small steps, so the per-step overhead greatly impacts runtime performance.
 Every reduction step modifies the stack and usually also the graph, resulting in high memory traffic.
@@
 % What did I research?
-% - Research problem / question
+% - Research problem/question
 %   * Definitions

### 2022-12-07 experiments/superg/src/compiled_expr.rs

@@ -112,7 +112,7 @@ impl CompiledExpr {
                 let r = CompiledExpr::compile(r);
                 cap(l, r)
             }
-            ast::Expr::Lam(_, _) => todo!(),
+            ast::Expr::Lam(x, e) => CompiledExpr::compile(e).abstract_var(x),
         }
     }
 
@@ -129,5 +129,28 @@ impl CompiledExpr {
             }
             i @ CompiledExpr::Int(_) => cap(CompiledExpr::Comb(Comb::K), i),
         }
+    }
+}
+
+#[cfg(test)]
+mod tests {
+    use super::*;
+    use crate::lexer::lex;
+    use crate::parser::parse_expr;
+
+    #[test]
+    fn compile_increment() {
+        let expr = "lam (x) (+ x 1)";
+        let mut tokens = lex(expr);
+        let parsed_expr = parse_expr(&mut tokens);
+        let compiled = CompiledExpr::compile(&parsed_expr);
+        use Comb::{Plus, I, K, S};
+        assert_eq!(
+            compiled,
+            cap(
+                cap(S, cap(cap(S, cap(K, Plus)), I)),
+                cap(K, CompiledExpr::Int(1))
+            )
+        );
     }
 }

### 2022-12-08 experiments/superg/doc/paper/paper.tex
Make intro more concise.

@@
 It is a difficult balance, and over the years many different evaluator designs, or \textit{abstract machines}, have been proposed~\cite{turner_new_1979,kieburtz_g-machine_1985,fairbairn_tim_1987,burn_spineless_1988,koopman_fresh_1989,jones_implementing_nodate}.
+
 One of the early approaches to implementing lazy functional languages is combinator graph reduction~\cite{turner_new_1979}.
+It was the basis for David Turner's Miranda language~\cite{turner_miranda_1985}, the popular choice for pure functional languages and one of the few to be commercially supported at the time.
+Eventually, the interest in Miranda and combinator graph reduction with it waned.
+Part of the reason may be that Miranda was until very recently a closed-source, licensed software (its source code was released only in 2020~\cite{noauthor_open_2021}), leading to the creation of the open-source Haskell language~\cite{hudak_history_2007}.
+But combinator reduction also deserves part of the blame.
+% Firstly, owing to the simplicity of individual combinators, expressions are reduced in many small steps, so the per-step overhead greatly impacts runtime performance.
+% Every reduction step modifies the stack and usually also the graph, resulting in high memory traffic.
+One of the major issues is that the translation to combinators may result in much larger programs.
+For a lambda calculus expression of size $N$, the traditional bracket abstraction algorithm produces an output of $O(N^2)$ in the worst case, where other approaches such as supercombinators are linear\cite{spj_impl}.
+Larger programs run slower, and along with other disadvantages, this led to combinator graph reduction being all but abandoned.
 
+% What did I research?
+% - Research problem/question
+%   * Definitions
+%   * Choices
+%   * Presuppositions
+% - Why specifically this research question?
+%   * Theoretical or societal relevance
+Recently, however, a new algorithm for the translation of lambda calculus expressions into combinatory logic has been proposed that is linear in the size of the input\cite{kiselyov_lambda_2018}.
+Kiselyov's translation algorithm is based on semantics rather than syntax and this deeper understanding of the program has been shown to produce more compact translations.
+While a clear improvement over previous translation strategies, it remains to be seen how big the impact is on real-world performance.
+Is it enough to bridge the performance gap with newer reduction engines based on different paradigms?
+This leads us to our first research question:
+
+\textbf{RQ1:} Does Kiselyov's semantic translation make combinator graph reduction competitive with contemporary lazy functional evaluators?
+
+% How did I research this?
+% - Methods
+% - Sub-questions
+We use GHC as the reference for a contemporary evaluator, and then measure the execution time of a set of benchmark programs.
+We also need a fast reduction engine that can run programs compiled to combinator expressions.
+Alas, there is no performant reduction engine tailored to executing such programs.
+The now open-sourced Miranda is a good starting point, but it is showing its age.
+The 2020 port compiles cleanly on a recent GCC, but enabling optimizations is known to break the garbage collector.
+Miranda's architecture is based on cells with a tag, head and tail as described in\cite{turner_new_1979}, but later work has shown that a tagless representation with direct pointers to code leads to better performance\cite{koopman_fresh_1989}.
+In absence of an up-to-date reference on building a practical combinator reduction engine, we formulate the second research question:
+
+\textbf{RQ2:} How is combinator graph reduction implemented efficiently on conventional hardware?
+
+We evaluate the state of the art in combinator graph reduction and build a fast implementation for a language that is simple but powerful enough to express our benchmark programs.
+Using this engine, we can then answer RQ1.
+
+% How did I organize the report?
+% - Structure Overview
+Concretely, our contributions are the following:
+\begin{itemize}
+    \item Kiselyov develops not one but two algorithms. The first does not have linear complexity, but tends to produce very compact code.
+          The second is guaranteed to be of linear complexity, at the cost of being less compact in certain cases.
+          Kiselyov's description is somewhat theoretical, so in section~\ref{sec:kiselyov} we describe both algorithms in a way that is straightforward to implement.
+    \item We implement a fast combinator graph reduction engine based on the ideas in Miranda~\cite{turner_miranda_1985} and later improvements~\cite{koopman_architecture_1992}.
+          We elaborate on its design in section~\ref{sec:engine}.
+    \item We evaluate the performance of Miranda, Haskell and our engine on a set of benchmark programs in section~\ref{sec:eval}.
+\end{itemize}
+
+For readers unfamiliar with combinatory logic and/or graph reduction, we provide a short introduction to these topics in section~\ref{sec:prelim}.
+
+\section{Preliminaries}
+\label{sec:prelim}
+\textbf{TODO: write section. Currently contains only a few small fragments}
+\subsection{Combinatory Logic}
 Starting from a lambda calculus expression, we first translate it into an expression consisting exclusively of the application of a small set of predefined functions, or \textit{combinators}.
@@
 The compiled program is executed using graph reduction, a technique that is also common in other reduction engines\cite{kieburtz_g-machine_1985,fairbairn_tim_1987}.
-\textbf{TODO: } Should we also cover graph reduction?
+
+\subsection{Graph Reduction}
 % In the reduction graph, vertices are combinators, constants or function applications, and edges connect the function and argument for application (combinators and constants have no outgoing edges, they are leaf nodes).
-
-Combinator reduction was the runtime for David Turner's Miranda language~\cite{turner_miranda_1985}, the popular choice for pure functional languages at the time, and one of the few to be commercially supported.
-Eventually, the interest in Miranda and combinator graph reduction with it waned.
-Part of the reason may be that Miranda was until very recently a closed-source, licensed software (its source code was released only in 2020~\cite{noauthor_open_2021}), leading to the creation of the open-source Haskell language.
-But combinator reduction also deserves part of the blame.
-Firstly, owing to the simplicity of individual combinators, expressions are reduced in many small steps, so the per-step overhead greatly impacts runtime performance.
-Every reduction step modifies the stack and usually also the graph, resulting in high memory traffic.
-Another important issue is that the translation to combinators may result in much larger programs.
-For a lambda calculus expression of size $N$, the traditional bracket abstraction algorithm produces an output of $O(N^2)$ in the worst case, where other approaches such as supercombinators are linear\cite{spj_impl}.
-These disadvantages have led to combinator graph reduction being all but abandoned.
-
-% What did I research?
-% - Research problem/question
-%   * Definitions
-%   * Choices
-%   * Presuppositions
-% - Why specifically this research question?
-%   * Theoretical or societal relevance
-Recently, however, a new algorithm for the translation of lambda calculus expressions into combinatory logic has been proposed that is linear in the size of the input\cite{kiselyov_lambda_2018}.
-While a clear improvement over previous translation strategies, it remains to be seen how big the impact is on real-world performance.
-Is it enough to bridge the performance gap with newer reduction engines based on different paradigms?
-This leads us to our first research question:
-
-\textbf{RQ1:} Does Kiselyov's semantic translation make combinator graph reduction competitive with contemporary lazy functional evaluators?
-
-% How did I research this?
-% - Methods
-% - Sub-questions
-We use GHC as the reference for a contemporary evaluator, and then measure the execution time of a set of benchmark programs.
-We also need a fast reduction engine that can run the program compiled with Kiselyov's algorithm.
-Alas, there is no performant combinator graph reduction engine tailored to executing such programs.
-The now open-sourced Miranda is a good starting point, but it is showing its age.
-The 2020 port compiles cleanly on a recent GCC, but enabling optimizations is known to break the garbage collector.
-Miranda's architecture is based on cells with a tag, head and tail as described in\cite{turner_new_1979}, but later work has shown that a tagless representation with direct pointers to code leads to better performance\cite{koopman_fresh_1989}.
-In absence of an up-to-date reference on building a practical combinator reduction engine, we formulate the second research question:
-
-\textbf{RQ2:} How is combinator graph reduction implemented efficiently on conventional hardware?
-
-We evaluate the state of the art in combinator graph reduction and build a fast implementation for a language that is simple but powerful enough to express our benchmark programs.
-Using this engine, we can then answer RQ1.
-
-% How did I organize the report?
-% - Structure Overview
-Concretely, our contributions are the following:
-\begin{itemize}
-    \item Kiselyov describes not one but two algorithms. The first does not have linear complexity, but tends to produce very compact code.
-          The second is guaranteed to be of linear complexity, at the cost of being less compact in certain cases.
-          In section~\ref{sec:kiselyov} we describe both algorithms in a way that is straightforward to implement.
-    \item We implement a fast combinator graph reduction engine based on the ideas in Miranda~\cite{turner_miranda_1985} and later improvements~\cite{koopman_architecture_1992}.
-          We elaborate on its design in section~\ref{sec:engine}.
-    \item We evaluate the performance of Miranda, Haskell and our engine on a set of benchmark programs in section~\ref{sec:eval}.
-\end{itemize}
 
@@
 \label{sec:kiselyov}
-Describes the strict and lazy translation algorithms in practical terms.
+Describes the algorithms as implementable functions. Both recursive and linear semantic functions are covered (the two algorithms).
 
@@
 \subsection*{Miranda}
-Describe the implementation of SASL/Miranda.
+Describe the implementation of a SASL/Miranda-style engine.
 
@@
 
-Compare performance with SASL-style engine.
+Compare performance with Miranda-style engine.
 
@@
 
-Evaluation performance of:
+Evaluate performance of:
 

### 2022-12-09 experiments/superg/doc/paper/paper.tex
Update link to source code.
Ref to RQ1 in bold.

@@
 %\usepackage{float}
-%\usepackage{url}
+\usepackage{hyperref}
+\hypersetup{
+    colorlinks=true,
+    linkcolor=blue,
+    filecolor=magenta,      
+    urlcolor=cyan,
+    pdftitle={Reviving the Lost Art of Combinator Graph Reduction},
+    pdfpagemode=FullScreen,
+    }
+
+\urlstyle{same}
 
@@
 We evaluate the state of the art in combinator graph reduction and build a fast implementation for a language that is simple but powerful enough to express our benchmark programs.
-Using this engine, we can then answer RQ1.
+Using this engine, we can then answer \textbf{RQ1}.
 
@@
 % Give the answer to the question
+Source code for our Superg evaluator is available at \url{https://github.com/wildarch/mono/tree/main/experiments/superg}.
 

### 2022-12-10 experiments/superg/src/turner/mod.rs
helpers to access tag,hd,tl.

@@ -97,19 +97,19 @@ impl TurnerEngine {
                         let l2 = self.stack[self.stack.len() - 4];
 
                         // If x is an int, we should transfer that to the new location
-                        let x_tag = self.tag[l2.0 as usize] & TAG_RHS_INT;
-                        let x = self.tl[l2.0 as usize];
-                        let g = self.tl[l1.0 as usize];
-                        let f = self.tl[l0.0 as usize];
+                        let x_tag = self.tag(l2) & TAG_RHS_INT;
+                        let x = self.tl(l2);
+                        let g = self.tl(l1);
+                        let f = self.tl(l0);
 
                         // Make lower cells
                         // TODO: check if we guarantee nobody else holds a reference to cells l0 and l1, so we can overwrite them
                         let left_cell = self.make_cell(x_tag | TAG_WANTED, f, x);
                         let right_cell = self.make_cell(x_tag | TAG_WANTED, g, x);
                         // Overwrite upper cell
-                        self.tag[l2.0 as usize] = TAG_WANTED;
-                        self.hd[l2.0 as usize] = left_cell;
-                        self.tl[l2.0 as usize] = right_cell;
+                        self.set_tag(l2, TAG_WANTED);
+                        self.set_hd(l2, left_cell);
+                        self.set_tl(l2, right_cell);
                         // Truncate to upper cell
                         self.stack.truncate(self.stack.len() - 3);
                     }
@@ -123,14 +123,14 @@ impl TurnerEngine {
                         // K x y = x
                         let l0 = self.stack[self.stack.len() - 2];
                         let l1 = self.stack[self.stack.len() - 3];
-                        let x = self.tl[l0.0 as usize];
+                        let x = self.tl(l0);
                         // If x is an int, we should transfer that to the new location
-                        let x_tag = self.tag[l0.0 as usize] & TAG_RHS_INT;
+                        let x_tag = self.tag(l0) & TAG_RHS_INT;
 
                         // Make the indirection node
-                        self.tag[l1.0 as usize] = x_tag | TAG_WANTED;
-                        self.hd[l1.0 as usize] = CellPtr(Comb::I as i32);
-                        self.tl[l1.0 as usize] = x;
[…]

### 2022-12-14 experiments/superg/src/kiselyov/lazy_opt.rs
Kiselyov-style compilation for Eta.

@@ -3,200 +3,79 @@ use crate::compiled_expr::{cap, Comb, CompiledExpr};
 
 use super::BExpr;
 
-#[derive(Debug, Clone, Copy, PartialEq)]
-enum ContextElem {
-    Ignored,
-    Used,
+pub fn compile_lazy_opt(e: &BExpr) -> CompiledExpr {
+    match compile(e) {
+        CExpr::Closed(e) => e,
+        _ => unreachable!("expect to reach a closed expression"),
+    }
 }
-type Context = Vec<ContextElem>;
 
-fn infer_context(e: &BExpr) -> Context {
-    use ContextElem::*;
+#[derive(Debug, Clone)]
+enum CExpr {
+    Closed(CompiledExpr),
+    Value,
+    Next(Box<CExpr>),
+    Weak(Box<CExpr>),
+}
+
+fn compile(e: &BExpr) -> CExpr {
     match e {
         BExpr::Var(i) => {
-            let mut v = vec![Used];
+            let mut e = CExpr::Value;
             for _ in 0..*i {
-                v.push(Ignored);
+                e = CExpr::Weak(Box::new(e));
             }
-            v
+            e
         }
-        BExpr::Lam(e) => {
-            let mut ctx = infer_context(e);
-            ctx.pop();
-            ctx
-        }
-        BExpr::Ap(f, a) => unify_contexts(infer_context(f), infer_context(a)),
+        BExpr::Lam(e) => match compile(e) {
+            CExpr::Closed(d) => CExpr::Closed(cap(Comb::K, d)),
+            CExpr::Value => CExpr::Closed(CompiledExpr::Comb(Comb::I)),
+            CExpr::Next(e) => *e,
+            CExpr::Weak(e) => semantic(CExpr::Closed(CompiledExpr::Comb(Comb::K)), *e),
+        },
+        BExpr::Ap(a, b) => semantic(compile(a), compile(b)),
         _ => todo!(),
     }
 }
 
-fn unify_contexts(mut a: Context, mut b: Context) -> Context {
-    use ContextElem::*;
-    let mut ctx = Context::new();
-    loop {
-        let e = match (a.pop(), b.pop()) {
-            (Some(Used), _) | (_, Some(Used)) => Used,
-            (None, Some(Ignored)) | (Some(Ignored), None) | (Some(Ignored), Some(Ignored)) => {
-                Ignored
-            }
-            (None, None) => {
[…]

### 2022-12-14 experiments/superg/src/kiselyov/lazy_opt.rs
Cleaning.

@@ -10,6 +10,7 @@ pub fn compile_lazy_opt(e: &BExpr) -> CompiledExpr {
     }
 }
 
+// Internal type to store context together with a compiled expression.
 #[derive(Debug, Clone)]
 enum CExpr {
     Closed(CompiledExpr),
@@ -40,42 +41,36 @@ fn compile(e: &BExpr) -> CExpr {
 
 fn semantic(e1: CExpr, e2: CExpr) -> CExpr {
     use CExpr::*;
-    fn weak(e: CExpr) -> CExpr {
-        Weak(Box::new(e))
+    use Comb::*;
+    fn weak<C: Into<Box<CExpr>>>(e: C) -> CExpr {
+        Weak(e.into())
     }
-    fn next(e: CExpr) -> CExpr {
-        Next(Box::new(e))
+    fn next<C: Into<Box<CExpr>>>(e: C) -> CExpr {
+        Next(e.into())
+    }
+    fn closed<C: Into<CompiledExpr>>(e: C) -> CExpr {
+        Closed(e.into())
+    }
+    fn semantic3(e1: CExpr, e2: CExpr, e3: CExpr) -> CExpr {
+        semantic(semantic(e1, e2), e3)
     }
     match (e1, e2) {
         (Weak(e1), Weak(e2)) => weak(semantic(*e1, *e2)),
-        (Weak(e), Closed(d)) => weak(semantic(*e, Closed(d))),
-        (Closed(d), Weak(e)) => weak(semantic(Closed(d), *e)),
-        (Weak(e), Value) => Next(e),
-        (Value, Weak(e)) => next(semantic(Closed(cap(Comb::C, Comb::I)), *e)),
-        (Weak(e1), Next(e2)) => next(semantic(
-            semantic(Closed(CompiledExpr::Comb(Comb::B)), *e1),
-            *e2,
-        )),
-        (Next(e1), Weak(e2)) => next(semantic(
-            semantic(Closed(CompiledExpr::Comb(Comb::C)), *e1),
-            *e2,
-        )),
-        (Next(e1), Next(e2)) => next(semantic(
-            semantic(Closed(CompiledExpr::Comb(Comb::S)), *e1),
-            *e2,
-        )),
-        (Next(e), Value) => next(semantic(
-            semantic(Closed(CompiledExpr::Comb(Comb::S)), *e),
-            Closed(CompiledExpr::Comb(Comb::I)),
-        )),
[…]

### 2022-12-14 experiments/superg/src/kiselyov/linear.rs
linear: pass tests.

@@ -85,15 +85,48 @@ mod tests {
     use super::*;
     use crate::{lex, parser::parse_expr};
 
+    // Testcases taken from Kiselyov's paper, Table 1.
     #[test]
-    fn example() {
-        let expr = "lam (x y) (y x)";
+    fn kiselyov_examples() {
+        assert_compiles_to("lam (x) (lam (y) y)", "KI");
+        assert_compiles_to("lam (x) (lam (y) x)", "BKI");
+        // Kiselyov' paper contains an error, it states the output is: C(BS(BKI))I
+        assert_compiles_to("lam (x) (lam (y) (x y))", "C(BS(BKI))I");
+        assert_compiles_to("lam (x) (lam (y) (y x))", "B(SI)(BKI)");
+        assert_compiles_to("lam (x) (lam (y) (lam (z) (z x)))", "B2(SI) (B2 K(BKI))");
+        assert_compiles_to(
+            "lam (x) (lam (y) (lam (z) ((lam (w) w) x)))",
+            "B3 I(B2 K(BKI))",
+        );
+        assert_compiles_to(
+            "lam (x) (lam (y) (lam (z) ((x z) (y z))))",
+            "C(B S2(C2 (B2 S(B2 K (BKI)))I)) (C(BS(BKI))I)",
+        );
+    }
+
+    #[test]
+    fn kiselyov_worsecase() {
+        assert_compiles_to("lam (x) (lam (y) (y x))", "B(SI)(BKI)");
+        assert_compiles_to(
+            "lam (x) (lam (y) (lam (z) (z y x)))",
+            "B(S2 (B(SI)(BKI))) (B2 K(BKI))",
+        );
+        assert_compiles_to(
+            "lam (x) (lam (y) (lam (z) (lam (a) (a z y x))))",
+            "B(S3 (B(S2 (B(SI)(BKI))) (B2 K(BKI)))) (B3 K(B2 K (BKI)))",
+        );
+    }
+
+    fn assert_compiles_to(expr: &str, compiled_expr: &str) {
+        // Compile expr
         let mut tokens = lex(expr);
         let parsed_expr = parse_expr(&mut tokens);
         let bexpr = to_debruijn(&parsed_expr, &mut vec![]);
-        let compiled = compile_linear(&bexpr);
-        use Comb::{B, I, K, S};
-        assert_eq!(compiled, cap(cap(B(1), cap(S(1), I)), cap(cap(B(1), K), I)))
[…]

### 2022-12-16 experiments/superg/doc/paper/paper.tex
prototype equation for strict algo.

@@
 \urlstyle{same}
+
+\usepackage{amsmath}
+\DeclareMathOperator{\strict}{strict}
 
@@
 \label{sec:kiselyov}
-Describes the algorithms as implementable functions. Both recursive and linear semantic functions are covered (the two algorithms).
+Kiselyov's paper~\cite{kiselyov_lambda_2018} describes three iterations of his non-linear semantic translation, and finally the linear algorithm.
+They are presented as a type system accompanied with operational semantics, with the exception of the $\eta$-optimization given as an OCaml code patch.
+OCaml's type system can represent these algorithms very concisely in tagless-final style, but how should we represent them in a more conventional style, possibly in a language that does not have such a strong type system?
+
+\subsection{Strict algorithm}
+This is based on the rules in figure 6 of Kiselyov's paper.
+We call it \textit{strict} it does not have lazy weakening: most notably it lacks the K-optimization required for full laziness.
+
+The most naive implementation is to perform the type checking, then use the built context to apply the elimination rules and the translated program.
+As Kiselyov already notes in his treatment of the linear algorithm further on in the paper, the translation does not actually care what elements are in the context, just that we keep track of the size of the context.
+This lets us express compilation as a recursive function, as shown in figure~\ref{fig:strict}.
+
+\begin{figure*}[h]
+    \begin{equation*}
+        \begin{array}{l l @{}l l}
+            \strict & z         & {}= 1 \models I                                                         &                                                                                  \\
+            \strict & s e       & {}= n+1 \models (0 \models K) \coprod (n \models c)                     & \text{where } (n \models c) := \strict e                                         \\
+            \strict & \lambda e & {}= (0 \models K c)                                                     & \text{where } (0 \models c) := \strict e                                         \\
+            \strict & \lambda e & {}= (n-1 \models K c)                                                   & \text{where } (n \models c) := \strict e                                         \\
+            \strict & e_1 \ e_2 & {}= (\max n_1 n_2 \models ((n_1 \models c_1) \coprod (n_2 \models c_2)) & \text{where } (n_1 \models c_1) := \strict e_1; (n_2 \models c_2) := \strict e_2 \\
+        \end{array}
+    \end{equation*}
+    \caption{Compilation function for the strict algorithm. We have reused $\models$ from the paper. In code one may represent it as a tuple, e.g. $n \models e$ becomes $(n, e)$.}
+    \label{fig:strict}
+\end{figure*}
 

### 2022-12-16 experiments/superg/doc/paper/paper.tex
write about lazy weakening.

@@
 \DeclareMathOperator{\strict}{strict}
+\DeclareMathOperator{\lazy}{lazy}
 
@@
 As Kiselyov already notes in his treatment of the linear algorithm further on in the paper, the translation does not actually care what elements are in the context, just that we keep track of the size of the context.
-This lets us express compilation as a recursive function, as shown in figure~\ref{fig:strict}.
+This lets us express compilation as a recursive function:
 
-\begin{figure*}[h]
 \begin{equation*}
-        \begin{array}{l l @{}l l}
-            \strict & z         & {}= 1 \models I                                                         &                                                                                  \\
-            \strict & s e       & {}= n+1 \models (0 \models K) \coprod (n \models c)                     & \text{where } (n \models c) := \strict e                                         \\
-            \strict & \lambda e & {}= (0 \models K c)                                                     & \text{where } (0 \models c) := \strict e                                         \\
-            \strict & \lambda e & {}= (n-1 \models K c)                                                   & \text{where } (n \models c) := \strict e                                         \\
-            \strict & e_1 \ e_2 & {}= (\max n_1 n_2 \models ((n_1 \models c_1) \coprod (n_2 \models c_2)) & \text{where } (n_1 \models c_1) := \strict e_1; (n_2 \models c_2) := \strict e_2 \\
+    \begin{array}{l l l l}
+        \strict & z         & = & 1 \models I                                              \\
+        \strict & s \ e     & = & n+1 \models (0 \models K) \coprod (n \models c)          \\
+                &           &   & \text{where } (n \models c) := \strict e                 \\
+        \\
+        \strict & \lambda e & = & 0 \models K \; c                                         \\
+                &           &   & \text{where } (0 \models c) := \strict e                 \\
+        \strict & \lambda e & = & (n-1) \models K \; c                                     \\
+                &           &   & \text{where } (n \models c) := \strict e                 \\
+        \\
+        \strict & e_1 \ e_2 & = & (n \models ((n_1 \models c_1) \coprod (n_2 \models c_2)) \\
+                &           &   & \text{where } (n_1 \models c_1) := \strict e_1           \\
+                &           &   & \text{where } (n_2 \models c_2) := \strict e_2           \\
+                &           &   & \text{where } n := \max n_1 \, n_2                       \\
     \end{array}
 \end{equation*}
-    \caption{Compilation function for the strict algorithm. We have reused $\models$ from the paper. In code one may represent it as a tuple, e.g. $n \models e$ becomes $(n, e)$.}
-    \label{fig:strict}
-\end{figure*}
+
+There are two cases for $\lambda e$, and this is precisely where the number of elements in the context matters:
+if the context is empty, the first rule applies, otherwise, we use the second one.
+The semantic function may be defined as:
+
+\begin{equation*}
+    \arraycolsep=1.4pt
+    \begin{array}{l l l l l}
+        (0 \models c_1)   & \coprod & (0 \models c_2)   & = & c_1 \ c_2                                         \\
+        \\
+        (0 \models c_1)   & \coprod & (n_2 \models c_2) & = & (0 \models B c_1) \coprod (n_2 - 1 \models c_2)   \\
+        \\
+        (n_1 \models c_1) & \coprod & (0 \models c_2)   & = & (0 \models C C c_2) \coprod (n_1 - 1 \models c_1) \\
+        \\
+        (n_1 \models c_1) & \coprod & (n_2 \models c_2) & = & (n_1 - 1 \models (s \coprod l)) \coprod r)        \\
+                          &         &                   &   & \text{where } s := (0 \models S )                 \\
+                          &         &                   &   & \text{where } l := (n_1-1 \models c_1 )           \\
+                          &         &                   &   & \text{where } r := (n_2-1 \models c_2 )           \\
+    \end{array}
+\end{equation*}
+
+These simplified functions can be implemented directly and efficiently in a conventional programming language.
+
+\subsection{Lazy weakening}
+Adding lazy weakening requires that we track which items in the context are ignored.
+We change our context representation from an integer to a list of boolean values.
+In the new compilation functions below, we use a shorthand $t$ for the value \textit{true} and $f$ for \textit{false}.
+We always match the last element of lists and represent the remainder of the list with $\Gamma$.
+
+\begin{equation*}
+    \begin{array}{l l l l}
+        \lazy & z         & = & t \models I                                                             \\
+        \lazy & s \ e     & = & \Gamma,f \models c                                                      \\
+              &           &   & \text{where } (\Gamma \models c) := \lazy e                             \\
+        \\
+        \lazy & \lambda e & = & \emptyset \models K \; c                                                \\
+              &           &   & \text{where } (\emptyset \models c) := \lazy e                          \\
+        \lazy & \lambda e & = & (\emptyset \models K) \coprod (\Gamma \models c)                        \\
+              &           &   & \text{where } (\Gamma,f \models c) := \lazy e                           \\
+        \lazy & \lambda e & = & \Gamma \models K \; c                                                   \\
+              &           &   & \text{where } (\Gamma,t \models c) := \lazy e                           \\
+        \\
+        \lazy & e_1 \ e_2 & = & (\Gamma \models ((\Gamma_1 \models c_1) \coprod (\Gamma_2 \models c_2)) \\
+              &           &   & \text{where } (\Gamma_1 \models c_1) := \lazy e_1                       \\
+              &           &   & \text{where } (\Gamma_2 \models c_2) := \lazy e_2                       \\
+              &           &   & \text{where } \Gamma := \Gamma_1 \sqcup \Gamma_2                        \\
+    \end{array}
+\end{equation*}
+
+We have introduced yet another case for $\lambda e$, this time to distinguish between $t$ or $f$ as the last element in the context of the inner expression.
+There is also the new function $\sqcup$, which we need to implement:
+
+\begin{equation*}
+    \begin{array}{l l l l l}
+        \Gamma_1    & \sqcup & \emptyset   & = & \Gamma_1                     \\
+        \emptyset   & \sqcup & \Gamma_2    & = & \Gamma_2                     \\
+        \Gamma_1,t  & \sqcup & \Gamma_2,\_ & = & (\Gamma_1 \sqcup \Gamma_2),t \\
+        \Gamma_1,\_ & \sqcup & \Gamma_2,t  & = & (\Gamma_1 \sqcup \Gamma_2),t \\
+        \Gamma_1,f  & \sqcup & \Gamma_2,f  & = & (\Gamma_1 \sqcup \Gamma_2),f \\
+    \end{array}
+\end{equation*}
+
+In our implementation, we unroll the recursion into a while loop, and repeatedly pop elements from the two arrays until both are empty.
+This builds the new context backwards, so as a final step the array is reversed.
+
+Alternatively, an implementation can pad the shorter list (at the start!) with $f$ and zip them.
+This is particularly convenient for languages that provide a zip function with a default value\footnote{such as \href{https://docs.python.org/3/library/itertools.html\#itertools.zip\_longest}{\texttt{ziplongest}} in Python.}.
+
+Our semantic function now has to handle more cases, and thus is somewhat larger:
+
+\begin{equation*}
+    \arraycolsep=1.4pt
+    \begin{array}{l l l l l}
+        (\emptyset \models c_1)  & \coprod & (\emptyset \models c_2)  & = & c_1 \ c_2                                                \\
+        \\
+        (\emptyset \models c_1)  & \coprod & (\Gamma_2,t \models c_2) & = & (\emptyset \models B c_1) \coprod (\Gamma_2 \models c_2) \\
+        \\
+        (\Gamma_1,t \models c_1) & \coprod & (\emptyset \models c_2)  & = & (\emptyset \models C C c_2) \coprod (\Gamma \models c_1) \\
+        \\
+        (\Gamma_1,t \models c_1) & \coprod & (\Gamma_2,t \models c_2) & = & (\Gamma_1 \models (s \coprod l)) \coprod r)              \\
+                                 &         &                          &   & \text{where } s := (\emptyset \models S )                \\
+                                 &         &                          &   & \text{where } l := (\Gamma_1 \models c_1 )               \\
+                                 &         &                          &   & \text{where } r := (\Gamma_2 \models c_2 )               \\
+        \\
+        % The new additions
+        (\Gamma_1,f \models c_1) & \coprod & (\Gamma_2,f \models c_2) & = & (\Gamma_1 \models c_1) \coprod (\Gamma_2 \models c_2)    \\
+        \\
+        (\Gamma_1,f \models c_1) & \coprod & (\Gamma_2,t \models c_2) & = & (\Gamma_1 \models (b \coprod l)) \coprod r)              \\
+                                 &         &                          &   & \text{where } b := (\emptyset \models B )                \\
+                                 &         &                          &   & \text{where } l := (\Gamma_1 \models c_1 )               \\
+                                 &         &                          &   & \text{where } r := (\Gamma_2 \models c_2 )               \\
+        \\
+        (\Gamma_1,t \models c_1) & \coprod & (\Gamma_2,f \models c_2) & = & (\Gamma_1 \models (c \coprod l)) \coprod r)              \\
+                                 &         &                          &   & \text{where } c := (\emptyset \models C )                \\
+                                 &         &                          &   & \text{where } l := (\Gamma_1 \models c_1 )               \\
+                                 &         &                          &   & \text{where } r := (\Gamma_2 \models c_2 )               \\
+        \\
+        (\emptyset \models c_1)  & \coprod & (\Gamma_2,f \models c_2) & = & (\emptyset \models c_1) \coprod (\Gamma_2 \models c_2)   \\
+        \\
+        (\Gamma_1,f \models c_1) & \coprod & (\emptyset \models c_2)  & = & (\Gamma_1 \models c_1) \coprod (\emptyset \models c_2)   \\
+    \end{array}
+\end{equation*}
+
+The first 4 cases are direct translations from the previous definitions, the others are new additions to handle ignored elements.
+The last 2 cases are not included in the original paper, but they are necessary to make the pattern matching exhaustive, and are present in the reference OCaml implementation.
+
+\textbf{TODO:} Eta optimization.
 

### 2022-12-17 experiments/superg/doc/paper/paper.tex
Write on eta optimization.

@@
 \DeclareMathOperator{\lazy}{lazy}
+\DeclareMathOperator{\lazyeta}{lazy_\eta}
 
@@
 
-\textbf{TODO:} Eta optimization.
+\subsection{Eta optimization}
+Kiselyov covers eta optimization as a change to his OCaml code rather than typing rules.
+Our recursive functions are derived from the typing rules, so these changes are not as straightforward to incorporate.
+%Previously Kiselyov's representation consisted of elements $C$, $N$ and $W$, representing closed expressions, a used context element, and an ignored context element respectively.
+The introduction of the $V$ element means that a context may now be associated with either a closed expression or the top variable (we reuse $V$ as a marker for this).
+Our implementation uses a tagged union to define a type that can be either an expression or the top variable.
+For the compilation function, we only need to change the definition of $\lazyeta \ z$ and add an extra case for $\lazyeta \ \lambda e$:
+
+\begin{equation*}
+    \begin{array}{l l l l}
+        \lazyeta & z         & = & \emptyset \models V                                                     \\
+        \lazyeta & s \ e     & = & \Gamma,f \models c                                                      \\
+                 &           &   & \text{where } (\Gamma \models c) := \lazyeta e                          \\
+        \\
+        \lazyeta & \lambda e & = & \emptyset \models I                                                     \\
+                 &           &   & \text{where } (\emptyset \models V) := \lazyeta e                       \\
+        \lazyeta & \lambda e & = & \emptyset \models K \; c                                                \\
+                 &           &   & \text{where } (\emptyset \models c) := \lazyeta e                       \\
+        \lazyeta & \lambda e & = & (\emptyset \models K) \coprod (\Gamma \models c)                        \\
+                 &           &   & \text{where } (\Gamma,f \models c) := \lazyeta e                        \\
+        \lazyeta & \lambda e & = & \Gamma \models K \; c                                                   \\
+                 &           &   & \text{where } (\Gamma,t \models c) := \lazyeta e                        \\
+        \\
+        \lazyeta & e_1 \ e_2 & = & (\Gamma \models ((\Gamma_1 \models c_1) \coprod (\Gamma_2 \models c_2)) \\
+                 &           &   & \text{where } (\Gamma_1 \models c_1) := \lazyeta e_1                    \\
+                 &           &   & \text{where } (\Gamma_2 \models c_2) := \lazyeta e_2                    \\
+                 &           &   & \text{where } \Gamma := \Gamma_1 \sqcup \Gamma_2                        \\
+    \end{array}
+\end{equation*}
+
+\textbf{TODO: } Check this, possibly need to take (C d, V) = N (C d)
+The context merge function $\sqcup$ is unchanged, but we do have some new cases for the semantic function:
+
+\begin{equation*}
+    \arraycolsep=1.4pt
+    \begin{array}{l l l l l}
+        % Added for eta optimization
+        (\Gamma_1,f \models c_1) & \coprod & (\emptyset \models V)    & = & c_1                                                      \\
+        (\emptyset \models V)    & \coprod & (\Gamma_2,f \models c_2) & = & (\emptyset \models CI) \coprod (\Gamma_2 \models c_2)    \\
+
+        (\Gamma_1,t \models c_1) & \coprod & (\emptyset \models V)    & = & (\Gamma_1 \models (s \coprod l)) \coprod i               \\
+                                 &         &                          &   & \text{where } l := (\Gamma_1 \models c_1 )               \\
+                                 &         &                          &   & \text{where } s := (\emptyset \models S )                \\
+                                 &         &                          &   & \text{where } i := (\emptyset \models I )                \\
+        (\emptyset \models V)    & \coprod & (\Gamma_2,t \models c_2) & = & (\emptyset \models SI) \coprod (\Gamma_2 \models c_2)    \\
+        (\emptyset \models c_1)  & \coprod & (\emptyset \models V)    & = & c_1                                                      \\
+        (\emptyset \models V)    & \coprod & (\emptyset \models c_2)  & = & CI c_2                                                   \\
+        \\
+
+        % Original
+        (\emptyset \models c_1)  & \coprod & (\emptyset \models c_2)  & = & c_1 \ c_2                                                \\
+        \\
+        (\emptyset \models c_1)  & \coprod & (\Gamma_2,t \models c_2) & = & (\emptyset \models B c_1) \coprod (\Gamma_2 \models c_2) \\
+        \\
+        (\Gamma_1,t \models c_1) & \coprod & (\emptyset \models c_2)  & = & (\emptyset \models C C c_2) \coprod (\Gamma \models c_1) \\
+        \\
+        (\Gamma_1,t \models c_1) & \coprod & (\Gamma_2,t \models c_2) & = & (\Gamma_1 \models (s \coprod l)) \coprod r)              \\
+                                 &         &                          &   & \text{where } s := (\emptyset \models S )                \\
+                                 &         &                          &   & \text{where } l := (\Gamma_1 \models c_1 )               \\
+                                 &         &                          &   & \text{where } r := (\Gamma_2 \models c_2 )               \\
+        \\
+        % Lazy
+        (\Gamma_1,f \models c_1) & \coprod & (\Gamma_2,f \models c_2) & = & (\Gamma_1 \models c_1) \coprod (\Gamma_2 \models c_2)    \\
+        \\
+        (\Gamma_1,f \models c_1) & \coprod & (\Gamma_2,t \models c_2) & = & (\Gamma_1 \models (b \coprod l)) \coprod r)              \\
+                                 &         &                          &   & \text{where } b := (\emptyset \models B )                \\
+                                 &         &                          &   & \text{where } l := (\Gamma_1 \models c_1 )               \\
+                                 &         &                          &   & \text{where } r := (\Gamma_2 \models c_2 )               \\
+        \\
+        (\Gamma_1,t \models c_1) & \coprod & (\Gamma_2,f \models c_2) & = & (\Gamma_1 \models (c \coprod l)) \coprod r)              \\
+                                 &         &                          &   & \text{where } c := (\emptyset \models C )                \\
+                                 &         &                          &   & \text{where } l := (\Gamma_1 \models c_1 )               \\
+                                 &         &                          &   & \text{where } r := (\Gamma_2 \models c_2 )               \\
+        \\
+        (\emptyset \models c_1)  & \coprod & (\Gamma_2,f \models c_2) & = & (\emptyset \models c_1) \coprod (\Gamma_2 \models c_2)   \\
+        \\
+        (\Gamma_1,f \models c_1) & \coprod & (\emptyset \models c_2)  & = & (\Gamma_1 \models c_1) \coprod (\emptyset \models c_2)   \\
+    \end{array}
+\end{equation*}
 

### 2022-12-17 experiments/superg/doc/paper/paper.tex
Cleanup eta optimization.

@@
                  &           &   & \text{where } (\emptyset \models c) := \lazyeta e                            \\
-        \lazyeta & \lambda e & = & (\emptyset \models K) \coprod (\Gamma \models c)                        \\
+        \lazyeta & \lambda e & = & (\emptyset \models K) \coprod_\eta (\Gamma \models c)                        \\
                  &           &   & \text{where } (\Gamma,f \models c) := \lazyeta e                             \\
@@
         \\
-        \lazyeta & e_1 \ e_2 & = & (\Gamma \models ((\Gamma_1 \models c_1) \coprod (\Gamma_2 \models c_2)) \\
+        \lazyeta & e_1 \ e_2 & = & (\Gamma \models ((\Gamma_1 \models c_1) \coprod_\eta (\Gamma_2 \models c_2)) \\
                  &           &   & \text{where } (\Gamma_1 \models c_1) := \lazyeta e_1                         \\
@@
         % Added for eta optimization
-        (\Gamma_1,f \models c_1) & \coprod & (\emptyset \models V)    & = & c_1                                                      \\
-        (\emptyset \models V)    & \coprod & (\Gamma_2,f \models c_2) & = & (\emptyset \models CI) \coprod (\Gamma_2 \models c_2)    \\
+        (\Gamma_1,f \models c_1) & \coprod_\eta & (\emptyset \models V)    & = & c_1                                                        \\
+        (\emptyset \models V)    & \coprod_\eta & (\Gamma_2,f \models c_2) & = & (\emptyset \models CI) \coprod_\eta (\Gamma_2 \models c_2) \\
 
-        (\Gamma_1,t \models c_1) & \coprod & (\emptyset \models V)    & = & (\Gamma_1 \models (s \coprod l)) \coprod i               \\
+        (\Gamma_1,t \models c_1) & \coprod_\eta & (\emptyset \models V)    & = & (\Gamma_1 \models (s \coprod_\eta l)) \coprod_\eta i       \\
                                  &              &                          &   & \text{where } l := (\Gamma_1 \models c_1 )                 \\
@@
                                  &              &                          &   & \text{where } i := (\emptyset \models I )                  \\
-        (\emptyset \models V)    & \coprod & (\Gamma_2,t \models c_2) & = & (\emptyset \models SI) \coprod (\Gamma_2 \models c_2)    \\
-        (\emptyset \models c_1)  & \coprod & (\emptyset \models V)    & = & c_1                                                      \\
-        (\emptyset \models V)    & \coprod & (\emptyset \models c_2)  & = & CI c_2                                                   \\
+        (\emptyset \models V)    & \coprod_\eta & (\Gamma_2,t \models c_2) & = & (\emptyset \models SI) \coprod_\eta (\Gamma_2 \models c_2) \\
+        (\emptyset \models c_1)  & \coprod_\eta & (\emptyset \models V)    & = & c_1                                                        \\
+        (\emptyset \models V)    & \coprod_\eta & (\emptyset \models c_2)  & = & CI c_2                                                     \\
+        (\emptyset \models V)    & \coprod_\eta & (\emptyset \models V)    & = & \mathbf{impossible}                                        \\
         \\
 
-        % Original
-        (\emptyset \models c_1)  & \coprod & (\emptyset \models c_2)  & = & c_1 \ c_2                                                \\
-        \\
-        (\emptyset \models c_1)  & \coprod & (\Gamma_2,t \models c_2) & = & (\emptyset \models B c_1) \coprod (\Gamma_2 \models c_2) \\
-        \\
-        (\Gamma_1,t \models c_1) & \coprod & (\emptyset \models c_2)  & = & (\emptyset \models C C c_2) \coprod (\Gamma \models c_1) \\
-        \\
-        (\Gamma_1,t \models c_1) & \coprod & (\Gamma_2,t \models c_2) & = & (\Gamma_1 \models (s \coprod l)) \coprod r)              \\
-                                 &         &                          &   & \text{where } s := (\emptyset \models S )                \\
-                                 &         &                          &   & \text{where } l := (\Gamma_1 \models c_1 )               \\
-                                 &         &                          &   & \text{where } r := (\Gamma_2 \models c_2 )               \\
-        \\
-        % Lazy
-        (\Gamma_1,f \models c_1) & \coprod & (\Gamma_2,f \models c_2) & = & (\Gamma_1 \models c_1) \coprod (\Gamma_2 \models c_2)    \\
-        \\
-        (\Gamma_1,f \models c_1) & \coprod & (\Gamma_2,t \models c_2) & = & (\Gamma_1 \models (b \coprod l)) \coprod r)              \\
-                                 &         &                          &   & \text{where } b := (\emptyset \models B )                \\
-                                 &         &                          &   & \text{where } l := (\Gamma_1 \models c_1 )               \\
-                                 &         &                          &   & \text{where } r := (\Gamma_2 \models c_2 )               \\
-        \\
-        (\Gamma_1,t \models c_1) & \coprod & (\Gamma_2,f \models c_2) & = & (\Gamma_1 \models (c \coprod l)) \coprod r)              \\
-                                 &         &                          &   & \text{where } c := (\emptyset \models C )                \\
-                                 &         &                          &   & \text{where } l := (\Gamma_1 \models c_1 )               \\
-                                 &         &                          &   & \text{where } r := (\Gamma_2 \models c_2 )               \\
-        \\
-        (\emptyset \models c_1)  & \coprod & (\Gamma_2,f \models c_2) & = & (\emptyset \models c_1) \coprod (\Gamma_2 \models c_2)   \\
-        \\
-        (\Gamma_1,f \models c_1) & \coprod & (\emptyset \models c_2)  & = & (\Gamma_1 \models c_1) \coprod (\emptyset \models c_2)   \\
+        (\Gamma_1 \models c_1)   & \coprod_\eta & (\Gamma_2 \models)       & = & (\Gamma_1 \models c_1) \coprod (\Gamma_2 \models)          \\
     \end{array}
 \end{equation*}
+
+The case $(\emptyset \models V) \coprod (\emptyset \models V) $ is impossible in this translation, implementations may throw an error or leave it as undefined behavior.
+\textbf{TODO:} Why is this an impossible case?
 

### 2022-12-18 experiments/superg/doc/paper/paper.tex
Note on folding ctx merge into semantic.

@@
                  &           &   & \text{where } (\Gamma_2 \models c_2) := \lazyeta e_2                         \\
-                 &           &   & \text{where } \Gamma := \Gamma_1 \sqcup \Gamma_2                             \\
+                 &           &   & \text{where } \Gamma := \Gamma_1 \sqcup_\eta \Gamma_2                        \\
     \end{array}
@@
 
-\textbf{TODO: } Check this, possibly need to take (C d, V) = N (C d)
-The context merge function $\sqcup$ is unchanged, but we do have some new cases for the semantic function:
+Our context merge function $\sqcup$ now also depends on the compiled expression, since we have a special case for $V$:
+
+\begin{equation*}
+    \begin{array}{l l l l l}
+        (\Gamma_1,\_ \models c_1) & \sqcup_\eta & (\emptyset \models V)     & = & \Gamma_1,t                        \\
+        (\emptyset \models V)     & \sqcup_\eta & (\Gamma_2,\_ \models c_2) & = & \Gamma_2,t                        \\
+        (\emptyset \models c_1)   & \sqcup_\eta & (\emptyset \models V)     & = & t                                 \\
+        (\emptyset \models V)     & \sqcup_\eta & (\emptyset \models c_2)   & = & t                                 \\
+        (\emptyset \models V)     & \sqcup_\eta & (\emptyset \models V)     & = & \mathbf{impossible}               \\
+        \\
+        % Original ones
+        (\Gamma_1 \models c_1)    & \sqcup_\eta & (\emptyset \models c_2)   & = & \Gamma_1                          \\
+        (\emptyset \models c_1)   & \sqcup_\eta & (\Gamma_2 \models c_2)    & = & \Gamma_2                          \\
+        (\Gamma_1,t \models c_1)  & \sqcup_\eta & (\Gamma_2,\_ \models c_2) & = & (\Gamma_1 \sqcup_\eta \Gamma_2),t \\
+        (\Gamma_1,\_ \models c_1) & \sqcup_\eta & (\Gamma_2,t \models c_2)  & = & (\Gamma_1 \sqcup_\eta \Gamma_2),t \\
+        (\Gamma_1,f \models c_1)  & \sqcup_\eta & (\Gamma_2,f \models c_2)  & = & (\Gamma_1 \sqcup_\eta \Gamma_2),f \\
+    \end{array}
+\end{equation*}
+
+The $V$ value effectively encodes an additional implicit used element in the context.
+We also have some new cases for the semantic function:
 
@@
 \textbf{TODO:} Why is this an impossible case?
+
+With the addition of the expression value to $\sqcup_\eta$, the function now takes the same input as $\coprod_\eta$ and has very similar matching patterns.
+Our implementation, therefore, folds context merging into $\coprod_\eta$ to make the code more concise.
 

### 2022-12-19 experiments/superg/doc/paper/paper.tex
Write about linear algorithm.

@@
 \DeclareMathOperator{\lazyeta}{lazy_\eta}
+\DeclareMathOperator{\linear}{linear}
 
@@
 
+\subsection{Linear algorithm}
+The implementation of the linear algorithm is much simpler because it is based on the strict algorithm.
+The compilation function is identical save for a different semantic function $\coprod_{\text{linear}}$:
+
+\begin{equation*}
+    \begin{array}{l l l l}
+        \linear & z         & = & 1 \models I                                                              \\
+        \linear & s \ e     & = & n+1 \models (0 \models K) \coprod (n \models c)                          \\
+                &           &   & \text{where } (n \models c) := \linear e                                 \\
+        \\
+        \linear & \lambda e & = & 0 \models K \; c                                                         \\
+                &           &   & \text{where } (0 \models c) := \linear e                                 \\
+        \linear & \lambda e & = & (n-1) \models K \; c                                                     \\
+                &           &   & \text{where } (n \models c) := \linear e                                 \\
+        \\
+        \linear & e_1 \ e_2 & = & (n \models ((n_1 \models c_1) \coprod_{\text{linear}} (n_2 \models c_2)) \\
+                &           &   & \text{where } (n_1 \models c_1) := \linear e_1                           \\
+                &           &   & \text{where } (n_2 \models c_2) := \linear e_2                           \\
+                &           &   & \text{where } n := \max n_1 \, n_2                                       \\
+    \end{array}
+\end{equation*}
+
+below we give the corresponding semantic function, which is no longer recursive, making this algorithm linear:
+
+\begin{equation*}
+    \arraycolsep=1.4pt
+    \begin{array}{l l l l l}
+        (0 \models c_1)   & \coprod & (0 \models c_2)   & = & c_1 \ c_2                               \\
+        \\
+        (0 \models c_1)   & \coprod & (n \models c_2)   & = & B_n \ c_1 \ c_2                         \\
+        \\
+        (n \models c_1)   & \coprod & (0 \models c_2)   & = & C_n \ c_1 \ c_2                         \\
+        \\
+        (n \models c_1)   & \coprod & (n \models c_2)   & = & S_n \ c_1 \ c_2                         \\
+        \\
+        (n_1 \models c_1) & \coprod & (n_2 \models c_2) & = & B_{n_2-n_1} \ (S_{n_1} \ c_1) \ c_2     \\
+                          &         &                   &   & \text{where } n_1 < n_2                 \\
+        (n_1 \models c_1) & \coprod & (n_2 \models c_2) & = & C_{n_1-n_2} \ (B_{n_1-n_2} \ c_1) \ c_2 \\
+                          &         &                   &   & \text{where } n_1 > n_2                 \\
+    \end{array}
+\end{equation*}
+
+\textbf{TODO: } Can we add lazy weakening and/or eta optimization?
+
 \section{Engine implementation}

### 2022-12-19 experiments/superg/doc/paper/paper.tex
Write up much of miranda impl.

@@
 \subsection*{Miranda}
-Describe the implementation of a SASL/Miranda-style engine.
+For our base implementation, we try to stick to Miranda's design as much as possible.
+While there is published work on Miranda~\cite{turner_miranda_1985}, it does not provide much detail on the reduction engine.
+Fortunately, the source code for Miranda is now freely available~\cite{noauthor_open_2021}, and from the code, we can deduce that its design is very similar to SASL~\cite{turner_new_1979}, an earlier language developed by Turner.
 
-Compare performance with Open Source Miranda.
+Miranda uses a graph reduction engine, so after a program has been converted into combinatory logic using Turner's abstraction algorithm~\cite{turner_another_1979}, it is loaded into a graph structure.
+The graph is binary: every node has exactly two children, and the node itself represents the application of the right node to the left one.
+Let us start with an example program $SKK \ 3$.
+The $S$ combinator takes 3 arguments, but currying lets us write it as $((SK) \; K)\ 3$.
+In this form, it is straightforward to construct the associated binary graph, shown in figure~\ref{fig:skk3}.
+
+\begin{figure}
+    \includegraphics[width=.23\columnwidth]{skk3}
+    \centering
+    \caption{$SKK \ 3$ in graph representation.}
+    \label{fig:skk3}
+\end{figure}
+
+To reduce this graph, we follow the left pointers until we find a combinator to apply.
+As we traverse the graph along what is often called the \textit{spine}, we keep the nodes we have seen on a stack, because their right nodes contain the arguments to the combinator we eventually find.
+
+In this example graph, after following two pointers we find the $S$ combinator, and we apply the definition $S f g x \Rightarrow (f x) (g x)$ to rewrite the graph, as shown in figure~\ref{fig:skk3_sred}.
+
+\begin{figure}
+    \includegraphics[width=.8\columnwidth]{skk3_sred}
+    \centering
+    \caption{
+        First reduction step for $SKK \ 3$.
+        One of the K combinators is underlined to make them easier to distinguish.
+    }
+    \label{fig:skk3_sred}
+\end{figure}
+
+The reduction process continues until the program has been reduced to a constant value, or there are insufficient arguments for the next combinator to reduce.
+The expression is said to be in \textit{weak head normal form}: some inner part of the expression may be reducible, but we only perform outermost reduction, so this is as close as we can get to normal form.
+
+\subsection{Sharing}
+In the reduction in figure~\ref{fig:skk3_sred}, we have copied the value $3$ to two nodes.
+For another reduction where the $x$ value is a pointer to a subgraph, performing the reduction may result in multiple pointers to the same expression.
+Expressions in Miranda are \textit{referentially transparent}: we can safely replace any expression with the value it computes.
+Computing $x$ in both branches would therefore be wasteful.
+Instead, we would like to compute it once (in whatever branch runs first) and have the second branch directly reuse the result.
+It turns out that this is not too difficult to implement: when we make a reduction step, we update the outermost node with the contents of the new outermost node.
+In figure~\ref{fig:skk3_sred}, this would mean we overwrite node 1 with the contents of node 4.
+Now any other pointers to node 1 automatically point to the reduced equivalent.
+If the expression reduces to a single value $v$ (say an integer), we can use the $I$ combinator to make an \textit{indirection node} $I v$.
+
+\subsection{Strict operators}
+\textbf{TODO}
+
+\subsection{Representing the graph}
+Miranda represents the graph using three arrays, \texttt{hd} (for head) stores the left pointer, \texttt{tl} (for tail) stores the right pointer, and \texttt{tag} stores metadata for the cell.
+Owing to its early inception when 64-bit machines were rare, \texttt{hd} and \texttt{tl} are 32 bits wide.
+\texttt{tag} only needs to store a few bit flags and thus is a single byte wide.
+Our implementation stores two flags in \texttt{tag}:
+\begin{itemize}
+    \item \texttt{WANTED}: Set if this cell is currently in use.
+Cells without this bit set are considered free and may be overwritten to store new nodes in the graph.
+More on this in the section on Garbage collection.
+    \item \texttt{RHS\_INT}: Set if the right pointer is not a pointer but an integer value.
+Note that the left pointer can never be an integer because the left pointer is always a function for which the right pointer is the first argument.
+\end{itemize}
+
+It is unclear why a representation with three separate arrays was chosen over a single array where each element contains a 'cell' struct with a head, tail and tag.
+We suspect decomposing the graph into separate arrays leads to poor data locality, as almost any operation will access all three parts of a cell.
+Our base implementation respects this design choice, later implementations do not.
+
+Miranda reserves the lower range of cell pointer values for special purposes:
+\begin{itemize}
+    \item 0..255 represents the Latin 1 character set.
+    \item 256..305 for internal use in the lexer.
+    \item 306..446 to encode combinators.
+\end{itemize}
+
+The first 446 cells in the heap are unused: when the reduction engine encounters their value, they are interpreted as a character, lexer rule or combinator.
+This explains why we have to tag integers but not combinators or character values.
+
+\subsection{Garbage collection}
+\textbf{TODO}
 

### 2022-12-19 experiments/superg/doc/paper/paper.tex
Finish up Miranda writing.

@@
 
-\subsection{Strict operators}
-\textbf{TODO}
-
 \subsection{Representing the graph}
@@
 \end{itemize}
+
+Miranda integers have unlimited precision, using GNU MP~\cite{gnump} internally.
+For simplicity (and performance), our implementation instead treats the cell pointer as a 32-bit signed integer.
 
@@
 
+\subsection{Strict combinators}
+Miranda has special built-in combinators to perform arithmetic operations and compare numbers.
+Where most combinators place no restrictions on their arguments, these combinators require their arguments to be fully reduced to integers (hence call them \textit{strict}), and so require special treatment.
+We found Miranda's implementation difficult to follow, the approach we describe here is the one taken for our implementation, and may not fully correspond to Miranda's.
+
+When a \textit{strict} combinator is about to be reduced, the engine inspects the arguments.
+If all arguments are reduced, the combinator reduction is performed as usual.
+If one or more arguments are not yet reduced, we push the pointers to those arguments onto the stack and proceed to reduce from the top of the stack.
+Eventually, the top argument will have been fully reduced, producing an integer value.
+In general, if a reduction results in an integer value, it can mean one of two things:
+\begin{itemize}
+    \item The value is the result of the program.
+    \item The value is one of the arguments for a strict combinator.
+\end{itemize}
+We distinguish between these cases by looking at the size of the stack: if it is empty, we have computed the final result of the program.
+Otherwise, we continue reducing from the top of the stack, which will be either another argument or the original combinator expression.
+The pointer to the original complex argument is replaced with an indirection node to the value, using the update logic described earlier.
+
+As an example, consider the expression $+ \ 1 \ (K \ 2 \ 3)$.
+After we have traversed the spine and encountered the $+$ combinator, the engine notices that the second argument is not yet reduced, and adds it to the stack.
+Figure~\ref{fig:add_step1} shows the state of the graph at this point in the reduction process.
+
+\begin{figure}
+    \includegraphics[width=.8\columnwidth]{add_step1}
+    \centering
+    \caption{
+        Graph of $+ \ 1 \ (K \ 2 \ 3)$ after the second argument has been added to the stack.
+    }
+    \label{fig:add_step1}
+\end{figure}
+
+Once $K \ 2 \ 3$ has been reduced to $3$, the engines observes that the stack is not yet empty, so it creates an indirection node, and resumes reduction from the top of the stack.
+Figure~\ref{fig:add_step2} reflects this state.
+
+\begin{figure}
+    \includegraphics[width=.8\columnwidth]{add_step2}
+    \centering
+    \caption{
+        Graph of $+ \ 1 \ (K \ 2 \ 3)$ after the second argument has been reduced.
+    }
+    \label{fig:add_step2}
+\end{figure}
+
+This example shows that it is not sufficient to check the \texttt{tag} of each argument to see if it is an integer: indirection nodes also count as fully reduced.
+
 \subsection{Garbage collection}
-\textbf{TODO}
+The graph reduction process may lead to some nodes becoming unreachable.
+Coming back to the example in figure~\ref{fig:skk3_sred}, after the reduction step nodes 2 and 3 are no longer used: they are \textit{garbage}.
+In this example, it is easy to conclude which nodes are garbage, but in larger programs, there are often multiple references to the same node, so we cannot mark nodes as garbage during the reduction step.
+Miranda employs a simple mark-and-sweep garbage collector that kicks in when the engine runs out of free cells while allocating a new node.
+In the first phase, it zeroes out the \texttt{WANTED} bit on all tags in the heap.
+Then in the marking phase, starting from the elements currently on the stack\footnote{
+    Miranda keeps its reduction stack on the C stack.
+    This makes it difficult to reliably enumerate all cell pointers on the stack at a given point in time, because the compiler may choose to keep some of the pointers in registers.
+    For this reason, users are instructed not to enable optimizations when compiling Miranda, which hurts performance but is necessary for correctness.
+    When Miranda was developed, compilers were not advanced enough to perform such optimizations, so this was not an issue.
+}, the garbage collector traverses the graph and sets the \texttt{WANTED} bit on all reachable nodes.
+There is no real sweep phase: instead, the allocation routine walks over the tag array until it finds a cell that is not marked as \texttt{WANTED}.
+This also means that there is no list of currently free cells, which we suspect may lead to performance issues in large heaps.
+
+We consider the garbage collector in Miranda to be somewhat simplistic and inefficient, and we suspect there is much to improve in terms of efficiency.
+Our benchmark programs do not use too much heap space, so we choose not to focus our efforts on this, and defer optimizations on this front to future work.
 

### 2022-12-19 experiments/superg/doc/paper/paper.tex
Split miranda and tigre into sections.

@@
     \item We implement a fast combinator graph reduction engine based on the ideas in Miranda~\cite{turner_miranda_1985} and later improvements~\cite{koopman_architecture_1992}.
-          We elaborate on its design in section~\ref{sec:engine}.
+          We elaborate on its design in sections~\ref{sec:miranda} and~\ref{sec:tigre}.
     \item We evaluate the performance of Miranda, Haskell and our engine on a set of benchmark programs in section~\ref{sec:eval}.
@@
 
-\section{Engine implementation}
-\label{sec:engine}
-\subsection*{Miranda}
+\section{Miranda}
+\label{sec:miranda}
 For our base implementation, we try to stick to Miranda's design as much as possible.
@@
 
-\subsection*{TIGRE}
+\section{TIGRE}
+\label{sec:tigre}
+\textbf{TODO}
+
 Describe changes made to align with TIGRE.

### 2022-12-19 experiments/superg/doc/paper/paper.tex
Combinator encoding.

@@
 \subsection{Linear algorithm}
-The implementation of the linear algorithm is much simpler because it is based on the strict algorithm.
+The implementation of the linear algorithm is much simpler because it is based on the $\strict$ algorithm.
 The compilation function is identical save for a different semantic function $\coprod_{\text{linear}}$:
@@
 
+Our implementation does not support character values and needs no reserved number for the lexer, so we can encode the first combinator as $0$.
+The bulk combinators introduced by the $\linear$ algorithm do complicate the encoding somewhat:
+in principle, there is now an infinite amount of possible combinators.
+However, compiled programs only contain a finite subset of those combinators, and the reduction process only creates combinators that were present in the original program.
+We can have an efficient encoding of combinators, but only if we generate it for specific programs.
+Our implementation keeps an array of function pointers, each pointing to the code to reduce a particular combinator.
+The array is populated as the compiled program is loaded into the graph representation:
+when we encounter a combinator we have not seen before, its implementation is added to the array, and the combinator is encoded into the graph as its index into the array.
+At runtime, combinator resolution is done by indexing into the array and calling the associated function.
+
 \subsection{Strict combinators}
@@
 
-Once $K \ 2 \ 3$ has been reduced to $3$, the engines observes that the stack is not yet empty, so it creates an indirection node, and resumes reduction from the top of the stack.
+Once $K \ 2 \ 3$ has been reduced to $3$, the engine observes that the stack is not yet empty, so it creates an indirection node and resumes reduction from the top of the stack.
 Figure~\ref{fig:add_step2} reflects this state.

### 2022-12-19 experiments/superg/src/turner/mod.rs
Separate functions per combinator.

@@ -118,78 +118,24 @@ impl TurnerEngine {
         let top = self.stack.last().unwrap();
         if let Some(comb) = top.comb() {
             match comb {
-                Comb::S => self.run_comb3(|engine, args| {
-                    // S f g x => f x (g x)
-                    let x = engine.tl(args[2]);
-                    let g = engine.tl(args[1]);
-                    let f = engine.tl(args[0]);
-                    // If x is an int, we should transfer that to the new location
-                    let mut tag = Tag::wanted();
-                    if engine.tag(args[2]).is_rhs_int() {
-                        tag = tag.set_rhs_int();
-                    }
-                    // Make lower cells
-                    let left_cell = engine.make_cell(tag, f, x);
-                    let right_cell = engine.make_cell(tag, g, x);
-                    StepResult::CellContents(Tag::wanted(), left_cell, right_cell)
-                }),
-                Comb::K => self.run_comb2(|engine, args| {
-                    // K x y = x
-                    if let Some(v) = engine.int_rhs(args[0]) {
-                        StepResult::Value(v)
-                    } else {
-                        StepResult::Cell(engine.tl(args[0]))
-                    }
-                }),
-                Comb::I => self.run_comb1(|engine, args| {
-                    // I x = x
-                    if let Some(v) = engine.int_rhs(args[0]) {
-                        StepResult::Value(v)
-                    } else {
-                        StepResult::Cell(engine.tl(args[0]))
-                    }
-                }),
-                Comb::Y => todo!("Y not implemented. Stack: {:?}", self.stack),
+                Comb::S => self.run_s_comb(),
+                Comb::K => self.run_k_comb(),
+                Comb::I => self.run_i_comb(),
+                Comb::Y => todo!(),
                 Comb::U => todo!(),
                 Comb::P => todo!(),
                 Comb::B => todo!(),
                 Comb::C => todo!(),
-                Comb::Plus => self.run_strict_binop(|a, b| a + b),
-                Comb::Minus => self.run_strict_binop(|a, b| a - b),
-                Comb::Times => self.run_strict_binop(|a, b| a * b),
-                Comb::Divide => self.run_strict_binop(|a, b| a / b),
-                Comb::Cond => self.run_comb3(|engine, args| {
-                    // COND c t f = if(c) t else f
-                    if let Some(c) = engine.int_rhs(args[0]) {
[…]

### 2022-12-19 experiments/superg/src/compiled_expr.rs
remove repr(i32) from Comb.

@@ -1,5 +1,4 @@
-#[derive(Debug, Clone, Copy, PartialEq)]
-#[repr(i32)]
+#[derive(Debug, Clone, Copy, PartialEq, Eq, Hash)]
 pub enum Comb {
     S,
     K,

### 2022-12-19 experiments/superg/src/turner/mod.rs

@@ -13,14 +13,36 @@ const CELLS: usize = 250_000;
 pub struct CellPtr(i32);
 
 impl CellPtr {
-    pub fn comb(&self) -> Option<Comb> {
-        Comb::all().get(self.0 as usize).copied()
+    pub fn comb(&self, engine: &TurnerEngine) -> Option<Comb> {
+        engine.comb_inst.get(self.0 as usize).copied()
+    }
+
+    pub fn is_comb(&self, engine: &TurnerEngine) -> bool {
+        (self.0 as usize) < engine.comb_impl.len()
     }
 }
 
-impl Into<CellPtr> for Comb {
-    fn into(self) -> CellPtr {
-        CellPtr(self as i32)
+trait IntoCellPtr {
+    fn into_cell_ptr(self, engine: &TurnerEngine) -> CellPtr;
+}
+
+impl IntoCellPtr for CellPtr {
+    fn into_cell_ptr(self, _: &TurnerEngine) -> CellPtr {
+        self
+    }
+}
+
+impl IntoCellPtr for Comb {
+    fn into_cell_ptr(self, engine: &TurnerEngine) -> CellPtr {
+        CellPtr(engine.comb_map.get(&self).copied().unwrap() as i32)
+    }
+}
+
+impl Comb {
+    fn implementation(&self) -> fn(&mut TurnerEngine) -> Option<CellPtr> {
+        match self {
+            _ => TurnerEngine::run_todo_comb,
+        }
     }
 }
 
@@ -64,7 +86,7 @@ pub struct TurnerEngine {
     // | TAG | HEAD | TAIL |
     // |=====|======|======|
     //
-    // We could store that in a Vec<(Tag, CellPtr,CellPtr), or even define a Cell type, but Miranda choosen this decomposed form, so we stick with that.
+    // We could store that in a Vec<(Tag, CellPtr,CellPtr), or even define a Cell type, but Miranda chooses this decomposed form, so we stick with that.
     // TODO: Experiment with storing tuples instead.
     tag: Vec<Tag>,
     hd: Vec<CellPtr>,
@@ -74,6 +96,13 @@ pub struct TurnerEngine {
     // Points to the cell containing the compiled definition
     def_lookup: HashMap<String, CellPtr>,
     stack: Vec<CellPtr>,
+
+    // Combinator implementation lookup table
+    comb_impl: Vec<fn(&mut TurnerEngine) -> Option<CellPtr>>,
[…]

### 2022-12-19 experiments/superg/src/turner/mod.rs
Get debugger working with new LUT setup.

@@ -13,12 +13,16 @@ const CELLS: usize = 250_000;
 pub struct CellPtr(i32);
 
 impl CellPtr {
-    pub fn comb(&self, engine: &TurnerEngine) -> Option<Comb> {
+    pub fn comb(&self, engine: &TurnerEngineDebug) -> Option<Comb> {
         engine.comb_inst.get(self.0 as usize).copied()
     }
 
     pub fn is_comb(&self, engine: &TurnerEngine) -> bool {
         (self.0 as usize) < engine.comb_impl.len()
+    }
+
+    pub fn is_i_comb(&self) -> bool {
+        return self.0 == 0;
     }
 }
 
@@ -34,14 +38,28 @@ impl IntoCellPtr for CellPtr {
 
 impl IntoCellPtr for Comb {
     fn into_cell_ptr(self, engine: &TurnerEngine) -> CellPtr {
-        CellPtr(engine.comb_map.get(&self).copied().unwrap() as i32)
+        if let Some(idx) = engine.comb_map.get(&self) {
+            CellPtr(*idx as i32)
+        } else {
+            panic!("Combinator {:?} not found in mapping", self);
+        }
     }
 }
 
 impl Comb {
     fn implementation(&self) -> fn(&mut TurnerEngine) -> Option<CellPtr> {
         match self {
-            _ => TurnerEngine::run_todo_comb,
+            Comb::S => TurnerEngine::run_s_comb,
+            Comb::K => TurnerEngine::run_k_comb,
+            Comb::I => TurnerEngine::run_i_comb,
+            Comb::Plus => TurnerEngine::run_plus_comb,
+            Comb::Minus => TurnerEngine::run_minus_comb,
+            Comb::Times => TurnerEngine::run_times_comb,
+            Comb::Cond => TurnerEngine::run_cond_comb,
+            Comb::Eq => TurnerEngine::run_eq_comb,
+            Comb::Lt => TurnerEngine::run_lt_comb,
+            Comb::Abort => TurnerEngine::run_abort_comb,
+            _ => todo!("No implementation for combinator {:?}", self),
         }
     }
 }
@@ -99,10 +117,13 @@ pub struct TurnerEngine {
 
     // Combinator implementation lookup table
     comb_impl: Vec<fn(&mut TurnerEngine) -> Option<CellPtr>>,
-    // Combinator instance lookup table
-    comb_inst: Vec<Comb>,
[…]

### 2022-12-21 experiments/superg/src/compiled_expr.rs
Add bulk combinators to regular Comb.

@@ -23,39 +23,11 @@ pub enum Comb {
     Or,
     Not,
     Abort,
+    // Kiselyov's bulk combinators
+    Sn(usize),
+    Bn(usize),
+    Cn(usize),
 }
-
-impl Comb {
-    pub fn all() -> &'static [Comb] {
-        return ALL_COMBS;
-    }
-}
-
-const ALL_COMBS: &'static [Comb] = &[
-    Comb::S,
-    Comb::K,
-    Comb::I,
-    Comb::Y,
-    Comb::U,
-    Comb::P,
-    Comb::B,
-    Comb::C,
-    Comb::Plus,
-    Comb::Minus,
-    Comb::Times,
-    Comb::Divide,
-    Comb::Cond,
-    Comb::Eq,
-    Comb::Neq,
-    Comb::Gt,
-    Comb::Gte,
-    Comb::Lt,
-    Comb::Lte,
-    Comb::And,
-    Comb::Or,
-    Comb::Not,
-    Comb::Abort,
-];
 
 #[derive(Clone, PartialEq)]
 pub enum CompiledExpr {
@@ -68,7 +40,12 @@ pub enum CompiledExpr {
 impl std::fmt::Debug for CompiledExpr {
     fn fmt(&self, f: &mut std::fmt::Formatter<'_>) -> std::fmt::Result {
         match self {
-            CompiledExpr::Comb(c) => write!(f, "{:?}", c),
+            CompiledExpr::Comb(c) => match c {
+                Comb::Sn(i) => write!(f, "S{}", i),
+                Comb::Bn(i) => write!(f, "B{}", i),
+                Comb::Cn(i) => write!(f, "C{}", i),
+                _ => write!(f, "{:?}", c),
+            },
             CompiledExpr::Ap(a, b) => write!(f, "({:?} {:?})", a, b),
             CompiledExpr::Var(v) => write!(f, "{}", v),
             CompiledExpr::Int(i) => write!(f, "{}", i),

### 2022-12-21 experiments/superg/src/kiselyov/linear.rs

@@ -1,7 +1,7 @@
 /// The strict compilation algorithm, as given in fig.10
 use super::BExpr;
 
-use compiled_expr::*;
+use crate::compiled_expr::*;
 
 fn infer_n(e: &BExpr) -> usize {
     match e {
@@ -62,18 +62,41 @@ pub fn compile_linear(e: &BExpr) -> CompiledExpr {
 }
 
 fn semantic(n1: usize, e1: CompiledExpr, n2: usize, e2: CompiledExpr) -> CompiledExpr {
-    use Comb::*;
+    #[allow(non_snake_case)]
+    fn Sn(i: usize) -> Comb {
+        if i == 1 {
+            Comb::S
+        } else {
+            Comb::Sn(i)
+        }
+    }
+    #[allow(non_snake_case)]
+    fn Bn(i: usize) -> Comb {
+        if i == 1 {
+            Comb::B
+        } else {
+            Comb::Bn(i)
+        }
+    }
+    #[allow(non_snake_case)]
+    fn Cn(i: usize) -> Comb {
+        if i == 1 {
+            Comb::C
+        } else {
+            Comb::Cn(i)
+        }
+    }
     match (n1, n2) {
         (0, 0) => cap(e1, e2),
-        (0, n) => cap(cap(B(n), e1), e2),
-        (n, 0) => cap(cap(C(n), e1), e2),
+        (0, n) => cap(cap(Bn(n), e1), e2),
+        (n, 0) => cap(cap(Cn(n), e1), e2),
         (n, m) => {
             if n == m {
-                cap(cap(S(n), e1), e2)
+                cap(cap(Sn(n), e1), e2)
             } else if n < m {
-                cap(cap(B(m - n), cap(S(n), e1)), e2)
+                cap(cap(Bn(m - n), cap(Sn(n), e1)), e2)
             } else {
-                cap(cap(C(n - m), cap(cap(B(n - m), S(m)), e1)), e2)
+                cap(cap(Cn(n - m), cap(cap(Bn(n - m), Sn(m)), e1)), e2)
             }
         }
     }
@@ -83,7 +106,8 @@ fn semantic(n1: usize, e1: CompiledExpr, n2: usize, e2: CompiledExpr) -> Compile
 mod tests {
     use super::super::to_debruijn;
     use super::*;
-    use crate::{lex, parser::parse_expr};
+    use crate::lex;
+    use crate::parser::{parse_compiled_expr, parse_expr};
 
[…]

### 2022-12-21 experiments/superg/src/parser.rs

@@ -137,6 +137,9 @@ fn parse_compiled_expr_inner(tokens: &mut VecDeque<Token>) -> CompiledExpr {
             "Not" => Not,
             "Abort" => Abort,
             _ => {
+                if let Some(bulk_comb) = parse_bulk_comb(s.as_str()) {
+                    return CompiledExpr::Comb(bulk_comb);
+                }
                 // Assume a string of combinators
                 return s
                     .chars()
@@ -147,6 +150,16 @@ fn parse_compiled_expr_inner(tokens: &mut VecDeque<Token>) -> CompiledExpr {
         }),
         Some(Token::RParen) => panic!("Unexpected right paren"),
     }
+}
+
+fn parse_bulk_comb(s: &str) -> Option<Comb> {
+    let comb = match s.chars().nth(0) {
+        Some('S') => Comb::Sn,
+        Some('B') => Comb::Bn,
+        Some('C') => Comb::Cn,
+        _ => return None,
+    };
+    s[1..].parse().ok().map(comb)
 }
 
 fn parse_comb(c: char) -> Comb {

### 2022-12-21 experiments/superg/src/bracket.rs
Enable testing strict compiler.
Currently has failing tests.

@@ -1,5 +1,13 @@
 use crate::ast;
-use crate::compiled_expr::{cap, Comb, CompiledExpr};
+use crate::compiled_expr::{cap, Comb, CompiledExpr, ExprCompiler};
+
+pub struct BracketCompiler;
+
+impl ExprCompiler for BracketCompiler {
+    fn compile(&mut self, expr: &crate::ast::Expr) -> CompiledExpr {
+        compile(expr)
+    }
+}
 
 pub fn compile(e: &ast::Expr) -> CompiledExpr {
     match e {
@@ -9,24 +17,9 @@ pub fn compile(e: &ast::Expr) -> CompiledExpr {
             s => CompiledExpr::Var(s.to_owned()),
         },
         ast::Expr::BinOp(l, o, r) => {
-            let op_comb = match o {
-                ast::BinOp::Cons => Comb::P,
-                ast::BinOp::Plus => Comb::Plus,
-                ast::BinOp::Minus => Comb::Minus,
-                ast::BinOp::Times => Comb::Times,
-                ast::BinOp::Divide => Comb::Divide,
-                ast::BinOp::Eq => Comb::Eq,
-                ast::BinOp::Neq => Comb::Neq,
-                ast::BinOp::Gt => Comb::Gt,
-                ast::BinOp::Gte => Comb::Gte,
-                ast::BinOp::Lt => Comb::Lt,
-                ast::BinOp::Lte => Comb::Lte,
-                ast::BinOp::And => Comb::And,
-                ast::BinOp::Or => Comb::Or,
-            };
             let l = compile(l);
             let r = compile(r);
-            cap(cap(op_comb, l), r)
+            cap(cap(Comb::from(*o), l), r)
         }
         ast::Expr::Not(e) => cap(Comb::Not, compile(e)),
         ast::Expr::Ap(l, r) => {

### 2022-12-21 experiments/superg/src/compiled_expr.rs

@@ -1,3 +1,5 @@
+use crate::ast::BinOp;
+
 #[derive(Debug, Clone, Copy, PartialEq, Eq, Hash)]
 pub enum Comb {
     S,
@@ -27,6 +29,26 @@ pub enum Comb {
     Sn(usize),
     Bn(usize),
     Cn(usize),
+}
+
+impl From<BinOp> for Comb {
+    fn from(o: BinOp) -> Self {
+        match o {
+            BinOp::Cons => Comb::P,
+            BinOp::Plus => Comb::Plus,
+            BinOp::Minus => Comb::Minus,
+            BinOp::Times => Comb::Times,
+            BinOp::Divide => Comb::Divide,
+            BinOp::Eq => Comb::Eq,
+            BinOp::Neq => Comb::Neq,
+            BinOp::Gt => Comb::Gt,
+            BinOp::Gte => Comb::Gte,
+            BinOp::Lt => Comb::Lt,
+            BinOp::Lte => Comb::Lte,
+            BinOp::And => Comb::And,
+            BinOp::Or => Comb::Or,
+        }
+    }
 }
 
 #[derive(Clone, PartialEq)]
@@ -61,4 +83,8 @@ impl Into<CompiledExpr> for Comb {
 
 pub fn cap<A: Into<CompiledExpr>, B: Into<CompiledExpr>>(a: A, b: B) -> CompiledExpr {
     CompiledExpr::Ap(Box::new(a.into()), Box::new(b.into()))
+}
+
+pub trait ExprCompiler {
+    fn compile(&mut self, expr: &crate::ast::Expr) -> CompiledExpr;
 }

### 2022-12-21 experiments/superg/src/kiselyov/mod.rs

@@ -1,9 +1,42 @@
 use crate::ast::Expr;
+use crate::compiled_expr::{CompiledExpr, ExprCompiler};
 
 pub mod lazy;
 pub mod lazy_opt;
 pub mod linear;
 pub mod strict;
+
+pub struct StrictCompiler;
+
+impl ExprCompiler for StrictCompiler {
+    fn compile(&mut self, expr: &crate::ast::Expr) -> CompiledExpr {
+        strict::compile_strict(&to_debruijn(expr, &mut Vec::new()))
+    }
+}
+
+pub struct LazyCompiler;
+
+impl ExprCompiler for LazyCompiler {
+    fn compile(&mut self, expr: &crate::ast::Expr) -> CompiledExpr {
+        lazy::compile_lazy(&to_debruijn(expr, &mut Vec::new()))
+    }
+}
+
+pub struct LazyOptCompiler;
+
+impl ExprCompiler for LazyOptCompiler {
+    fn compile(&mut self, expr: &crate::ast::Expr) -> CompiledExpr {
+        lazy_opt::compile_lazy_opt(&to_debruijn(expr, &mut Vec::new()))
+    }
+}
+
+pub struct LinearCompiler;
+
+impl ExprCompiler for LinearCompiler {
+    fn compile(&mut self, expr: &crate::ast::Expr) -> CompiledExpr {
+        linear::compile_linear(&to_debruijn(expr, &mut Vec::new()))
+    }
+}
 
 #[derive(Debug, Clone, PartialEq)]
 pub enum Type {
@@ -30,7 +63,6 @@ pub fn to_debruijn(e: &Expr, vars: &mut Vec<String>) -> BExpr {
             if let Some(i) = vars.iter().position(|x| x == v) {
                 BExpr::Var(vars.len() - 1 - i)
             } else {
-                println!("Did not find {} in {:?}", v, vars);
                 BExpr::SVar(v.clone())
             }
         }

### 2022-12-21 experiments/superg/src/kiselyov/strict.rs

@@ -1,10 +1,9 @@
+use super::BExpr;
 /// The strict compilation algorithm, as given in fig.6
 use crate::compiled_expr::{cap, Comb, CompiledExpr};
 
-use super::BExpr;
-
 fn infer_n(e: &BExpr) -> usize {
-    let n = match e {
+    match e {
         BExpr::Var(i) => i + 1,
         BExpr::Lam(e) => {
             let n = infer_n(e);
@@ -23,13 +22,11 @@ fn infer_n(e: &BExpr) -> usize {
         BExpr::SVar(_) => 0,
         BExpr::BinOp(l, _, r) => std::cmp::max(infer_n(l), infer_n(r)),
         BExpr::Not(e) => infer_n(e),
-    };
-    println!("infer_n({:?}) = {}", e, n);
-    n
+    }
 }
 
 pub fn compile_strict(e: &BExpr) -> CompiledExpr {
-    let c = match e {
+    match e {
         BExpr::Var(i) => {
             let n = infer_n(e);
             if n == 1 {
@@ -57,17 +54,26 @@ pub fn compile_strict(e: &BExpr) -> CompiledExpr {
             compile_strict(e2),
         ),
         BExpr::Int(i) => CompiledExpr::Int(*i),
-        BExpr::SVar(s) => CompiledExpr::Var(s.clone()),
-        BExpr::BinOp(_, _, _) => todo!(),
+        BExpr::SVar(s) => match s.as_str() {
+            "if" => CompiledExpr::Comb(Comb::Cond),
+            _ => CompiledExpr::Var(s.clone()),
+        },
+        BExpr::BinOp(l, o, r) => {
+            let nl = infer_n(l);
+            // l <op> r => ((<op> l) r)
+            semantic(
+                nl,
+                semantic(0, CompiledExpr::Comb(Comb::from(*o)), nl, compile_strict(l)),
+                infer_n(r),
+                compile_strict(r),
+            )
+        }
         BExpr::Not(_) => todo!(),
-    };
-    println!("compile({:?}) = {:#?}", e, c);
-    c
+    }
 }
 
 fn semantic(n1: usize, e1: CompiledExpr, n2: usize, e2: CompiledExpr) -> CompiledExpr {
-    let dbg = format!("semantic({}, {:?}, {}, {:?}) = ", n1, e1, n2, e2);
-    let c = match (n1, n2) {
[…]

### 2022-12-21 experiments/superg/src/turner/mod.rs

@@ -1,8 +1,8 @@
-use crate::compiled_expr::{Comb, CompiledExpr};
+use crate::compiled_expr::{Comb, CompiledExpr, ExprCompiler};
 use std::io::Write;
 use std::{collections::HashMap, fs::File, io::BufWriter, path::PathBuf};
 
-use crate::{ast, bracket};
+use crate::ast;
 
 // TODO: Allow dynamically growing the heap.
 const CELLS: usize = 250_000;
@@ -52,6 +52,8 @@ impl Comb {
             Comb::S => TurnerEngine::run_s_comb,
             Comb::K => TurnerEngine::run_k_comb,
             Comb::I => TurnerEngine::run_i_comb,
+            Comb::B => TurnerEngine::run_b_comb,
+            Comb::C => TurnerEngine::run_c_comb,
             Comb::Plus => TurnerEngine::run_plus_comb,
             Comb::Minus => TurnerEngine::run_minus_comb,
             Comb::Times => TurnerEngine::run_times_comb,
@@ -127,7 +129,7 @@ struct CompiledDef {
 }
 
 impl TurnerEngine {
-    pub fn compile(program: &ast::Program) -> TurnerEngine {
+    pub fn compile<C: ExprCompiler>(compiler: &mut C, program: &ast::Program) -> TurnerEngine {
         let mut engine = TurnerEngine {
             tag: vec![Tag::wanted(); CELLS],
             hd: vec![CellPtr(0i32); CELLS],
@@ -146,7 +148,7 @@ impl TurnerEngine {
             .iter()
             .map(|def| CompiledDef {
                 name: def.name.clone(),
-                expr: bracket::compile(&def.as_lam()),
+                expr: compiler.compile(&def.as_lam()),
             })
             .collect();
 
@@ -258,6 +260,38 @@ impl TurnerEngine {
             } else {
                 StepResult::Cell(engine.tl(args[0]))
             }
+        })
+    }
+
+    fn run_b_comb(&mut self) -> Option<CellPtr> {
+        self.run_comb3(|engine, args| {
+            // B f g x => f (g x)
+            let x = engine.tl(args[2]);
+            let g = engine.tl(args[1]);
+            let f = engine.tl(args[0]);
+            // If x is an int, we should transfer that to the new location
+            let mut tag = Tag::wanted();
+            if engine.tag(args[2]).is_rhs_int() {
+                tag = tag.set_rhs_int();
+            }
+            let gx = engine.make_cell(tag, g, x);
[…]

### 2022-12-21 experiments/superg/tests/turner.rs

@@ -1,0 +1,219 @@
+use superg::bracket::BracketCompiler;
+use superg::compiled_expr::ExprCompiler;
+use superg::kiselyov::{LazyCompiler, LazyOptCompiler, LinearCompiler, StrictCompiler};
+use superg::lexer::lex;
+use superg::parser::parse;
+use superg::turner::TurnerEngine;
+
+#[test]
+fn test_id() {
+    assert_runs_to_int(
+        "test_id",
+        r#"
+(defun id (x) x)
+(defun main () (id 42))
+        "#,
+        42,
+        StepLimit(10),
+    );
+}
+
+#[test]
+fn test_k() {
+    assert_runs_to_int(
+        "test_k",
+        r#"
+(defun k (x y) x)
+(defun main () (k 42 84))
+        "#,
+        42,
+        StepLimit(20),
+    );
+}
+
+#[test]
+fn test_s() {
+    assert_runs_to_int(
+        "test_s",
+        r#"
+(defun s (f g x) (f x (g x)))
+(defun k (x y) x)
+(defun main () (s k k 42))
+        "#,
+        42,
+        StepLimit(200),
+    );
+}
+
+#[test]
+fn test_b() {
+    assert_runs_to_int(
+        "test_b",
+        r#"
+(defun b (f g x) (f (g x)))
+(defun k (x y) x)
+(defun i (x) x)
+(defun main () (b i i 42))
+    "#,
+        42,
+        StepLimit(200),
+    );
+}
+
+#[test]
+fn test_add() {
+    assert_runs_to_int(
+        "test_add",
+        r#"
+(defun main () (+ 2 40))
+        "#,
+        42,
+        StepLimit(10),
+    );
+}
+
+#[test]
+fn test_add_indirect() {
+    assert_runs_to_int(
+        "test_add_indirect",
+        r#"
+(defun id (x) x)
+(defun main () (+ (id 2) (id 40)))
+        "#,
+        42,
+        StepLimit(20),
+    );
+}
+
+#[test]
+fn test_cond() {
+    assert_runs_to_int(
+        "test_cond",
+        r#"
+(defun main () (if 0 1000 (if 1 42 2000)))
+        "#,
+        42,
+        StepLimit(10),
+    )
+}
+
+#[test]
+fn test_cond_add() {
+    assert_runs_to_int(
+        "test_cond_add1",
+        r#"
+(defun main () (+ 2 (if 0 30 40)))
+        "#,
+        42,
+        StepLimit(10),
+    );
+
[…]

### 2022-12-21 experiments/superg/src/turner/mod.rs
Get strict compiler working.

@@ -286,12 +286,17 @@ impl TurnerEngine {
             let g = engine.tl(args[1]);
             let f = engine.tl(args[0]);
             // If x is an int, we should transfer that to the new location
-            let mut tag = Tag::wanted();
+            let mut x_tag = Tag::wanted();
             if engine.tag(args[2]).is_rhs_int() {
-                tag = tag.set_rhs_int();
+                x_tag = x_tag.set_rhs_int();
             }
-            let fx = engine.make_cell(tag, f, x);
-            StepResult::CellContents(Tag::wanted(), fx, g)
+            // If g is an int, we should transfer that to the new location
+            let mut g_tag = Tag::wanted();
+            if engine.tag(args[1]).is_rhs_int() {
+                g_tag = g_tag.set_rhs_int();
+            }
+            let fx = engine.make_cell(x_tag, f, x);
+            StepResult::CellContents(g_tag, fx, g)
         })
     }
 

### 2022-12-21 experiments/superg/tests/turner.rs

@@ -146,7 +146,6 @@ fn test_cond_eq() {
 
 #[test]
 fn test_factorial() {
-    // TODO: was 5! = 120
     assert_runs_to_int(
         "test_factorial",
         r#"
@@ -154,9 +153,9 @@ fn test_factorial() {
       (if (= n 1)
           1
           (* n (fac (- n 1)))))
-    (defun main () (fac 2))
+    (defun main () (fac 5))
             "#,
-        2,
+        120,
         StepLimit(1000),
     );
 }

### 2022-12-21 experiments/superg/src/kiselyov/lazy.rs
Enable kiselyov's lazy algorithm.

@@ -26,7 +26,9 @@ fn infer_context(e: &BExpr) -> Context {
             ctx
         }
         BExpr::Ap(f, a) => unify_contexts(infer_context(f), infer_context(a)),
-        _ => todo!(),
+        BExpr::BinOp(l, _, r) => unify_contexts(infer_context(l), infer_context(r)),
+        BExpr::SVar(_) | BExpr::Int(_) => Context::new(),
+        BExpr::Not(e) => infer_context(e),
     }
 }
 
@@ -78,7 +80,27 @@ pub fn compile_lazy(e: &BExpr) -> CompiledExpr {
             infer_context(e2),
             compile_lazy(e2),
         ),
-        _ => todo!(),
+        BExpr::Int(i) => CompiledExpr::Int(*i),
+        BExpr::SVar(s) => match s.as_str() {
+            "if" => CompiledExpr::Comb(Comb::Cond),
+            _ => CompiledExpr::Var(s.clone()),
+        },
+        BExpr::BinOp(l, o, r) => {
+            let cl = infer_context(l);
+            // l <op> r => ((<op> l) r)
+            semantic(
+                cl.clone(),
+                semantic(
+                    Context::new(),
+                    CompiledExpr::Comb(Comb::from(*o)),
+                    cl,
+                    compile_lazy(l),
+                ),
+                infer_context(r),
+                compile_lazy(r),
+            )
+        }
+        BExpr::Not(_) => todo!(),
     }
 }
 

### 2022-12-21 experiments/superg/tests/turner.rs

@@ -197,6 +197,8 @@ fn assert_runs_to_int(_test_name: &str, program: &str, v: i32, l: StepLimit) {
     assert_runs_to_int_gen(BracketCompiler, _test_name, program, v, l);
     println!("Strict");
     assert_runs_to_int_gen(StrictCompiler, _test_name, program, v, l);
+    println!("Lazy");
+    assert_runs_to_int_gen(LazyCompiler, _test_name, program, v, l);
 }
 
 fn assert_runs_to_int_gen<C: ExprCompiler>(

### 2022-12-21 experiments/superg/src/kiselyov/lazy_opt.rs
Add lazy opt.

@@ -35,7 +35,19 @@ fn compile(e: &BExpr) -> CExpr {
             CExpr::Weak(e) => semantic(CExpr::Closed(CompiledExpr::Comb(Comb::K)), *e),
         },
         BExpr::Ap(a, b) => semantic(compile(a), compile(b)),
-        _ => todo!(),
+        BExpr::Int(i) => CExpr::Closed(CompiledExpr::Int(*i)),
+        BExpr::SVar(s) => CExpr::Closed(match s.as_str() {
+            "if" => CompiledExpr::Comb(Comb::Cond),
+            _ => CompiledExpr::Var(s.clone()),
+        }),
+        BExpr::BinOp(l, o, r) => semantic(
+            semantic(
+                CExpr::Closed(CompiledExpr::Comb(Comb::from(*o))),
+                compile(l),
+            ),
+            compile(r),
+        ),
+        BExpr::Not(_) => todo!(),
     }
 }
 

### 2022-12-21 experiments/superg/tests/turner.rs

@@ -199,6 +199,8 @@ fn assert_runs_to_int(_test_name: &str, program: &str, v: i32, l: StepLimit) {
     assert_runs_to_int_gen(StrictCompiler, _test_name, program, v, l);
     println!("Lazy");
     assert_runs_to_int_gen(LazyCompiler, _test_name, program, v, l);
+    println!("Lazy opt");
+    assert_runs_to_int_gen(LazyOptCompiler, _test_name, program, v, l);
 }
 
 fn assert_runs_to_int_gen<C: ExprCompiler>(

### 2022-12-21 experiments/superg/src/kiselyov/linear.rs
Generate a correct bulk combinator.

@@ -55,8 +55,20 @@ pub fn compile_linear(e: &BExpr) -> CompiledExpr {
             compile_linear(e2),
         ),
         BExpr::Int(i) => CompiledExpr::Int(*i),
-        BExpr::SVar(s) => CompiledExpr::Var(s.clone()),
-        BExpr::BinOp(_, _, _) => todo!(),
+        BExpr::SVar(s) => match s.as_str() {
+            "if" => CompiledExpr::Comb(Comb::Cond),
+            _ => CompiledExpr::Var(s.clone()),
+        },
+        BExpr::BinOp(l, o, r) => {
+            let nl = infer_n(l);
+            // l <op> r => ((<op> l) r)
+            semantic(
+                nl,
+                semantic(0, CompiledExpr::Comb(Comb::from(*o)), nl, compile_linear(l)),
+                infer_n(r),
+                compile_linear(r),
+            )
+        }
         BExpr::Not(_) => todo!(),
     }
 }

### 2022-12-21 experiments/superg/src/turner/mod.rs

@@ -61,6 +61,9 @@ impl Comb {
             Comb::Eq => TurnerEngine::run_eq_comb,
             Comb::Lt => TurnerEngine::run_lt_comb,
             Comb::Abort => TurnerEngine::run_abort_comb,
+            Comb::Sn(2) => TurnerEngine::run_s2_comb,
+            Comb::Bn(2) => TurnerEngine::run_b2_comb,
+            Comb::Cn(2) => TurnerEngine::run_c2_comb,
             _ => todo!("No implementation for combinator {:?}", self),
         }
     }
@@ -666,6 +669,166 @@ impl TurnerEngine {
     run_comb!(run_comb1, 1);
     run_comb!(run_comb2, 2);
     run_comb!(run_comb3, 3);
+
+    // Runtime dynamic version of run_combn
+    fn run_combn<F: FnOnce(&mut TurnerEngine, &[CellPtr]) -> StepResult>(
+        &mut self,
+        arg_count: usize,
+        handler: F,
+    ) -> Option<CellPtr> {
+        let frame_size = arg_count + 1;
+        debug_assert!(self.stack.len() >= frame_size);
+        // Arguments are in reverse order because the stack is LIFO.
+        let mut args = vec![CellPtr(0); arg_count];
+        for i in 0..arg_count {
+            args[i] = *unsafe { self.stack.get_unchecked(self.stack.len() - 2 - i) };
+        }
+        let frame_start = args[arg_count - 1];
+
+        match handler(self, &args) {
+            StepResult::Value(v) => {
+                // Make an indirection node
+                self.set_tag(frame_start, Tag::wanted().set_rhs_int());
+                self.set_hd(frame_start, Comb::I);
+                self.set_tl(frame_start, CellPtr(v));
+
+                // If nothing is on the stack, return the node
+                if self.stack.len() == frame_size {
+                    return Some(frame_start);
+                }
+                // Remove this frame, returning to the parent expression on the stack
+                unsafe { self.stack.set_len(self.stack.len() - frame_size) };
+                None
+            }
+            StepResult::Cell(c) => {
+                // Make an indirection node
+                self.set_tag(frame_start, Tag::wanted());
+                self.set_hd(frame_start, Comb::I);
+                self.set_tl(frame_start, c);
+
+                // Make the new cell top of the stack
+                let new_len = self.stack.len() - arg_count;
+                unsafe {
+                    *self.stack.get_unchecked_mut(new_len - 1) = c;
[…]

### 2022-12-21 experiments/superg/tests/turner.rs

@@ -201,6 +201,8 @@ fn assert_runs_to_int(_test_name: &str, program: &str, v: i32, l: StepLimit) {
     assert_runs_to_int_gen(LazyCompiler, _test_name, program, v, l);
     println!("Lazy opt");
     assert_runs_to_int_gen(LazyOptCompiler, _test_name, program, v, l);
+    println!("Linear");
+    assert_runs_to_int_gen(LinearCompiler, _test_name, program, v, l);
 }
 
 fn assert_runs_to_int_gen<C: ExprCompiler>(

### 2022-12-21 experiments/superg/src/turner/mod.rs
A dynamic implementation.

@@ -47,7 +47,7 @@ impl IntoCellPtr for Comb {
 }
 
 impl Comb {
-    fn implementation(&self) -> fn(&mut TurnerEngine) -> Option<CellPtr> {
+    fn implementation(&self) -> fn(&mut TurnerEngine, n: usize) -> Option<CellPtr> {
         match self {
             Comb::S => TurnerEngine::run_s_comb,
             Comb::K => TurnerEngine::run_k_comb,
@@ -61,9 +61,9 @@ impl Comb {
             Comb::Eq => TurnerEngine::run_eq_comb,
             Comb::Lt => TurnerEngine::run_lt_comb,
             Comb::Abort => TurnerEngine::run_abort_comb,
-            Comb::Sn(2) => TurnerEngine::run_s2_comb,
-            Comb::Bn(2) => TurnerEngine::run_b2_comb,
-            Comb::Cn(2) => TurnerEngine::run_c2_comb,
+            Comb::Sn(_) => TurnerEngine::run_sn_comb,
+            Comb::Bn(_) => TurnerEngine::run_bn_comb,
+            Comb::Cn(_) => TurnerEngine::run_cn_comb,
             _ => todo!("No implementation for combinator {:?}", self),
         }
     }
@@ -121,7 +121,7 @@ pub struct TurnerEngine {
     stack: Vec<CellPtr>,
 
     // Combinator implementation lookup table
-    comb_impl: Vec<fn(&mut TurnerEngine) -> Option<CellPtr>>,
+    comb_impl: Vec<(fn(&mut TurnerEngine, usize) -> Option<CellPtr>, usize)>,
     // Maps combinator to the index in comb_impl
     comb_map: HashMap<Comb, usize>,
 }
@@ -198,7 +198,13 @@ impl TurnerEngine {
             return;
         }
         // Add the combinator implementation
-        self.comb_impl.push(c.implementation());
+        let n = match c {
+            Comb::Sn(n) => n,
+            Comb::Bn(n) => n,
+            Comb::Cn(n) => n,
+            _ => 0,
+        };
+        self.comb_impl.push((c.implementation(), n));
         let idx = self.comb_impl.len() - 1;
         self.comb_map.insert(c, idx);
         self.next_cell = CellPtr(idx as i32);
@@ -217,8 +223,8 @@ impl TurnerEngine {
 
     fn step(&mut self) -> Option<CellPtr> {
         let top = self.stack.last().unwrap();
[…]

### 2023-01-01 experiments/superg/doc/paper/paper.tex
JIT call instructions.

@@
 \label{sec:tigre}
+\subsection*{Eliminating tags}
+Tags are commonly used to distinguish pointers, combinators and literal values.
+We get rid of the literal value tag by making all literals go through an indirection node, and introducing a special \texttt{LIT} combinator for such indirection nodes.
+To determine if a right-hand side is a literal, we can check for the \texttt{LIT} combinator rather than a tag.
+In the TIGRE paper, there is also a pointer and combinator tag, but with Miranda's design, this is not necessary.
+If we insist, that final tag value can be put as the lowest bit in the pointer value, since a pointer is typically aligned to some boundary.
+We are working with 32-bit (4-byte) wide values, for which on X86 we can assume 4-byte alignment, giving us 2 bits to play with.
+
+\subsection*{Threaded code interpreter}
+We make the nodes themselves executable code.
+For nodes that represent the combinators, a pointer to that node is just a pointer to the subroutine that implements the combinator.
+For application nodes, they are a jump instruction together with the left node as the target (so the left node is the width of one instruction).
+The jump instruction should push the current PC onto the stack, which means the stack now directly represents the spine.
+
+The TIGRE style requires that we make our memory writable and executable, which is problematic in many systems.
+It seems to work okay on Linux though, so we can use it.
+
+The instruction to use on X86 is \texttt{call}.
+It takes one parameter, which is a PC-relative offset.
+When we create a node, we should subtract the target position from the location where we are writing the instruction, and use that as the jump parameter.
+
+The trampolines can be generated or hand-coded ahead of time, so our runtime only needs to assemble \texttt{CALL} instructions.
+
+For combinators, we can generate a trampoline that grabs the arguments from the stack, puts them in registers and calls the rust implementation.
+
+It seems to be a good idea to align nodes at a 16-byte boundary for better jump performance:
+https://groups.google.com/g/golang-nuts/c/dhGbgC1pAmA
+do not put more than 2 branches in a 16-byte block.
+
+Encoding of the call instruction is \texttt{E8} \textit{rel32}, where \textit{rel32} is a 32-bit (little-endian) offset from the \emph{next} instruction.
 \textbf{TODO}

### 2023-01-01 experiments/superg/src/lib.rs

@@ -9,3 +9,5 @@ pub mod kiselyov;
 
 pub use lexer::lex;
 pub use parser::parse;
+
+pub mod tigre;

### 2023-01-01 experiments/superg/src/tigre/jit.rs

@@ -1,0 +1,211 @@
+pub fn assemble_call(off: i32) -> [u8; 5] {
+    let mut buf = [0u8; 5];
+    // CALL rel32
+    buf[0] = 0xE8;
+    buf[1..5].copy_from_slice(&off.to_le_bytes());
+    buf
+}
+
+pub struct JitMem {
+    buf: *mut u8,
+    len: usize,
+}
+
+impl JitMem {
+    pub fn new(len: usize) -> JitMem {
+        // Allow RWX, everything!
+        let prot = libc::PROT_READ | libc::PROT_WRITE | libc::PROT_EXEC;
+        let flags = libc::MAP_ANONYMOUS | libc::MAP_PRIVATE;
+        // Unused because the mmap is anonymous
+        let fd = -1;
+        let offset = 0;
+        let buf =
+            unsafe { libc::mmap(std::ptr::null_mut(), len, prot, flags, fd, offset) as *mut u8 };
+        if buf == libc::MAP_FAILED as *mut u8 {
+            panic!("mmap failed: {}", std::io::Error::last_os_error());
+        }
+
+        JitMem { buf, len }
+    }
+
+    pub fn slice_mut(&mut self) -> &mut [u8] {
+        unsafe { std::slice::from_raw_parts_mut(self.buf, self.len) }
+    }
+}
+
+impl core::ops::Drop for JitMem {
+    fn drop(&mut self) {
+        let res = unsafe { libc::munmap(self.buf as *mut libc::c_void, self.len) };
+        if res != 0 {
+            panic!(
+                "munmap({:?}, {}) failed: {}",
+                self.buf,
+                self.len,
+                std::io::Error::last_os_error()
+            );
+        }
+    }
+}
+
+#[cfg(test)]
+mod tests {
+    use super::*;
+
+    #[test]
+    fn write_exec() {
+        let code = &[
+            0x55, //    push   %rbp
+            0x48, 0x89, 0xe5, //    mov    %rsp,%rbp
+            0xb8, 0x37, 0x00, 0x00, 0x00, //    mov    $0x37,%eax
+            0xc9, //    leaveq
+            0xc3, //    retq
+        ];
+
+        unsafe {
+            let ptr = libc::mmap(
+                std::ptr::null_mut(),
+                code.len(),
+                libc::PROT_READ | libc::PROT_WRITE | libc::PROT_EXEC,
+                libc::MAP_ANONYMOUS | libc::MAP_PRIVATE,
+                -1,
[…]

### 2023-01-05 experiments/superg/doc/paper/paper.tex
Add a JIT placer.

@@
 
+The \texttt{call} instruction pushes the \texttt{EIP} register onto the stack, which contains the address of the 'instruction' after the \texttt{call}.
+In our calling convention, this means the stack consists of pointers to the right-hand sides, very convenient.
+
+A \texttt{push} instruction first decrements the stack pointer \texttt{rsp}, then writes the new value to the address of  \texttt{rsp}.
+Assembly to load the value of the top of the stack is therefore: \texttt{mov rax, [rsp]}.
+To load the second value: \texttt{mov rax, [rsp+8]}.
+
 It seems to be a good idea to align nodes at a 16-byte boundary for better jump performance:
@@
 
+Suggestion from \textit{Common Benchmarking Pitfalls}: Always look at your distribution.
+When you have run benchmarks, take a look at the times for each sample.
+
 \section{Conclusion}

### 2023-01-05 experiments/superg/src/tigre/comb.rs

@@ -1,0 +1,39 @@
+use std::arch::global_asm;
+
+global_asm!(
+    r#"
+    .global comb_LIT
+    comb_LIT:
+        pop rax
+        ret
+    .global comb_LIT_end
+    comb_LIT_end:
+"#
+);
+extern "C" {
+    pub fn comb_LIT() -> usize;
+}
+
+#[cfg(test)]
+mod tests {
+    use super::*;
+
+    #[test]
+    fn lit_dispatch() {
+        let res = unsafe { lit_dispatch_helper() };
+        assert_eq!(res, 42);
+    }
+
+    global_asm!(
+        r#"
+        .global lit_dispatch_helper
+        lit_dispatch_helper:
+            // Emulates a cell with lhs LIT and rhs 42
+            push 42
+            jmp comb_LIT
+    "#
+    );
+    extern "C" {
+        fn lit_dispatch_helper() -> u64;
+    }
+}

### 2023-01-05 experiments/superg/src/tigre/jit.rs

@@ -6,30 +6,73 @@ pub fn assemble_call(off: i32) -> [u8; 5] {
     buf
 }
 
+const PAGE_SIZE: usize = 4096;
+
 pub struct JitMem {
     buf: *mut u8,
     len: usize,
 }
 
 impl JitMem {
-    pub fn new(len: usize) -> JitMem {
+    pub fn new_at(ptr: *mut u8, len: usize) -> std::io::Result<JitMem> {
         // Allow RWX, everything!
         let prot = libc::PROT_READ | libc::PROT_WRITE | libc::PROT_EXEC;
-        let flags = libc::MAP_ANONYMOUS | libc::MAP_PRIVATE;
+        let flags = libc::MAP_ANONYMOUS | libc::MAP_PRIVATE | libc::MAP_FIXED_NOREPLACE;
         // Unused because the mmap is anonymous
         let fd = -1;
         let offset = 0;
-        let buf =
-            unsafe { libc::mmap(std::ptr::null_mut(), len, prot, flags, fd, offset) as *mut u8 };
+        let buf = unsafe {
+            libc::mmap(ptr as *mut libc::c_void, len, prot, flags, fd, offset) as *mut u8
+        };
         if buf == libc::MAP_FAILED as *mut u8 {
-            panic!("mmap failed: {}", std::io::Error::last_os_error());
+            Err(std::io::Error::last_os_error())
+        } else {
+            Ok(JitMem { buf, len })
         }
+    }
 
-        JitMem { buf, len }
+    pub fn new_in_range(min_ptr: *mut u8, max_ptr: *mut u8, len: usize) -> std::io::Result<JitMem> {
+        // Check inputs
+        assert_eq!(
+            min_ptr as usize % PAGE_SIZE,
+            0,
+            "min_ptr must be aligned to {}",
+            PAGE_SIZE
+        );
+        assert_eq!(
+            max_ptr as usize % PAGE_SIZE,
+            0,
+            "max_ptr must be aligned to {}",
+            PAGE_SIZE
+        );
+        assert!(min_ptr < max_ptr, "min_ptr must be less than max_ptr");
+
+        let mut ptr = min_ptr;
+        loop {
+            match JitMem::new_at(ptr, len) {
+                Ok(jit) => return Ok(jit),
+                Err(e) => {
+                    // Try next page
+                    ptr = ptr.wrapping_add(PAGE_SIZE);
+                    if ptr >= max_ptr {
+                        // Out of pages to try
[…]

### 2023-01-05 experiments/superg/src/tigre/mod.rs

@@ -1,1 +1,4 @@
+use std::arch::global_asm;
+
+pub mod comb;
 pub mod jit;

### 2023-01-05 experiments/superg/src/tigre/comb.rs
Dispatch LIT.

@@ -5,13 +5,36 @@ global_asm!(
     .global comb_LIT
     comb_LIT:
         pop rax
+        mov rax, [rax]
         ret
-    .global comb_LIT_end
-    comb_LIT_end:
 "#
 );
 extern "C" {
     pub fn comb_LIT() -> usize;
+}
+
+global_asm!(
+    r#"
+    .global comb_I
+    comb_I:
+        mov rcx, [rsp]
+        add rsp, 8
+        jmp [rcx]
+"#
+);
+extern "C" {
+    pub fn comb_I() -> usize;
+}
+
+global_asm!(
+    r#"
+    .global comb_Abort
+    comb_Abort:
+        jmp comb_Abort
+"#
+);
+extern "C" {
+    pub fn comb_Abort() -> usize;
 }
 
 #[cfg(test)]
@@ -28,9 +51,9 @@ mod tests {
         r#"
         .global lit_dispatch_helper
         lit_dispatch_helper:
-            // Emulates a cell with lhs LIT and rhs 42
-            push 42
-            jmp comb_LIT
+        call comb_LIT
+        .long 42
+        .long 0
     "#
     );
     extern "C" {

### 2023-01-05 experiments/superg/src/tigre/mod.rs

@@ -1,4 +1,190 @@
-use std::arch::global_asm;
-
 pub mod comb;
 pub mod jit;
+
+use std::collections::HashMap;
+
+use jit::JitMem;
+
+use crate::{
+    ast,
+    compiled_expr::{Comb, CompiledExpr, ExprCompiler},
+};
+
+use self::jit::JitPlacer;
+
+// Cells are 16 bytes for optimal alignment.
+// See https://groups.google.com/g/golang-nuts/c/dhGbgC1pAmA/m/Rcqwcd5mGmoJ.
+#[repr(packed)]
+pub struct Cell {
+    // CALL rel32 (0xE8)
+    call_opcode: u8,
+    call_rel_addr: i32,
+    // right-hand side.
+    // Can be an integer or a pointer
+    arg: i64,
+    used: bool,
+    _padding: [u8; 2],
+}
+
+const CALL_OPCODE: u8 = 0xE8;
+// OPCODE  + rel32
+const CALL_LEN: isize = 5;
+
+impl Cell {
+    fn new(lhs: i32, rhs: i32) -> Cell {
+        Cell {
+            call_opcode: CALL_OPCODE,
+            call_rel_addr: lhs,
+            arg: rhs as i64,
+            used: true,
+            _padding: [0u8; 2],
+        }
+    }
+}
+
+#[derive(Debug, Copy, Clone)]
+struct CellPtr(*mut Cell);
+
+trait IntoCellPtr {
+    fn into_cell_ptr(&self, engine: &TigreEngine) -> CellPtr;
+}
+
+impl IntoCellPtr for Comb {
+    fn into_cell_ptr(&self, engine: &TigreEngine) -> CellPtr {
+        match self {
+            Comb::I => CellPtr(comb::comb_I as *mut Cell),
+            Comb::Abort => CellPtr(comb::comb_Abort as *mut Cell),
+            _ => unimplemented!("into_cell_ptr {:?}", self),
+        }
+    }
+}
+
+struct Lit;
+impl IntoCellPtr for Lit {
+    fn into_cell_ptr(&self, engine: &TigreEngine) -> CellPtr {
+        CellPtr(comb::comb_LIT as *mut Cell)
+    }
+}
+
+impl IntoCellPtr for i32 {
+    fn into_cell_ptr(&self, engine: &TigreEngine) -> CellPtr {
+        CellPtr(*self as *mut Cell)
+    }
+}
+
[…]

### 2023-01-08 experiments/superg/src/tigre/comb.rs
Implement S and K.

@@ -1,5 +1,8 @@
 use std::arch::global_asm;
 
+use super::{Cell, CellPtr, TigreEngine, CALL_LEN, CALL_OPCODE};
+
+// LIT combinator
 global_asm!(
     r#"
     .global comb_LIT
@@ -13,28 +16,71 @@ extern "C" {
     pub fn comb_LIT() -> usize;
 }
 
+// I combinator
 global_asm!(
     r#"
     .global comb_I
     comb_I:
-        mov rcx, [rsp]
-        add rsp, 8
-        jmp [rcx]
+        pop rax
+        jmp [rax]
 "#
 );
 extern "C" {
     pub fn comb_I() -> usize;
 }
 
+// K combinator
 global_asm!(
     r#"
-    .global comb_Abort
-    comb_Abort:
-        jmp comb_Abort
+    .global comb_K
+    comb_K:
+        mov rax, [rsp]
+        add rsp, 16
+        jmp [rax]
 "#
 );
 extern "C" {
-    pub fn comb_Abort() -> usize;
+    pub fn comb_K() -> usize;
+}
+
+// S combinator
+global_asm!(
+    r#"
+    .global comb_S
+    comb_S:
+        // Load f, g, x pointers as arguments to make_s 
+        mov rdi, [rsp]
+        mov rsi, [rsp+8]
+        mov rdx, [rsp+16]
+        call make_s
+        // Pop arguments and jump to the updated node 
+        add rsp, 24
+        jmp rax
+    "#
+);
+extern "C" {
+    pub fn comb_S() -> usize;
+}
+#[no_mangle]
+unsafe extern "C" fn make_s(f: *const CellPtr, g: *const CellPtr, x: *const CellPtr) -> CellPtr {
+    TigreEngine::with_current(|engine| {
+        // Make new cells for (f x), (g x)
+        let fx = engine.make_cell(*f, *x);
+        let gx = engine.make_cell(*g, *x);
+        // Pointers on the stack are to the right pointer of a cell, after the call instruction.
+        // To get a pointer to the full cell, we subtract the length of a call instruction.
+        let top_cell_ptr = CellPtr(x.offset(-CALL_LEN) as *mut Cell);
+        let top_cell = engine.cell_mut(top_cell_ptr);
+        top_cell.call_opcode = CALL_OPCODE;
+        top_cell.set_call_addr(fx.0 as usize);
+        top_cell.arg = gx.0 as i64;
[…]

### 2023-01-08 experiments/superg/src/tigre/mod.rs

@@ -26,34 +26,39 @@ pub struct Cell {
     _padding: [u8; 2],
 }
 
+impl Cell {
+    pub fn set_call_addr(&mut self, target_addr: usize) {
+        let base_addr = self as *const Cell as isize + CALL_LEN;
+        let call_rel_addr = target_addr as isize - base_addr;
+        self.call_rel_addr = call_rel_addr.try_into().expect("have to jump too far");
+    }
+}
+
 const CALL_OPCODE: u8 = 0xE8;
 // OPCODE  + rel32
 const CALL_LEN: isize = 5;
 
-impl Cell {
-    fn new(lhs: i32, rhs: i32) -> Cell {
-        Cell {
-            call_opcode: CALL_OPCODE,
-            call_rel_addr: lhs,
-            arg: rhs as i64,
-            used: true,
-            _padding: [0u8; 2],
-        }
-    }
-}
-
 #[derive(Debug, Copy, Clone)]
+#[repr(transparent)]
 struct CellPtr(*mut Cell);
 
 trait IntoCellPtr {
     fn into_cell_ptr(&self, engine: &TigreEngine) -> CellPtr;
 }
 
+impl IntoCellPtr for CellPtr {
+    fn into_cell_ptr(&self, _: &TigreEngine) -> CellPtr {
+        *self
+    }
+}
+
 impl IntoCellPtr for Comb {
-    fn into_cell_ptr(&self, engine: &TigreEngine) -> CellPtr {
+    fn into_cell_ptr(&self, _: &TigreEngine) -> CellPtr {
         match self {
             Comb::I => CellPtr(comb::comb_I as *mut Cell),
             Comb::Abort => CellPtr(comb::comb_Abort as *mut Cell),
+            Comb::S => CellPtr(comb::comb_S as *mut Cell),
+            Comb::K => CellPtr(comb::comb_K as *mut Cell),
             _ => unimplemented!("into_cell_ptr {:?}", self),
         }
     }
@@ -61,13 +66,13 @@ impl IntoCellPtr for Comb {
 
 struct Lit;
 impl IntoCellPtr for Lit {
-    fn into_cell_ptr(&self, engine: &TigreEngine) -> CellPtr {
+    fn into_cell_ptr(&self, _: &TigreEngine) -> CellPtr {
         CellPtr(comb::comb_LIT as *mut Cell)
     }
 }
 
 impl IntoCellPtr for i32 {
[…]

### 2023-01-08 experiments/superg/src/tigre/comb.rs
Pass test_s.

@@ -69,9 +69,9 @@ unsafe extern "C" fn make_s(f: *const CellPtr, g: *const CellPtr, x: *const Cell
         let gx = engine.make_cell(*g, *x);
         // Pointers on the stack are to the right pointer of a cell, after the call instruction.
         // To get a pointer to the full cell, we subtract the length of a call instruction.
-        let top_cell_ptr = CellPtr(x.offset(-CALL_LEN) as *mut Cell);
+        let top_cell_ptr = CellPtr((x as *mut u8).offset(-CALL_LEN) as *mut Cell);
         let top_cell = engine.cell_mut(top_cell_ptr);
-        top_cell.call_opcode = CALL_OPCODE;
+        debug_assert_eq!(top_cell.call_opcode, CALL_OPCODE);
         top_cell.set_call_addr(fx.0 as usize);
         top_cell.arg = gx.0 as i64;
         top_cell_ptr

### 2023-01-08 experiments/superg/src/tigre/mod.rs

@@ -113,21 +113,16 @@ impl TigreEngine {
     }
 
     pub fn run(&mut self) -> i64 {
-        /*
-        ENGINE.with(|e| {
-            e.set(self as *mut TigreEngine);
-        });
-        */
-        unsafe { ENGINE = self as *mut TigreEngine };
         let main_ptr = self.def_lookup.get("main").expect("No main function");
         let func: fn() -> i64 = unsafe { std::mem::transmute(main_ptr.0) };
+
+        // Configure the global engine reference
+        unsafe { ENGINE = self as *mut TigreEngine };
+
         let res = func();
-        /*
-        ENGINE.with(|e| {
-            e.set(std::ptr::null_mut());
-        });
-        */
-        //unsafe { ENGINE = std::ptr::null_mut() };
+
+        // Deregister global engine reference
+        unsafe { ENGINE = std::ptr::null_mut() };
         res
     }
 
@@ -194,21 +189,9 @@ impl TigreEngine {
     pub fn with_current<R, F: FnOnce(&mut TigreEngine) -> R>(f: F) -> R {
         let engine = unsafe { &mut *ENGINE };
         f(engine)
-        /*
-        ENGINE.with(|e| {
-            let e = e.get();
-            assert!(
-                !e.is_null(),
-                "with_current must be called within a call to run"
-            );
-            let engine = unsafe { &mut *e };
-            f(engine)
-        })
-        */
     }
 }
 
-//thread_local!(static ENGINE: std::cell::Cell<*mut TigreEngine> = std::cell::Cell::new(std::ptr::null_mut()));
 static mut ENGINE: *mut TigreEngine = std::ptr::null_mut();
 
 #[cfg(test)]
@@ -244,6 +227,18 @@ mod tests {
             r#"
 (defun k (x y) x)
 (defun main () (k 42 84))
+        "#,
+            42,
+        );
+    }
+
+    #[test]
+    fn test_s() {
+        assert_runs_to_int(
+            r#"
+(defun s (f g x) (f x (g x)))
+(defun k (x y) x)
+(defun main () (s k k 42))
         "#,
             42,
         );

### 2023-01-08 experiments/superg/src/tigre/comb.rs
Implement plus combinator.

@@ -16,6 +16,12 @@ extern "C" {
     pub fn comb_LIT() -> usize;
 }
 
+// Abort combinator
+#[allow(non_snake_case)]
+pub extern "C" fn comb_Abort() -> usize {
+    panic!("Abort called")
+}
+
 // I combinator
 global_asm!(
     r#"
@@ -78,9 +84,41 @@ unsafe extern "C" fn make_s(f: *const CellPtr, g: *const CellPtr, x: *const Cell
     })
 }
 
-#[allow(non_snake_case)]
-pub extern "C" fn comb_Abort() -> usize {
-    panic!("Abort called")
+// Plus combinator
+global_asm!(
+    r#"
+    .global comb_plus
+    comb_plus:
+        // Load argument pointers as arguments
+        mov rdi, [rsp]
+        mov rsi, [rsp+8]
+        call apply_plus
+        // Pop arguments
+        add rsp, 16
+        // Return the computed value
+        ret
+"#
+);
+extern "C" {
+    pub fn comb_plus() -> usize;
+}
+// A pointer to a function that evaluates an argument to a strict operator.
+type ArgFn = unsafe extern "C" fn() -> i64;
+#[no_mangle]
+unsafe extern "C" fn apply_plus(a0: *const ArgFn, a1: *const ArgFn) -> i64 {
+    let res = (*a0)() + (*a1)();
+
+    TigreEngine::with_current(|engine| {
+        // Update the top cell with the new number.
+        // See `make_s` for details.
+        let top_cell_ptr = CellPtr((a1 as *mut u8).offset(-CALL_LEN) as *mut Cell);
+        let top_cell = engine.cell_mut(top_cell_ptr);
+        debug_assert_eq!(top_cell.call_opcode, CALL_OPCODE);
+        top_cell.set_call_addr(comb_LIT as usize);
+        top_cell.arg = res;
+    });
+
+    res
 }
 
 #[cfg(test)]

### 2023-01-08 experiments/superg/src/tigre/mod.rs

@@ -59,6 +59,7 @@ impl IntoCellPtr for Comb {
             Comb::Abort => CellPtr(comb::comb_Abort as *mut Cell),
             Comb::S => CellPtr(comb::comb_S as *mut Cell),
             Comb::K => CellPtr(comb::comb_K as *mut Cell),
+            Comb::Plus => CellPtr(comb::comb_plus as *mut Cell),
             _ => unimplemented!("into_cell_ptr {:?}", self),
         }
     }
@@ -239,6 +240,38 @@ mod tests {
 (defun s (f g x) (f x (g x)))
 (defun k (x y) x)
 (defun main () (s k k 42))
+        "#,
+            42,
+        );
+    }
+
+    #[test]
+    fn test_add() {
+        assert_runs_to_int(
+            r#"
+(defun main () (+ 2 40))
+        "#,
+            42,
+        );
+    }
+
+    #[test]
+    fn test_add_indirect() {
+        assert_runs_to_int(
+            r#"
+(defun id (x) x)
+(defun main () (+ (id 2) (id 40)))
+        "#,
+            42,
+        );
+    }
+
+    #[test]
+    fn add_reuse() {
+        assert_runs_to_int(
+            r#"
+(defun f (x) (+ x x))
+(defun main() (f (+ 20 1)))
         "#,
             42,
         );

### 2023-01-09 experiments/superg/src/tigre/comb.rs
Implement cond combinator.

@@ -121,6 +121,51 @@ unsafe extern "C" fn apply_plus(a0: *const ArgFn, a1: *const ArgFn) -> i64 {
     res
 }
 
+// Cond combinator
+global_asm!(
+    r#"
+    .global comb_cond
+    comb_cond:
+        // Load c, t, f pointers as arguments to apply_cond
+        mov rdi, [rsp]
+        mov rsi, [rsp+8]
+        mov rdx, [rsp+16]
+        call apply_cond
+        // Pop arguments and jump to the updated node 
+        add rsp, 24
+        jmp rax
+"#
+);
+extern "C" {
+    pub fn comb_cond() -> usize;
+}
+#[no_mangle]
+unsafe extern "C" fn apply_cond(c: *const ArgFn, t: *const CellPtr, f: *const CellPtr) -> CellPtr {
+    // TODO: avoid loading pointers to both branches in the assembly
+    let c_res = (*c)();
+    let branch_ptr = match c_res {
+        0 => f,
+        1 => t,
+        v => {
+            // We expect never to reach this code.
+            // In debug mode we panic, in release the behaviour is undefined.
+            debug_assert!(v == 0 || v == 1, "condition variable should be 0 or 1");
+            unsafe { std::hint::unreachable_unchecked() }
+        }
+    };
+
+    TigreEngine::with_current(|engine| {
+        // Update the top cell with an indirection to the taken branch.
+        // See `make_s` for details.
+        let top_cell_ptr = CellPtr((f as *mut u8).offset(-CALL_LEN) as *mut Cell);
+        let top_cell = engine.cell_mut(top_cell_ptr);
+        debug_assert_eq!(top_cell.call_opcode, CALL_OPCODE);
+        top_cell.set_call_addr(comb_I as usize);
+        top_cell.arg = (*branch_ptr).0 as i64;
+    });
+    *branch_ptr
+}
+
 #[cfg(test)]
 mod tests {
     use super::*;

### 2023-01-09 experiments/superg/src/tigre/mod.rs

@@ -60,6 +60,7 @@ impl IntoCellPtr for Comb {
             Comb::S => CellPtr(comb::comb_S as *mut Cell),
             Comb::K => CellPtr(comb::comb_K as *mut Cell),
             Comb::Plus => CellPtr(comb::comb_plus as *mut Cell),
+            Comb::Cond => CellPtr(comb::comb_cond as *mut Cell),
             _ => unimplemented!("into_cell_ptr {:?}", self),
         }
     }
@@ -272,6 +273,33 @@ mod tests {
             r#"
 (defun f (x) (+ x x))
 (defun main() (f (+ 20 1)))
+        "#,
+            42,
+        );
+    }
+
+    #[test]
+    fn test_cond() {
+        assert_runs_to_int(
+            r#"
+(defun main () (if 0 1000 (if 1 42 2000)))
+        "#,
+            42,
+        )
+    }
+
+    #[test]
+    fn test_cond_add() {
+        assert_runs_to_int(
+            r#"
+(defun main () (+ 2 (if 0 30 40)))
+        "#,
+            42,
+        );
+
+        assert_runs_to_int(
+            r#"
+(defun main () (if 0 30 (+ 40 2)))
         "#,
             42,
         );

### 2023-01-09 experiments/superg/src/tigre/comb.rs
Macro for generating binops.

@@ -84,42 +84,52 @@ unsafe extern "C" fn make_s(f: *const CellPtr, g: *const CellPtr, x: *const Cell
     })
 }
 
-// Plus combinator
-global_asm!(
-    r#"
-    .global comb_plus
-    comb_plus:
-        // Load argument pointers as arguments
-        mov rdi, [rsp]
-        mov rsi, [rsp+8]
-        call apply_plus
-        // Pop arguments
-        add rsp, 16
-        // Return the computed value
-        ret
-"#
-);
-extern "C" {
-    pub fn comb_plus() -> usize;
-}
 // A pointer to a function that evaluates an argument to a strict operator.
 type ArgFn = unsafe extern "C" fn() -> i64;
-#[no_mangle]
-unsafe extern "C" fn apply_plus(a0: *const ArgFn, a1: *const ArgFn) -> i64 {
-    let res = (*a0)() + (*a1)();
+macro_rules! comb_bin_op {
+    ($name:ident, $name_str:literal, $helper_func:ident, $helper_func_str:literal, $op:expr) => {
+        global_asm!(
+            concat!(".global ", $name_str),
+            concat!($name_str, ":"),
+            r#"
+                // Load argument pointers as arguments
+                mov rdi, [rsp]
+                mov rsi, [rsp+8]
+            "#,
+            concat!("call ", $helper_func_str),
+            r#"
+                // Pop arguments
+                add rsp, 16
+                // Return the computed value
+                ret
+            "#,
+        );
+        extern "C" {
+            pub fn $name() -> usize;
+        }
+        #[no_mangle]
+        unsafe extern "C" fn $helper_func(a0: *const ArgFn, a1: *const ArgFn) -> i64 {
+            let op: fn(i64, i64) -> i64 = $op;
+            let res = op((*a0)(), (*a1)());
 
-    TigreEngine::with_current(|engine| {
-        // Update the top cell with the new number.
-        // See `make_s` for details.
-        let top_cell_ptr = CellPtr((a1 as *mut u8).offset(-CALL_LEN) as *mut Cell);
-        let top_cell = engine.cell_mut(top_cell_ptr);
-        debug_assert_eq!(top_cell.call_opcode, CALL_OPCODE);
-        top_cell.set_call_addr(comb_LIT as usize);
-        top_cell.arg = res;
-    });
+            TigreEngine::with_current(|engine| {
+                // Update the top cell with the new number.
[…]

### 2023-01-09 experiments/superg/src/tigre/comb.rs
Make work with multiple threads.

@@ -58,9 +58,24 @@ global_asm!(
         mov rdi, [rsp]
         mov rsi, [rsp+8]
         mov rdx, [rsp+16]
+
+        // Align stack to 16 bytes if needed
+        mov rax, rsp
+        and rax, 8
+        cmp rax, 0
+        jne comb_S_need_align
+
+    comb_S_no_align:
         call make_s
         // Pop arguments and jump to the updated node 
         add rsp, 24
+        jmp rax
+
+    comb_S_need_align:
+        sub rsp, 8
+        call make_s
+        // Pop arguments and jump to the updated node 
+        add rsp, 32
         jmp rax
     "#
 );
@@ -80,6 +95,7 @@ unsafe extern "C" fn make_s(f: *const CellPtr, g: *const CellPtr, x: *const Cell
         debug_assert_eq!(top_cell.call_opcode, CALL_OPCODE);
         top_cell.set_call_addr(fx.0 as usize);
         top_cell.arg = gx.0 as i64;
+
         top_cell_ptr
     })
 }
@@ -95,11 +111,29 @@ macro_rules! comb_bin_op {
                 // Load argument pointers as arguments
                 mov rdi, [rsp]
                 mov rsi, [rsp+8]
+
+                // Align stack to 16 bytes if needed
+                mov rax, rsp
+                and rax, 8
+                cmp rax, 0
             "#,
+            concat!("jne ", $name_str, "_need_align"),
+            // No alignment needed
+            concat!($name_str, "_no_align:"),
             concat!("call ", $helper_func_str),
             r#"
                 // Pop arguments
                 add rsp, 16
+                // Return the computed value
+                ret
+            "#,
+            // Alignment needed
+            concat!($name_str, "_need_align:"),
+            "sub rsp, 8",
+            concat!("call ", $helper_func_str),
+            r#"
+                // Pop arguments
+                add rsp, 24
                 // Return the computed value
                 ret
             "#,
@@ -109,10 +143,12 @@ macro_rules! comb_bin_op {
         }
         #[no_mangle]
         unsafe extern "C" fn $helper_func(a0: *const ArgFn, a1: *const ArgFn) -> i64 {
-            let op: fn(i64, i64) -> i64 = $op;
-            let res = op((*a0)(), (*a1)());
-
+            // Place all code within `TigreEngine::with_current` since that also takes
[…]

### 2023-01-09 experiments/superg/src/tigre/mod.rs

@@ -1,7 +1,7 @@
 pub mod comb;
 pub mod jit;
 
-use std::collections::HashMap;
+use std::{cell::UnsafeCell, collections::HashMap, panic::UnwindSafe};
 
 use jit::JitMem;
 
@@ -60,7 +60,11 @@ impl IntoCellPtr for Comb {
             Comb::S => CellPtr(comb::comb_S as *mut Cell),
             Comb::K => CellPtr(comb::comb_K as *mut Cell),
             Comb::Plus => CellPtr(comb::comb_plus as *mut Cell),
+            Comb::Minus => CellPtr(comb::comb_min as *mut Cell),
             Comb::Cond => CellPtr(comb::comb_cond as *mut Cell),
+            Comb::Eq => CellPtr(comb::comb_eq as *mut Cell),
+            Comb::Times => CellPtr(comb::comb_times as *mut Cell),
+            Comb::Lt => CellPtr(comb::comb_lt as *mut Cell),
             _ => unimplemented!("into_cell_ptr {:?}", self),
         }
     }
@@ -92,6 +96,16 @@ impl TigreEngine {
     pub fn compile<C: ExprCompiler>(compiler: &mut C, program: &ast::Program) -> TigreEngine {
         let mut jit_placer = JitPlacer::new();
         jit_placer.add_target(comb::comb_LIT as usize);
+        jit_placer.add_target(comb::comb_Abort as usize);
+        jit_placer.add_target(comb::comb_I as usize);
+        jit_placer.add_target(comb::comb_K as usize);
+        jit_placer.add_target(comb::comb_S as usize);
+        jit_placer.add_target(comb::comb_plus as usize);
+        jit_placer.add_target(comb::comb_min as usize);
+        jit_placer.add_target(comb::comb_eq as usize);
+        jit_placer.add_target(comb::comb_times as usize);
+        jit_placer.add_target(comb::comb_cond as usize);
+        jit_placer.add_target(comb::comb_lt as usize);
         let mut engine = TigreEngine {
             mem: jit_placer
                 .place_jit(JIT_MEM_LEN)
@@ -115,16 +129,23 @@ impl TigreEngine {
     }
 
     pub fn run(&mut self) -> i64 {
+        eprintln!("Test124");
         let main_ptr = self.def_lookup.get("main").expect("No main function");
         let func: fn() -> i64 = unsafe { std::mem::transmute(main_ptr.0) };
 
         // Configure the global engine reference
-        unsafe { ENGINE = self as *mut TigreEngine };
[…]

### 2023-01-09 experiments/superg/src/tigre/mod.rs
Get ackermann running.

@@ -83,7 +83,7 @@ impl IntoCellPtr for i32 {
     }
 }
 
-const JIT_MEM_LEN: usize = std::mem::size_of::<Cell>() * 10_000;
+const JIT_MEM_LEN: usize = std::mem::size_of::<Cell>() * 1_000_000;
 
 pub struct TigreEngine {
     mem: JitMem,
@@ -394,10 +394,9 @@ mod tests {
         );
     }
 
-    /*
-        #[test]
-        fn test_ackermann() {
-            let program = r#"
+    #[test]
+    fn test_ackermann() {
+        let program = r#"
     (defun ack (x z) (if (= x 0)
                          (+ z 1)
                          (if (= z 0)
@@ -405,9 +404,8 @@ mod tests {
                              (ack (- x 1) (ack x (- z 1))))))
     (defun main () (ack 3 4))
         "#;
-            assert_runs_to_int(program, 125);
-        }
-        */
+        assert_runs_to_int(program, 125);
+    }
 
     fn assert_runs_to_int(program: &str, v: i64) {
         let parsed = parse(lex(program));

### 2023-01-15 experiments/superg/build.rs
Generated bulk combinator for tigre.

@@ -1,7 +1,7 @@
 use std::fs::File;
 use std::io::Write;
 
-// Generate bulk combinators Sn,Bn and Cn up to this n
+// Generate bulk combinators Sn, Bn and Cn up to this n
 const BULK_COMB_MAX_N: usize = 10;
 
 fn main() -> std::io::Result<()> {
@@ -10,13 +10,16 @@ fn main() -> std::io::Result<()> {
     // This script depends only on itself
     println!("cargo:rerun-if-changed=build.rs");
 
-    let combgen = std::fs::File::create(&format!("{}/combgen.rs", out_dir))?;
-    generate_combgen(combgen)?;
+    let turner_combinators = std::fs::File::create(&format!("{}/turner_combinators.rs", out_dir))?;
+    generate_turner_combinators(turner_combinators)?;
+
+    let tigre_combinators = std::fs::File::create(&format!("{}/tigre_combinators.rs", out_dir))?;
+    generate_tigre_combinators(tigre_combinators)?;
 
     Ok(())
 }
 
-fn generate_combgen(mut f: File) -> std::io::Result<()> {
+fn generate_turner_combinators(mut f: File) -> std::io::Result<()> {
     let max_combn = std::cmp::max(
         // Regular combinators use up to 3 arguments.
         3,
@@ -94,5 +97,74 @@ impl Comb {{
     )?;
     writeln!(f, "    }}")?;
     writeln!(f, "}}")?;
+    Ok(())
+}
+
+fn generate_tigre_combinators(mut f: File) -> std::io::Result<()> {
+    writeln!(f, "use super::*;")?;
+
+    for i in 2..=BULK_COMB_MAX_N {
+        for (k, k_upper) in [('s', 'S'), ('b', 'B'), ('c', 'C')] {
+            writeln!(
+                f,
+                "macros::bulk_comb_{k}!(comb_{k_upper}{}, {}, {}, {}, make_{k}{});",
+                // Number of arguments to distribute
+                i,
+                // Number of values on the stack (f, g and xs)
+                i + 2,
+                // Size of the values on the stack, in bytes
+                (i + 2) * 8,
+                // Size of the values on the stack, plus an extra 8 bytes for alignment to a 16 byte boundary.
+                (i + 3) * 8,
+                i
+            )?;
+        }
[…]

### 2023-01-15 experiments/superg/src/tigre/comb/mod.rs

@@ -5,10 +5,9 @@ use super::{Cell, CellPtr, TigreEngine, CALL_LEN, CALL_OPCODE};
 #[macro_use]
 mod macros;
 
-pub const ALL_COMBINATORS: &[unsafe extern "C" fn() -> usize] = &[
-    comb_LIT, comb_Abort, comb_I, comb_K, comb_S, comb_B, comb_C, comb_plus, comb_min, comb_eq,
-    comb_lt, comb_times, comb_cond, comb_S2, comb_B2, comb_C2,
-];
+// Bulk combinators are automatically generated using a build script.
+mod generated;
+pub use generated::*;
 
 // LIT combinator
 global_asm!(
@@ -134,34 +133,6 @@ macros::comb3!(comb_cond, apply_cond, |engine, c, t, f| {
 
     *branch_ptr
 });
-
-/*
-macros::bulk_comb!(comb_S2, 4, 32, 40, make_s2, |engine, args| {
-    let f = *args[0];
-    let g = *args[1];
-
-    let mut left_cell = f;
-    let mut right_cell = g;
-    for arg in &args[2..] {
-        left_cell = engine.make_cell(left_cell, **arg);
-        right_cell = engine.make_cell(right_cell, **arg);
-    }
-
-    // Update the top cell to point to the new left_cell and right_cell
-    // See `make_s` for details.
-    let top_arg_ptr = *args.last().unwrap() as *const CellPtr;
-    let top_cell_ptr = CellPtr((top_arg_ptr as *mut u8).offset(-CALL_LEN) as *mut Cell);
-    let top_cell = engine.cell_mut(top_cell_ptr);
-    debug_assert_eq!(top_cell.call_opcode, CALL_OPCODE);
-    top_cell.set_call_addr(left_cell.0 as usize);
-    top_cell.arg = right_cell.0 as i64;
-
-    top_cell_ptr
-});
-*/
-macros::bulk_comb_s!(comb_S2, 4, 32, 40, make_s2);
-macros::bulk_comb_b!(comb_B2, 4, 32, 40, make_b2);
-macros::bulk_comb_c!(comb_C2, 4, 32, 40, make_c2);
 
 #[cfg(test)]
 mod tests {

### 2023-01-15 experiments/superg/src/tigre/mod.rs

@@ -68,9 +68,9 @@ impl IntoCellPtr for Comb {
             Comb::Eq => CellPtr(comb::comb_eq as *mut Cell),
             Comb::Times => CellPtr(comb::comb_times as *mut Cell),
             Comb::Lt => CellPtr(comb::comb_lt as *mut Cell),
-            Comb::Sn(2) => CellPtr(comb::comb_S2 as *mut Cell),
-            Comb::Bn(2) => CellPtr(comb::comb_B2 as *mut Cell),
-            Comb::Cn(2) => CellPtr(comb::comb_C2 as *mut Cell),
+            Comb::Sn(n) => CellPtr(comb::COMB_SN_IMPLS[*n] as *mut Cell),
+            Comb::Bn(n) => CellPtr(comb::COMB_BN_IMPLS[*n] as *mut Cell),
+            Comb::Cn(n) => CellPtr(comb::COMB_CN_IMPLS[*n] as *mut Cell),
             _ => unimplemented!("into_cell_ptr {:?}", self),
         }
     }

### 2023-01-15 experiments/superg/src/turner/mod.rs

@@ -8,7 +8,7 @@ use crate::ast;
 mod macros;
 // Bulk combinators are automatically generated using a build script.
 mod combgen {
-    include!(concat!(env!("OUT_DIR"), "/combgen.rs"));
+    include!(concat!(env!("OUT_DIR"), "/turner_combinators.rs"));
 }
 
 // TODO: Allow dynamically growing the heap.

### 2023-01-15 experiments/superg/src/tigre/comb/macros.rs
Cleanup comb.rs.

@@ -1,3 +1,17 @@
+//! Macros to generate implementations of combinators that require calls to Rust code.
+//!
+//! # Alignment
+//! The System V ABI that Rust uses on x86-64 Linux (at least for an extern "C" `fn`) mandates that before a `call` instruction, the stack is aligned to 16 bytes (as documented [here](https://c9x.me/compile/doc/abi.html)).
+//! In particular it seems that SIMD instructions require this alignment, without it I have observed crashes in Rust library code that uses such instructions.
+//!
+//! We regularly chain `call` instructions until a combinator a reached, so we cannot guarantee that the stack is aligned properly.
+//! What we do know for sure is that the stack is aligned to 8 bytes, because we always push 64-bit values on the stack.
+//! So, before we can jump to a helper function in Rust, we must check if the stack is aligned to 16 bytes.
+//! If it is not, we jump to a `*_need_align` label and substract 8 bytes from the stack pointer.
+//! Since the stack pointer is always aligned to at least 8 bytes, this must result in alignment to 16 bytes.
+//! If the stack is already properly aligned, we can simply jump to the function.
+//! After the call returns, we pop the argument off the stack, and if extra alignment was needed, add an extra 8 bytes offset to the stack pointer.
+
 // Generate an assembly target and helper function for a combinator that takes three arguments.
 macro_rules! comb3 {
     ($name:ident, $helper_func:ident, $impl:expr) => {
@@ -107,14 +121,8 @@ macro_rules! comb_bin_op {
                 let op: fn(i64, i64) -> i64 = $op;
                 let res = op((*a0)(), (*a1)());
 
-                // Update the top cell with the new number.
-                // See `make_s` for details.
-                let top_cell_ptr = CellPtr((a1 as *mut u8).offset(-CALL_LEN) as *mut Cell);
-                let top_cell = engine.cell_mut(top_cell_ptr);
-                debug_assert_eq!(top_cell.call_opcode, CALL_OPCODE);
-                top_cell.set_call_addr(comb_LIT as usize);
-                top_cell.arg = res;
-
+                // Update the top cell to the literal value
+                engine.update_top_cell(Lit, res, a1 as *const CellPtr);
                 res
             })
         }
@@ -122,6 +130,7 @@ macro_rules! comb_bin_op {
 }
[…]

### 2023-01-15 experiments/superg/src/tigre/comb/mod.rs

@@ -1,13 +1,16 @@
 use std::arch::global_asm;
 
-use super::{Cell, CellPtr, TigreEngine, CALL_LEN, CALL_OPCODE};
+use crate::compiled_expr::Comb;
+
+use super::{Cell, CellPtr, IntoCellPtr, TigreEngine, CALL_LEN, CALL_OPCODE};
 
 #[macro_use]
-mod macros;
+pub mod macros;
 
+/// Implementations of bulk combinators.
 // Bulk combinators are automatically generated using a build script.
-mod generated;
-pub use generated::*;
+pub mod bulk;
+pub use bulk::ALL_COMBINATORS;
 
 // LIT combinator
 global_asm!(
@@ -21,6 +24,14 @@ global_asm!(
 );
 extern "C" {
     pub fn comb_LIT() -> usize;
+}
+// The LIT combinator is an implementation detail of Tigre, and therefore not added to the `Comb` enum.
+// Instead we define a struct type for it, so it can be used as an argument in functions generic over [`IntoCellPtr`]
+pub struct Lit;
+impl IntoCellPtr for Lit {
+    fn into_cell_ptr(&self, _: &TigreEngine) -> CellPtr {
+        CellPtr(comb_LIT as *mut Cell)
+    }
 }
 
 // Abort combinator
@@ -56,47 +67,45 @@ extern "C" {
     pub fn comb_K() -> usize;
 }
 
+impl TigreEngine {
+    // Updates the left and right values of the cell for which `top_arg_ptr` points to the right (argument) value.
+    unsafe fn update_top_cell<CP0: IntoCellPtr, CP1: IntoCellPtr>(
+        &mut self,
+        left: CP0,
+        right: CP1,
+        top_arg_ptr: *const CellPtr,
+    ) -> CellPtr {
+        let left = left.into_cell_ptr(self);
+        let right = right.into_cell_ptr(self);
+        // Pointers on the stack are to the right pointer of a cell, after the call instruction.
+        // To get a pointer to the full cell, we subtract the length of a call instruction.
+        let top_cell_ptr = CellPtr((top_arg_ptr as *mut u8).offset(-CALL_LEN) as *mut Cell);
+        let top_cell = self.cell_mut(top_cell_ptr);
+        debug_assert_eq!(top_cell.call_opcode, CALL_OPCODE);
+        top_cell.set_call_addr(left.0 as usize);
+        top_cell.arg = right.0 as i64;
[…]

### 2023-01-15 experiments/superg/src/tigre/mod.rs

@@ -14,6 +14,7 @@ use crate::{
 use self::jit::JitPlacer;
 
 // Cells are 16 bytes for optimal alignment.
+// A unit test at the end of this file verifies the size of this type.
 // See https://groups.google.com/g/golang-nuts/c/dhGbgC1pAmA/m/Rcqwcd5mGmoJ.
 #[repr(packed)]
 pub struct Cell {
@@ -68,22 +69,21 @@ impl IntoCellPtr for Comb {
             Comb::Eq => CellPtr(comb::comb_eq as *mut Cell),
             Comb::Times => CellPtr(comb::comb_times as *mut Cell),
             Comb::Lt => CellPtr(comb::comb_lt as *mut Cell),
-            Comb::Sn(n) => CellPtr(comb::COMB_SN_IMPLS[*n] as *mut Cell),
-            Comb::Bn(n) => CellPtr(comb::COMB_BN_IMPLS[*n] as *mut Cell),
-            Comb::Cn(n) => CellPtr(comb::COMB_CN_IMPLS[*n] as *mut Cell),
+            Comb::Sn(n) => CellPtr(comb::bulk::COMB_SN_IMPLS[*n] as *mut Cell),
+            Comb::Bn(n) => CellPtr(comb::bulk::COMB_BN_IMPLS[*n] as *mut Cell),
+            Comb::Cn(n) => CellPtr(comb::bulk::COMB_CN_IMPLS[*n] as *mut Cell),
             _ => unimplemented!("into_cell_ptr {:?}", self),
         }
     }
 }
 
-struct Lit;
-impl IntoCellPtr for Lit {
+impl IntoCellPtr for i32 {
     fn into_cell_ptr(&self, _: &TigreEngine) -> CellPtr {
-        CellPtr(comb::comb_LIT as *mut Cell)
+        CellPtr(*self as *mut Cell)
     }
 }
 
-impl IntoCellPtr for i32 {
+impl IntoCellPtr for i64 {
     fn into_cell_ptr(&self, _: &TigreEngine) -> CellPtr {
         CellPtr(*self as *mut Cell)
     }
@@ -194,7 +194,7 @@ impl TigreEngine {
 
     fn alloc_expr(&mut self, expr: CompiledExpr) -> CellPtr {
         match expr {
-            CompiledExpr::Int(i) => self.make_cell(Lit, i),
+            CompiledExpr::Int(i) => self.make_cell(comb::Lit, i),
[…]

### 2023-01-16 experiments/superg/doc/paper/paper.tex
Implement cond combinator.
Get ackermann running.

--- 2022-12-19 experiments/superg/doc/paper/paper.tex
@@
 \DeclareMathOperator{\linear}{linear}
+
+\usepackage{listings}
 
@@
 
-To reduce this graph, we follow the left pointers until we find a combinator to apply.
+To reduce this graph, we follow the left pointer until we find a combinator to apply.
 As we traverse the graph along what is often called the \textit{spine}, we keep the nodes we have seen on a stack, because their right nodes contain the arguments to the combinator we eventually find.
@@
 \label{sec:tigre}
-\textbf{TODO}
+The TIGRE engine, first presented in~\cite{koopman_fresh_1989}, is a combinator graph reduction engine like Miranda.
+Originally designed for the VAX instruction set, our implementation targets the x86-64 instruction set but mirrors the TIGRE design otherwise.
+The TIGRE paper focuses on key differences from Miranda, the elimination of tags and directly executable graph nodes, so where the paper omits implementation details we take the same approach as our reference implementation based on Miranda.
 
-Describe changes made to align with TIGRE.
+\subsection{Eliminating tags}
+The TIGRE authors point to tags as a major source of inefficiency in graph reduction engines.
+At evaluation time Miranda must check the tag of nodes to see if their right-hand side is a value or a pointer to a node.
+TIGRE does away with this by adding a special \texttt{LIT} combinator.
+The \texttt{LIT} combinator takes one argument guaranteed to be a literal value and returns it, like an \texttt{I} combinator specialized for constants.
+An apparent downside of this is that simple expressions require more indirection, and therefore a larger graph to reduce.
+Take for example the expression $1+2$, which in the Miranda-style engine can be represented with just two nodes, shown in figure~\ref{fig:1_plus_2_miranda}.
 
-Compare performance with Miranda-style engine.
+\begin{figure}
+    \includegraphics[width=.2\columnwidth]{1_plus_2_miranda}
+    \centering
+    \caption{
+        Graph of $1 + 2$ in the Miranda-style engine.
+        Tags omitted for brevity.
+    }
+    \label{fig:1_plus_2_miranda}
+\end{figure}
+
+The TIGRE version of this program, shown in figure~\ref{fig:1_plus_2_tigre}, is larger, occupying four nodes.
+
+\begin{figure}
+    \includegraphics[width=.46\columnwidth]{1_plus_2_tigre}
+    \centering
+    \caption{
+        Graph of $1 + 2$ in the TIGRE-style engine.
+    }
+    \label{fig:1_plus_2_tigre}
+\end{figure}
+
+In this example, the TIGRE graph may indeed be slower to evaluate, but the same does not hold for the programs in general.
+If the values $1$ and $2$ were the results of a previous computation, they would also be behind an indirection node in the Miranda engine.
+The TIGRE approach is only slower when performing operations on values that are constant in the source program, in other cases the reduced branching on tag values is expected to speed up execution.
+
+The TIGRE authors report that some engines also use the tag to distinguish between combinators and cell pointers.
+In Miranda the lowest cell pointers are reserved for combinators, so this does not need to be encoded in the tag, but the engine still needs to do a case distinction for low values of the pointer.
+TIGRE also eliminates this overhead by encoding nodes as executable machine instructions.
+We discuss this encoding in more detail in the next section.
+With this change any pointer in the graph points to executable code: nodes are directly executable, and for combinators, we can insert a pointer to their implementation function.
+The overhead of interpreting tags is thus fully eliminated.
+
+Our implementation still has a tag byte in each cell, but it is exclusively used by the garbage collector to keep track of free and occupied cells.
+The TIGRE paper provides no details on garbage collection, so we adopt the same mark-and-sweep collection approach as Miranda.
+
+\subsection{Executable graph nodes}
+Due to the elimination of tags, graph nodes must be executable x86-64 machine instructions.
+Executing a node is expected to \emph{unwind the spine}: a pointer to the current node should be pushed onto the stack, after which we proceed to evaluate the cell to the left.
+It turns out that this can be neatly encoded as a subroutine call in both the VAX and x86-64 instruction sets.
+
+The x86-64 \texttt{CALL} \textit{rel32} instruction\footnote{\url{https://c9x.me/x86/html/file_module_x86_id_26.html}} modifies the processor state as follows:
+\begin{enumerate}
+    \item Pushes onto the stack a pointer to the instruction directly after the \texttt{CALL}.
+          In a regular program, this is the \emph{return address}, the address the program should jump to after the subroutine finishes.
+    \item Adds \textit{rel32} to the program counter register \texttt{\%rip}.
+          \textit{rel32} encodes the address of the subroutine as a signed 32-bit offset from the location of the \texttt{CALL}.
+\end{enumerate}
+
+We exploit this by encoding nodes as a \texttt{CALL} instruction to the left pointer, followed by the right pointer as a raw value.
+The memory layout of a node is illustrated in figure~\ref{fig:tigre_layout}.
+
+\begin{figure*}
+    \includegraphics[width=2\columnwidth]{tigre_layout}
+    \centering
+    \caption{
+        Memory layout of a node in the TIGRE-style engine.
+        Each square represents a byte of memory.
+    }
+    \label{fig:tigre_layout}
+\end{figure*}
+
+When execution jumps to a node, the \texttt{CALL} instruction executes.
+The address of the right pointer is pushed to the stack.
+This is convenient because when later a combinator executes, it only needs to read the value of the right pointer (the left is only needed to unwind the stack).
+Execution then jumps to the address of the left cell pointer, which is either another node or the code for a combinator.
+The \texttt{CALL} instruction and 64-bit right pointer take up 13 bytes.
+We add an extra byte to store the tag (only one bit is currently used) and 2 bytes of padding so the full node is 16 bytes wide.
+Having 16-byte wide nodes is convenient because Intel processors can execute jumps faster if the target is aligned to a 16-byte boundary\footnote{\url{https://groups.google.com/g/golang-nuts/c/dhGbgC1pAmA/m/Rcqwcd5mGmoJ}}.
+
+The simpler combinators are easily expressed as a short sequence of assembly instructions.
+Figure~\ref{fig:combI} shows the implementation of the I combinator.
+
+\begin{figure}
+    \begin{lstlisting}[
+        language={[x86masm]Assembler},
+    ]
+    comb_I:
+        ; Pop pointer to argument
+        ; into rax
+        pop rax
+        ; Dereference and jump
+        jmp [rax]
+    \end{lstlisting}
+    \centering
+    \caption{I combinator in x86-64 assembly.}
+    \label{fig:combI}
+\end{figure}
+
+Combinators that return a value (\texttt{LIT} or strict combinators) can be implemented by computing the value and returning it with \texttt{RET}.
+See for example the implementation of \texttt{LIT} in figure~\ref{fig:combLIT}.
+
+\begin{figure}
+    \begin{lstlisting}[
+        language={[x86masm]Assembler},
+    ]
+    comb_LIT:
+        ; Pop pointer to argument
+        ; into rax
+        pop rax
+        ; Dereference and store 
+        ; in return value register
+        mov rax, [rax]
+        ; Return to caller
+        ret
+    \end{lstlisting}
+    \centering
+    \caption{LIT combinator in x86-64 assembly.}
+    \label{fig:combLIT}
+\end{figure}
+
+A strict combinator that needs to evaluate its arguments is easier to implement in a TIGRE-style engine compared to Miranda: the implementation can strictly evaluate arguments by jumping to the argument node with a \texttt{CALL} instruction, and control will return to the combinator implementation after the argument has been evaluated and the \texttt{RET} instruction executes.
+The fully evaluated argument is then available in the \texttt{\%rax} register.
+This even works from Rust code: a pointer to a node can be cast to a function pointer with an integer as a return value and invoked normally.
+
+\textbf{TODO: Maybe add a step-by-step reduction example?}
+
+To our surprise, we did not find the TIGRE-style engine more difficult to implement than the Miranda version, mainly because the reduction process happens to map rather naturally to subroutine calls.
+We discuss the main challenges encountered during the implementation in the next section.
+
+\subsection{Implementation issues}
+\subsubsection{Writeable and executable memory}
+For our reduction engine to work, nodes must be stored in executable memory.
+Modern x86-64 CPUs support the NX-bit~\cite{paulson_new_2004}, which allows the operating system to mark regions of the memory as non-executable, triggering an exception if a program jumps to a page with the NX-bit set.
+The NX-bit is typically enabled by default on dynamically allocated heap memory.
+To allocate memory without the NX-bit set on Linux, we allocate memory for the graph using \texttt{mmap}\footnote{\url{https://man7.org/linux/man-pages/man2/mmap.2.html}} with the \texttt{PROT\_EXEC} (to allow execution) and \texttt{PROT\_WRITE} (to allow writing) flags.
+Some operating systems do not allow unprivileged programs to set \texttt{PROT\_EXEC} and \texttt{PROT\_WRITE} at the same time.
+One solution to this is to flip the memory protection settings from executable to writeable, make the changes, and flip back to executable and continue execution~\cite{de_mooij_wx_2015}.
+We anticipate this would be prohibitively expensive for our engine: After a combinator executes, it updates the graph with the new value, so that if the same value is used in a different subgraph, it is only evaluated once (just as Miranda does).
+The engine would have to issue two \texttt{mprotect} system calls at every reduction step, a heavy performance penalty.
+Luckily, the author's Linux distribution does not enforce a W \^{} X policy and allows the allocation of memory that is both writeable and executable.
+
+\subsubsection{Relative jumps}
+The \texttt{CALL} instruction takes as argument a 32-bit offset relative to the instruction pointer (the address of the \texttt{CALL} instruction), rather than a full 64-bit address.
+Therefore, some pointer arithmetic is required to calculate the value of this offset.
+More importantly, a 32-bit offset is too small to reach any target within the 64-bit address space.
+This is problematic for large programs where nodes may be very far apart.
+The programs we have used to evaluate the engine are relatively small, so we did not encounter such an issue.
+However, the offset may also point to an implementation of a combinator.
+The location of this code is determined by the linker when the engine itself is compiled, so it is difficult to influence its location in memory.
+To get around this, our engine registers the addresses of all combinators.
+Given this set of addresses, the engine can compute the \emph{safe range}: a region of the address space from which a 32-bit offset would be sufficient to reach any combinator.
+When allocating memory to store the graph nodes, it finds an unmapped page within the safe range and allocates it.
+As an additional safeguard, all offsets are initially computed as 64-bit values, after which the engine verifies that the offset indeed fits in 32 bits.
+
+The technique is not without its problems: for large programs (or a large set of combinators), allocation of large enough block of memory may be unsuccessful.
+A more robust solution would be to use trampolines: locations in memory close to nodes that contain assembly instructions to jump to locations far away in memory.
+This requires more bookkeeping and is therefore left as future work.
+
+\subsection{Stack alignment}
+The System-V ABI that is used on x86-64-Linux mandates that before a \texttt{CALL} instruction the stack is aligned to 16 bytes\footnote{https://c9x.me/compile/doc/abi.html}.
+Our engine regularly chains \texttt{CALL} instructions (pushing an 8-byte return address onto the stack) until a combinator is reached, so it is impossible to guarantee statically the alignment of the stack upon a call to a combinator implementation.
+The entry point of each combinator that needs to call into Rust code checks the alignment of the stack right before the \texttt{CALL} instruction and adjusts it if necessary.
+
+\subsection{Panicking inside combinators}
+Rust has the concept of \emph{panic}: an exceptional situation that indicates a bug in the program, for example, an index out-of-bounds error.
+If a panic occurs, the program collects a stack trace and terminates with an error code.
+A panic inside of a combinator implementation may cause the program to unwind into the node graph, which is undesirable because it is undefined behaviour\footnote{\url{https://doc.rust-lang.org/nomicon/ffi.html\#ffi-and-unwinding}}.
+To prevent this, all combinator implementations are wrapped in a call to \emph{catch\_unwind}\footnote{\url{https://doc.rust-lang.org/std/panic/fn.catch_unwind.html}}, so we catch panics before they leave Rust code, log them and abort the program.
 

### 2023-01-17 experiments/superg/benches/bench.rs
Benchmark Miranda.

@@ -1,0 +1,49 @@
+use criterion::{criterion_group, criterion_main, Criterion};
+use superg::compiled_expr::ExprCompiler;
+
+use std::time::{Duration, Instant};
+use superg::bracket::BracketCompiler;
+use superg::turner::TurnerEngine;
+use superg::{lex, parse, Engine};
+
+mod miranda;
+
+criterion_group!(turner, bench_fib_20);
+criterion_main!(turner, miranda::benches);
+
+const FIB_20: &str = r#"
+    (defun fib (n)
+      (if (< n 2) 
+          n
+          (+ (fib (- n 1)) (fib (- n 2)))))
+    (defun main () (fib 20))
+"#;
+
+pub fn bench_fib_20(c: &mut Criterion) {
+    assert_runs_to_int::<_, TurnerEngine>(BracketCompiler, c, "turner_fib_20", FIB_20, 6765);
+}
+
+fn assert_runs_to_int<C: ExprCompiler, E: Engine>(
+    mut compiler: C,
+    c: &mut Criterion,
+    test_name: &str,
+    program: &str,
+    v: i32,
+) {
+    let parsed = parse(lex(program));
+
+    c.bench_function(test_name, |b| {
+        b.iter_custom(|iters| {
+            let mut elapsed = Duration::ZERO;
+            for _ in 0..iters {
+                let mut engine = E::compile(&mut compiler, &parsed);
+                // Measure only the time spent reducing the graph
+                let start = Instant::now();
+                let res = engine.run();
+                elapsed += start.elapsed();
+                assert_eq!(res, v);
+            }
+            elapsed
+        })
+    });
+}

### 2023-01-17 experiments/superg/benches/benchmarks.rs

@@ -1,38 +1,0 @@
-use criterion::{criterion_group, criterion_main, Criterion};
-use std::time::{Duration, Instant};
-use superg::turner::TurnerEngine;
-use superg::{lex, parse};
-
-pub fn bench_ackermann(c: &mut Criterion) {
-    let program = r#"
-(defun ack (x z) (if (= x 0)
-                     (+ z 1)
-                     (if (= z 0)
-                         (ack (- x 1) 1)
-                         (ack (- x 1) (ack x (- z 1))))))
-(defun main () (ack 3 4))
-    "#;
-    assert_runs_to_int(c, "bench_ackermann", program, 125);
-}
-
-fn assert_runs_to_int(c: &mut Criterion, test_name: &str, program: &str, v: i32) {
-    let parsed = parse(lex(program));
-
-    c.bench_function(test_name, |b| {
-        b.iter_custom(|iters| {
-            let mut elapsed = Duration::ZERO;
-            for _ in 0..iters {
-                let mut engine = TurnerEngine::compile(&parsed);
-                // Measure only the time spent reducing the graph
-                let start = Instant::now();
-                let ptr = engine.run();
-                elapsed += start.elapsed();
-                assert_eq!(engine.get_int(ptr), Some(v));
-            }
-            elapsed
-        })
-    });
-}
-
-criterion_group!(benches, bench_ackermann);
-criterion_main!(benches);

### 2023-01-17 experiments/superg/benches/miranda.rs

@@ -1,0 +1,80 @@
+use criterion::{criterion_group, Criterion};
+use regex::Regex;
+use std::io::Write;
+use std::path::Path;
+use std::process::{Command, Stdio};
+use std::time::Duration;
+
+criterion_group!(benches, bench_fib_20);
+
+pub fn bench_fib_20(c: &mut Criterion) {
+    bench_fib(20, c)
+}
+
+fn bench_fib(n: u16, c: &mut Criterion) {
+    bench_miranda(
+        c,
+        &format!("miranda_fib_{n}"),
+        r#"
+fib n = n,                              if n <= 2
+      = fib(n - 1) + fib(n - 2),        otherwise
+        "#,
+        &format!("fib {n}"),
+    );
+}
+
+fn bench_miranda(c: &mut Criterion, test_name: &str, module_file_content: &str, command: &str) {
+    // Create the temp file with the module contents
+    let mut temp_file = tempfile::Builder::new().suffix(".m").tempfile().unwrap();
+    write!(temp_file, "{}", module_file_content).unwrap();
+    temp_file.flush().unwrap();
+    c.bench_function(test_name, |b| {
+        b.iter_custom(|iters| {
+            let mut elapsed = Duration::ZERO;
+
+            for _ in 0..iters {
+                elapsed += run_miranda(temp_file.path(), command).runtime;
+            }
+            elapsed
+        })
+    });
+}
+
+#[derive(Debug, Clone, Copy, PartialEq)]
+struct Stats {
+    runtime: Duration,
+    cycles: usize,
+}
+
+fn run_miranda(file_path: &Path, command: &str) -> Stats {
+    let miranda_path = std::env::var("MIRANDA_PATH").expect("'MIRANDA_PATH' must be set");
+    let mut mira = Command::new("./mira")
+        .arg(file_path)
+        .current_dir(miranda_path)
+        .stdin(Stdio::piped())
+        .stdout(Stdio::null())
+        .stderr(Stdio::piped())
+        .spawn()
+        .expect("error starting miranda");
+    {
+        let mut stdin = mira.stdin.take().unwrap();
+        writeln!(stdin, "{}", command).unwrap();
+        // stdin dropped and closed here
+    }
+
+    let output = mira.wait_with_output().expect("error running miranda");
+
+    let stderr = String::from_utf8(output.stderr).unwrap();
+
[…]

### 2023-01-17 experiments/superg/benches/bench.rs
Benchmarking.

@@ -1,6 +1,8 @@
-use criterion::{criterion_group, criterion_main, Criterion};
+use criterion::{criterion_group, criterion_main, Bencher, BenchmarkId, Criterion};
 use superg::compiled_expr::ExprCompiler;
+use superg::tigre::TigreEngine;
 
+use std::io::Write;
 use std::time::{Duration, Instant};
 use superg::bracket::BracketCompiler;
 use superg::turner::TurnerEngine;
@@ -8,42 +10,96 @@ use superg::{lex, parse, Engine};
 
 mod miranda;
 
-criterion_group!(turner, bench_fib_20);
-criterion_main!(turner, miranda::benches);
+criterion_group!(turner_vs_miranda, bench_fib_turner_vs_miranda);
+criterion_group!(turner_vs_tigre, bench_fib_turner_vs_tigre);
+criterion_main!(turner_vs_miranda, turner_vs_tigre);
 
-const FIB_20: &str = r#"
+fn fib_program(n: u16) -> String {
+    format!(
+        r#"
     (defun fib (n)
       (if (< n 2) 
           n
           (+ (fib (- n 1)) (fib (- n 2)))))
-    (defun main () (fib 20))
-"#;
-
-pub fn bench_fib_20(c: &mut Criterion) {
-    assert_runs_to_int::<_, TurnerEngine>(BracketCompiler, c, "turner_fib_20", FIB_20, 6765);
+    (defun main () (fib {n}))
+    "#
+    )
 }
 
-fn assert_runs_to_int<C: ExprCompiler, E: Engine>(
-    mut compiler: C,
-    c: &mut Criterion,
-    test_name: &str,
+fn fib(n: u16) -> i32 {
+    if n < 2 {
+        n as i32
+    } else {
+        fib(n - 1) + fib(n - 2)
+    }
+}
+
+fn bench_instance<C: ExprCompiler, E: Engine>(
+    b: &mut Bencher,
     program: &str,
-    v: i32,
+    expected_res: i32,
+    mut compiler: C,
 ) {
     let parsed = parse(lex(program));
+    b.iter_custom(|iters| {
+        let mut elapsed = Duration::ZERO;
+        for _ in 0..iters {
+            let mut engine = E::compile(&mut compiler, &parsed);
+            // Measure only the time spent reducing the graph
+            let start = Instant::now();
+            let res = engine.run();
+            elapsed += start.elapsed();
+            assert_eq!(res, expected_res);
+        }
[…]

### 2023-01-17 experiments/superg/benches/miranda.rs

@@ -1,59 +1,24 @@
-use criterion::{criterion_group, Criterion};
 use regex::Regex;
 use std::io::Write;
 use std::path::Path;
 use std::process::{Command, Stdio};
 use std::time::Duration;
 
-criterion_group!(benches, bench_fib_20);
-
-pub fn bench_fib_20(c: &mut Criterion) {
-    bench_fib(20, c)
-}
-
-fn bench_fib(n: u16, c: &mut Criterion) {
-    bench_miranda(
-        c,
-        &format!("miranda_fib_{n}"),
-        r#"
-fib n = n,                              if n <= 2
-      = fib(n - 1) + fib(n - 2),        otherwise
-        "#,
-        &format!("fib {n}"),
-    );
-}
-
-fn bench_miranda(c: &mut Criterion, test_name: &str, module_file_content: &str, command: &str) {
-    // Create the temp file with the module contents
-    let mut temp_file = tempfile::Builder::new().suffix(".m").tempfile().unwrap();
-    write!(temp_file, "{}", module_file_content).unwrap();
-    temp_file.flush().unwrap();
-    c.bench_function(test_name, |b| {
-        b.iter_custom(|iters| {
-            let mut elapsed = Duration::ZERO;
-
-            for _ in 0..iters {
-                elapsed += run_miranda(temp_file.path(), command).runtime;
-            }
-            elapsed
-        })
-    });
-}
-
 #[derive(Debug, Clone, Copy, PartialEq)]
-struct Stats {
-    runtime: Duration,
-    cycles: usize,
+pub struct Stats {
+    pub runtime: Duration,
+    pub cycles: usize,
+    pub result: i32,
 }
 
-fn run_miranda(file_path: &Path, command: &str) -> Stats {
+pub fn run_miranda(file_path: &Path, command: &str) -> Stats {
     let miranda_path = std::env::var("MIRANDA_PATH").expect("'MIRANDA_PATH' must be set");
     let mut mira = Command::new("./mira")
         .arg(file_path)
         .current_dir(miranda_path)
         .stdin(Stdio::piped())
-        .stdout(Stdio::null())
-        .stderr(Stdio::piped())
+        .stdout(Stdio::piped())
+        .stderr(Stdio::inherit())
         .spawn()
         .expect("error starting miranda");
     {
@@ -64,17 +29,19 @@ fn run_miranda(file_path: &Path, command: &str) -> Stats {
 
[…]

### 2023-01-17 experiments/superg/src/tigre/mod.rs
Improve TIGRE perf.

@@ -32,7 +32,11 @@ impl Cell {
     pub fn set_call_addr(&mut self, target_addr: usize) {
         let base_addr = self as *const Cell as isize + CALL_LEN;
         let call_rel_addr = target_addr as isize - base_addr;
-        self.call_rel_addr = call_rel_addr.try_into().expect("have to jump too far");
+        debug_assert!(
+            call_rel_addr >= (i32::MIN as isize) && call_rel_addr <= (i32::MAX as isize),
+            "have to jump too far"
+        );
+        self.call_rel_addr = call_rel_addr as i32;
     }
 }
 
@@ -211,27 +215,39 @@ impl TigreEngine {
     pub fn with_current<R, F: FnOnce(&mut TigreEngine) -> R + UnwindSafe>(f: F) -> R {
         // This function is always called from a combinator implementation.
         // Unwinding into non-rust code is undefined behaviour, so catch any panics here.
-        let maybe_panic = std::panic::catch_unwind(|| {
+        with_catch_unwind_in_debug(|| {
             let engine_ptr = ENGINE.with(|engine_cell| unsafe { *engine_cell.get() });
             let engine = unsafe {
                 debug_assert!(!engine_ptr.is_null());
                 &mut *engine_ptr
             };
             f(engine)
-        });
+        })
+    }
+}
 
-        match maybe_panic {
-            Ok(res) => res,
-            Err(e) => {
-                if let Some(msg) = e.downcast_ref::<&str>() {
-                    eprintln!("panic within TigreEngine::run: {}", msg);
-                } else {
-                    eprintln!("panic within TigreEngine::run");
-                }
-                std::process::abort();
+// Catching unwinds is expensive, so we only do it in debug mode
+#[cfg(debug_assertions)]
+fn with_catch_unwind_in_debug<R, F: FnOnce() -> R + UnwindSafe>(f: F) -> R {
+    let maybe_panic = std::panic::catch_unwind(f);
+
+    match maybe_panic {
+        Ok(res) => res,
+        Err(e) => {
+            if let Some(msg) = e.downcast_ref::<&str>() {
+                eprintln!("panic within TigreEngine::run: {}", msg);
+            } else {
+                eprintln!("panic within TigreEngine::run");
             }
+            std::process::abort();
         }
     }
+}
+
+// In release mode, this is a no-op.
+#[cfg(not(debug_assertions))]
[…]

### 2023-01-17 experiments/superg/src/tigre/comb/macros.rs
Put engine in r15.

@@ -19,10 +19,11 @@ macro_rules! comb3 {
             concat!(".global ", stringify!($name)),
             concat!(stringify!($name), ":"),
             r#"
+                mov rdi, r15
                 // Load f, g, x pointers as arguments
-                mov rdi, [rsp]
-                mov rsi, [rsp+8]
-                mov rdx, [rsp+16]
+                mov rsi, [rsp]
+                mov rdx, [rsp+8]
+                mov rcx, [rsp+16]
 
                 // Align stack to 16 bytes if needed
                 mov rax, rsp
@@ -53,19 +54,23 @@ macro_rules! comb3 {
         }
         #[no_mangle]
         unsafe extern "C" fn $helper_func(
+            reg_engine: *mut TigreEngine,
             f: *const CellPtr,
             g: *const CellPtr,
             x: *const CellPtr,
         ) -> CellPtr {
+            // TODO: Restore unwinding behaviour
             // Place all code within `TigreEngine::with_current` since that also takes
             // care of catching unwinds before they escape Rust code.
+            let fimpl: fn(
+                &mut TigreEngine,
+                *const CellPtr,
+                *const CellPtr,
+                *const CellPtr,
+            ) -> CellPtr = $impl;
             TigreEngine::with_current(|engine| {
-                let fimpl: fn(
-                    &mut TigreEngine,
-                    *const CellPtr,
-                    *const CellPtr,
-                    *const CellPtr,
-                ) -> CellPtr = $impl;
+                assert!(std::ptr::eq(reg_engine, engine as *mut TigreEngine));
+                println!("OK");
                 fimpl(engine, f, g, x)
             })
         }

### 2023-01-17 experiments/superg/src/tigre/mod.rs

@@ -1,7 +1,7 @@
 pub mod comb;
 pub mod jit;
 
-use std::{cell::UnsafeCell, collections::HashMap, panic::UnwindSafe};
+use std::{arch::global_asm, cell::UnsafeCell, collections::HashMap, panic::UnwindSafe};
 
 use jit::JitMem;
 
@@ -132,7 +132,7 @@ impl Engine for TigreEngine {
 
     fn run(&mut self) -> i32 {
         let main_ptr = self.def_lookup.get("main").expect("No main function");
-        let func: fn() -> i64 = unsafe { std::mem::transmute(main_ptr.0) };
+        let func: extern "C" fn() -> i64 = unsafe { std::mem::transmute(main_ptr.0) };
 
         // Configure the global engine reference
         ENGINE.with(|engine_cell| {
@@ -140,7 +140,7 @@ impl Engine for TigreEngine {
             unsafe { *engine_cell = self as *mut TigreEngine };
         });
 
-        let res = func();
+        let res = unsafe { run_comb_graph(self as *mut TigreEngine, func) };
 
         // Deregister global engine reference
         ENGINE.with(|engine_cell| {
@@ -149,6 +149,24 @@ impl Engine for TigreEngine {
         });
         res.try_into().expect("result does not fit into i32")
     }
+}
+
+global_asm!(
+    r#"
+    .global run_comb_graph
+    run_comb_graph:
+        // Set rbx to engine reference
+        push r15
+        mov r15, rdi
+        call rsi
+        pop r15
+        ret
+"#
+);
+
+extern "C" {
+    #[allow(improper_ctypes)]
+    fn run_comb_graph(engine: *mut TigreEngine, func: extern "C" fn() -> i64) -> i64;
 }
 
 impl TigreEngine {

### 2023-01-23 experiments/superg/benches/bench.rs
Benchmarks and perf improvements to engines.

@@ -1,7 +1,10 @@
-use criterion::{criterion_group, criterion_main, Bencher, BenchmarkId, Criterion};
+use criterion::measurement::WallTime;
+use criterion::{criterion_group, criterion_main, Bencher, BenchmarkGroup, BenchmarkId, Criterion};
 use superg::compiled_expr::ExprCompiler;
+use superg::kiselyov::{LazyCompiler, LazyOptCompiler, LinearCompiler, StrictCompiler};
 use superg::tigre::TigreEngine;
 
+use std::fmt::Display;
 use std::io::Write;
 use std::time::{Duration, Instant};
 use superg::bracket::BracketCompiler;
@@ -12,7 +15,8 @@ mod miranda;
 
 criterion_group!(turner_vs_miranda, bench_fib_turner_vs_miranda);
 criterion_group!(turner_vs_tigre, bench_fib_turner_vs_tigre);
-criterion_main!(turner_vs_miranda, turner_vs_tigre);
+criterion_group!(compilers, bench_compilers_fib, bench_compilers_ack);
+criterion_main!(turner_vs_miranda, turner_vs_tigre, compilers);
 
 fn fib_program(n: u16) -> String {
     format!(
@@ -31,6 +35,27 @@ fn fib(n: u16) -> i32 {
         n as i32
     } else {
         fib(n - 1) + fib(n - 2)
+    }
+}
+
+fn ack_program(m: u16, n: u16) -> String {
+    format!(
+        r#"
+(defun ack (x z) (if (= x 0)
+                     (+ z 1)
+                     (if (= z 0)
+                         (ack (- x 1) 1)
+                         (ack (- x 1) (ack x (- z 1))))))
+(defun main () (ack {m} {n}))
+    "#
+    )
+}
+
+fn ack(m: i32, n: i32) -> i32 {
+    match (m, n) {
+        (0, n) => n + 1,
+        (m, 0) => ack(m - 1, 1),
+        (m, n) => ack(m - 1, ack(m, n - 1)),
     }
 }
 
@@ -57,7 +82,7 @@ fn bench_instance<C: ExprCompiler, E: Engine>(
 
 fn bench_fib_turner_vs_miranda(c: &mut Criterion) {
     let mut group = c.benchmark_group("Fibonacci (Turner vs. Miranda)");
-    for n in (10..=20).step_by(2) {
+    for n in (10..=20).step_by(5) {
         let expected_res = fib(n);
         // Turner
[…]

### 2023-01-23 experiments/superg/build.rs

@@ -112,10 +112,10 @@ fn generate_tigre_combinators(mut f: File) -> std::io::Result<()> {
                 i,
                 // Number of values on the stack (f, g and xs)
                 i + 2,
-                // Size of the values on the stack, in bytes
-                (i + 2) * 8,
-                // Size of the values on the stack, plus an extra 8 bytes for alignment to a 16 byte boundary.
+                // Size of the values on the stack plus the engine pointer, in bytes
                 (i + 3) * 8,
+                // Size of the values on the stack, plus the engine pointer and an extra 8 bytes for alignment to a 16 byte boundary.
+                (i + 4) * 8,
                 i
             )?;
         }

### 2023-01-23 experiments/superg/src/tigre/comb/macros.rs

@@ -19,11 +19,12 @@ macro_rules! comb3 {
             concat!(".global ", stringify!($name)),
             concat!(stringify!($name), ":"),
             r#"
-                mov rdi, r15
                 // Load f, g, x pointers as arguments
                 mov rsi, [rsp]
                 mov rdx, [rsp+8]
                 mov rcx, [rsp+16]
+                // Save engine pointer
+                push rdi
 
                 // Align stack to 16 bytes if needed
                 mov rax, rsp
@@ -35,8 +36,10 @@ macro_rules! comb3 {
             concat!(stringify!($name), "_no_align:"),
             concat!("call ", stringify!($helper_func)),
             r#"
-                // Pop arguments
-                add rsp, 24
+                // Restore engine
+                mov rdi, [rsp]
+                // Pop arguments (+engine)
+                add rsp, 32
                 jmp rax
             "#,
             // Alignment needed
@@ -44,8 +47,10 @@ macro_rules! comb3 {
             "sub rsp, 8",
             concat!("call ", stringify!($helper_func)),
             r#"
+                // Restore engine
+                mov rdi, [rsp+8]
                 // Pop arguments
-                add rsp, 32
+                add rsp, 40
                 jmp rax
             "#,
         );
@@ -54,25 +59,18 @@ macro_rules! comb3 {
         }
         #[no_mangle]
         unsafe extern "C" fn $helper_func(
-            reg_engine: *mut TigreEngine,
+            engine: *mut TigreEngine,
             f: *const CellPtr,
             g: *const CellPtr,
             x: *const CellPtr,
         ) -> CellPtr {
-            // TODO: Restore unwinding behaviour
-            // Place all code within `TigreEngine::with_current` since that also takes
-            // care of catching unwinds before they escape Rust code.
             let fimpl: fn(
                 &mut TigreEngine,
                 *const CellPtr,
                 *const CellPtr,
                 *const CellPtr,
             ) -> CellPtr = $impl;
-            TigreEngine::with_current(|engine| {
-                assert!(std::ptr::eq(reg_engine, engine as *mut TigreEngine));
-                println!("OK");
-                fimpl(engine, f, g, x)
-            })
+            fimpl(&mut *engine, f, g, x)
         }
     };
 }
@@ -86,8 +84,11 @@ macro_rules! comb_bin_op {
             concat!(stringify!($name), ":"),
             r#"
                 // Load argument pointers as arguments
[…]

### 2023-01-23 experiments/superg/src/tigre/comb/mod.rs

@@ -109,7 +109,7 @@ macros::comb3!(comb_C, make_c, |engine, f, g, x| {
 });
 
 // A pointer to a function that evaluates an argument to a strict operator.
-type ArgFn = unsafe extern "C" fn() -> i64;
+type ArgFn = unsafe extern "C" fn(*mut TigreEngine) -> i64;
 macros::comb_bin_op!(comb_plus, apply_plus, |a, b| a + b);
 macros::comb_bin_op!(comb_min, apply_min, |a, b| a - b);
 macros::comb_bin_op!(comb_eq, apply_eq, |a, b| if a == b { 1 } else { 0 });
@@ -120,7 +120,7 @@ macros::comb3!(comb_cond, apply_cond, |engine, c, t, f| {
     // Evaluate the c cell as a strict argument
     let c = c as *const ArgFn;
     // TODO: avoid loading pointers to both branches in the assembly
-    let c_res = (*c)();
+    let c_res = (*c)(engine as *mut TigreEngine);
     let branch_ptr = match c_res {
         0 => f,
         1 => t,

### 2023-01-23 experiments/superg/src/tigre/mod.rs

@@ -1,7 +1,7 @@
 pub mod comb;
 pub mod jit;
 
-use std::{arch::global_asm, cell::UnsafeCell, collections::HashMap, panic::UnwindSafe};
+use std::{cell::UnsafeCell, collections::HashMap};
 
 use jit::JitMem;
 
@@ -115,6 +115,7 @@ impl Engine for TigreEngine {
             next_free_cell: 0,
             def_lookup: HashMap::new(),
         };
+        engine.init_cells();
         // Reserve a cell for each definition
         for def in &program.defs {
             let cell_ptr = engine.make_cell(Comb::I, Comb::Abort);
@@ -132,41 +133,13 @@ impl Engine for TigreEngine {
 
     fn run(&mut self) -> i32 {
         let main_ptr = self.def_lookup.get("main").expect("No main function");
-        let func: extern "C" fn() -> i64 = unsafe { std::mem::transmute(main_ptr.0) };
+        let func: extern "C" fn(*mut TigreEngine) -> i64 =
+            unsafe { std::mem::transmute(main_ptr.0) };
 
-        // Configure the global engine reference
-        ENGINE.with(|engine_cell| {
-            let engine_cell = UnsafeCell::raw_get(engine_cell as *const UnsafeCell<_>);
-            unsafe { *engine_cell = self as *mut TigreEngine };
-        });
+        let res = func(self as *mut TigreEngine);
 
-        let res = unsafe { run_comb_graph(self as *mut TigreEngine, func) };
-
-        // Deregister global engine reference
-        ENGINE.with(|engine_cell| {
-            let engine_cell = UnsafeCell::raw_get(engine_cell as *const UnsafeCell<_>);
-            unsafe { *engine_cell = std::ptr::null_mut() };
-        });
         res.try_into().expect("result does not fit into i32")
     }
-}
-
-global_asm!(
-    r#"
-    .global run_comb_graph
-    run_comb_graph:
-        // Set rbx to engine reference
-        push r15
-        mov r15, rdi
-        call rsi
-        pop r15
-        ret
-"#
-);
-
-extern "C" {
-    #[allow(improper_ctypes)]
-    fn run_comb_graph(engine: *mut TigreEngine, func: extern "C" fn() -> i64) -> i64;
 }
 
 impl TigreEngine {
@@ -177,10 +150,10 @@ impl TigreEngine {
         let arg = rhs.into_cell_ptr(self);
 
[…]

### 2023-01-23 experiments/superg/src/turner/macros.rs

@@ -19,7 +19,7 @@ macro_rules! run_comb {
                 StepResult::Value(v) => {
                     // Make an indirection node
                     self.set_tag(frame_start, Tag::wanted().set_rhs_int());
-                    self.set_hd(frame_start, Comb::I);
+                    self.set_hd(frame_start, COMB_I);
                     self.set_tl(frame_start, CellPtr(v));
 
                     // If nothing is on the stack, return the node
@@ -33,7 +33,7 @@ macro_rules! run_comb {
                 StepResult::Cell(c) => {
                     // Make an indirection node
                     self.set_tag(frame_start, Tag::wanted());
-                    self.set_hd(frame_start, Comb::I);
+                    self.set_hd(frame_start, COMB_I);
                     self.set_tl(frame_start, c);
 
                     // Make the new cell top of the stack

### 2023-01-23 experiments/superg/src/turner/mod.rs

@@ -19,6 +19,9 @@ const CELLS: usize = 250_000;
 #[derive(Debug, Copy, Clone, PartialEq)]
 pub struct CellPtr(i32);
 
+// The I combinator is used so often that we hardwire a cell pointer to it.
+pub const COMB_I: CellPtr = CellPtr(0);
+
 impl CellPtr {
     pub fn comb(&self, engine: &TurnerEngineDebug) -> Option<Comb> {
         engine.comb_inst.get(self.0 as usize).copied()
@@ -29,7 +32,7 @@ impl CellPtr {
     }
 
     pub fn is_i_comb(&self) -> bool {
-        return self.0 == 0;
+        return *self == COMB_I;
     }
 }
 
@@ -45,6 +48,7 @@ impl IntoCellPtr for CellPtr {
 
 impl IntoCellPtr for Comb {
     fn into_cell_ptr(self, engine: &TurnerEngine) -> CellPtr {
+        debug_assert!(self != Comb::I, "Use hardwired COMB_I instead (faster)");
         if let Some(idx) = engine.comb_map.get(&self) {
             CellPtr(*idx as i32)
         } else {
@@ -152,7 +156,7 @@ impl crate::Engine for TurnerEngine {
         // Reserve a spot for each definition
         engine.add_comb_impl(Comb::Abort);
         for def in &compiled_defs {
-            let cell_ptr = engine.make_cell(Tag::wanted(), Comb::I, Comb::Abort);
+            let cell_ptr = engine.make_cell(Tag::wanted(), COMB_I, Comb::Abort);
             if let Some(_) = engine.def_lookup.insert(def.name.clone(), cell_ptr) {
                 panic!("Duplicate definition of {}", def.name);
             }
@@ -491,6 +495,7 @@ impl TurnerEngine {
 
     fn alloc_compiled_expr(&mut self, expr: CompiledExpr) -> CellPtr {
         match expr {
+            CompiledExpr::Comb(Comb::I) => COMB_I,
             CompiledExpr::Comb(c) => c.into_cell_ptr(self),
             CompiledExpr::Var(s) => self
                 .def_lookup

### 2023-01-23 experiments/superg/doc/paper/paper.tex
Write comparison of TIGRE-style engine.

@@
 \usepackage{listings}
+
+\usepackage{svg}
+
+\ifCLASSOPTIONcompsoc
+    \usepackage[caption=false, font=normalsize, labelfont=sf, textfont=sf]{subfig}
+\else
+\usepackage[caption=false, font=footnotesize]{subfig}
+\fi
 
@@
 
-\subsection{Panicking inside combinators}
-Rust has the concept of \emph{panic}: an exceptional situation that indicates a bug in the program, for example, an index out-of-bounds error.
-If a panic occurs, the program collects a stack trace and terminates with an error code.
-A panic inside of a combinator implementation may cause the program to unwind into the node graph, which is undesirable because it is undefined behaviour\footnote{\url{https://doc.rust-lang.org/nomicon/ffi.html\#ffi-and-unwinding}}.
-To prevent this, all combinator implementations are wrapped in a call to \emph{catch\_unwind}\footnote{\url{https://doc.rust-lang.org/std/panic/fn.catch_unwind.html}}, so we catch panics before they leave Rust code, log them and abort the program.
+% Removed because it was too slow
+%\subsection{Panicking inside combinators}
+%Rust has the concept of \emph{panic}: an exceptional situation that indicates a bug in the program, for example, an index out-of-bounds error.
+%If a panic occurs, the program collects a stack trace and terminates with an error code.
+%A panic inside of a combinator implementation may cause the program to unwind into the node graph, which is undesirable because it is undefined behaviour\footnote{\url{https://doc.rust-lang.org/nomicon/ffi.html\#ffi-and-unwinding}}.
+%To prevent this, all combinator implementations are wrapped in a call to \emph{catch\_unwind}\footnote{\url{https://doc.rust-lang.org/std/panic/fn.catch_unwind.html}}, so we catch panics before they leave Rust code, log them and abort the program.
 
@@
 \label{sec:eval}
+Our performance evaluation consists of three parts:
+\begin{itemize}
+    \item We compare our Miranda-style engine with the original Miranda, to demonstrate that our implementation is competitive.
+    \item We compare our TIGRE-style engine to our Miranda-style engine, to see if the techniques described in~\cite{koopman_fresh_1989} are viable on the X86 architecture.
+    \item We compare the traditional bracket abstraction and the different semantic translation algorithms from~\cite{kiselyov_lambda_2018} by compiling the same source programs with each translation algorithm and measuring the runtime on our Miranda-style engine, to see if the semantic translation algorithms produce compiled programs that are faster to execute.
+\end{itemize}
+
+\subsection{Comparison with original Miranda}
+Figure~\ref{fig:turner-vs-miranda} shows the performance of our Miranda-style engine with the traditional bracket abstraction compared to the original Miranda~\cite{noauthor_open_2021}.
+The test program used for the benchmark is the function given in figure~\ref{fig:fib}.
+
+\begin{figure*}
+    \includesvg[width=\textwidth]{turner_vs_miranda}
+    \centering
+    \caption{Time to execute the Fibonacci benchmark program. A lower value on the Y-axis is better.}
+    \label{fig:turner-vs-miranda}
+\end{figure*}
+
+\begin{figure}
+    \begin{lstlisting}
+-- Miranda
+fib n = n,                       if n < 2
+      = fib(n - 1) + fib(n - 2), otherwise
+
+-- Superg (LISP-like)
+(defun fib (n)
+    (if (< n 2) 
+        n
+        (+ (fib (- n 1)) (fib (- n 2)))))
+    \end{lstlisting}
+    \centering
+    \caption{
+        Recursive function to compute the $n$-th element of the Fibonacci sequence.
+    }
+    \label{fig:fib}
+\end{figure}
+
+We observe that our Miranda-style engine is more than twice as fast as the original Miranda.
+While we have not attempted an in-depth evaluation of the performance differences between these two systems, we can point to two likely reasons that our implementation performs better on this benchmark:
+\begin{itemize}
+    \item Miranda supports (exclusively) arbitrary precision integers.
+          They are implemented by the high-performance GNU MP~\cite{gnump} library, but are still slower than the native 32-bit machine integers that our engine uses.
+    \item As mentioned in~\ref{sec:miranda}, the original Miranda must be compiled with optimizations disabled.
+          Without these optimizations, the compiler is likely to generate a slower implementation of key functions.
+\end{itemize}
+
+As the points above indicate, the comparison of the two systems is not fair.
+Therefore, we do not claim that our implementation is strictly better than the original, but rather that it is a reasonably fast reference implementation of a combinator graph reduction engine.
+The other benchmarks in this section all use the Miranda-style engine with bracket abstraction as a reference point for performance, based on which we can judge if the proposed changes are beneficial.
+
+\subsection{TIGRE-style Engine}
+To benchmark the TIGRE-style engine we use the same Fibonacci program and bracket abstraction algorithm, and plot the execution time against our Miranda-style engine.
+Both engines are configured to use the same compiler, so they will execute identical combinator expressions.
+The results can be seen in figure~\ref{fig:turner-vs-tigre}.
+
+\begin{figure*}
+    \includesvg[width=\textwidth]{turner_vs_tigre}
+    \centering
+    \caption{Time to execute the Fibonacci benchmark program. A lower value on the Y-axis is better.}
+    \label{fig:turner-vs-tigre}
+\end{figure*}
+
+Contrary to our expectations, the TIGRE-style engine is significantly slower than the reference Miranda-style engine.
+Based on the flame graphs~\cite{gregg_flame_2016} of the two engines, available in Appendix~\ref{app:flame-graph}, we can conclude that the S combinator is significantly slower on this engine.
+
+We used Valgrind's cachegrind~\cite{NethercotePhd2004} tool to inspect the \texttt{make\_s} function and found that its execution time is dominated by data write operation to update the graph.
+While the TIGRE-style engine executes fewer instructions (22M vs. 79M) and performs fewer data accesses (13M vs. 49M), it has a much higher cache miss rate for writes (3\% vs. 0.5\%).
+
+We hypothesize that updates to nodes in the graph are slow in the TIGRE-style engine because those same nodes have been loaded into the instruction cache, and therefore trigger an expensive cache invalidation to maintain coherency between the instruction and data caches.
+The node to be updated is part of the current spine, so its \texttt{CALL} instruction will have been executed before the update, loading it into the instruction cache.
+When a write occurs to this node, the CPU must conservatively invalidate or update that cached value, or it would risk executing a stale instruction if the code were to jump back to that address.
+Writes to addresses in the instruction cache are frequent in the TIGRE-style engine, but in most software, they are very rare.
+On some processor architectures, like ARM, programs must manually clear the instruction cache if they issue a write to a location that may be cached~\cite{bramley_caches_2013}.
+
+It appears that the TIGRE approach to implementing combinator graph reduction is a poor fit for modern architectures due to the split between instruction and data caches.
+Other innovations from TIGRE, such as \texttt{LIT} combinators and removing tags, may still be beneficial.
+In particular, it may be possible to get the benefits of TIGRE without cache coherency issues by storing jump addresses rather than call instructions in nodes.
+Rather than jump directly to a node and execute it, the runtime could load the jump address and push the argument pointer to the stack normally, then jump to the address instead.
+The implementation of such an engine is left as future work.
+
+\subsection{Kiselyov's algorithms}
+
+\textbf{TEXT BELOW ARE NOTES!}
 Programs:
@@
 
+\appendix
+
+\subsection{Flame graphs}
+\label{app:flame-graph}
+\begin{figure*}
+    \includesvg[inkscapeformat=png, width=\textwidth]{flamegraph_turner}
+    \centering
+    \caption{Flame graph for $fib(20)$ on Miranda-style engine.}
+    \label{fig:flamegraph-turner}
+\end{figure*}
+
+\begin{figure*}
+    \includesvg[inkscapeformat=png, width=\textwidth]{flamegraph_tigre}
+    \centering
+    \caption{Flame graph for $fib(20)$ on TIGRE-style engine.}
+    \label{fig:flamegraph-tigre}
+\end{figure*}
+
 \end{document}

### 2023-01-23 experiments/superg/doc/paper/paper.tex
Write Kiselyov's algorithms comparison.

@@
 \fi
+
+\usepackage{pgfplots}
 
@@
     \centering
-    \caption{Time to execute the Fibonacci benchmark program. A lower value on the Y-axis is better.}
+    \caption{
+        Time to execute the Fibonacci benchmark program on the original Miranda and our Miranda-style engine.
+        A lower value on the Y-axis is better.}
     \label{fig:turner-vs-miranda}
@@
     \centering
-    \caption{Time to execute the Fibonacci benchmark program. A lower value on the Y-axis is better.}
+    \caption{
+        Time to execute the Fibonacci benchmark program on Miranda-style and TIGRE-style engines.
+        A lower value on the Y-axis is better.}
     \label{fig:turner-vs-tigre}
@@
 \subsection{Kiselyov's algorithms}
+To compare the semantic translation algorithms from~\cite{kiselyov_lambda_2018} with each other, we again use the Fibonacci test program and execute all compiled programs on the Miranda-style engine.
+The bracket abstraction algorithm has been added as a reference point.
+The results can be seen in figure~\ref{fig:bracket-vs-kiselyov}.
 
-\textbf{TEXT BELOW ARE NOTES!}
-Programs:
-\begin{itemize}
-    \item ackermann
-    \item digits of e
-    \item linfib
-    \item lsort
-    \item nfib
-    \item primes
-    \item queens
-    \item tak
-    \item towers of hanoi
-    \item treesort
-\end{itemize}
+\begin{figure*}
+    \includesvg[width=\textwidth]{bracket_vs_kiselyov}
+    \centering
+    \caption{Time to execute the Fibonacci benchmark program compiled with the algorithms from~\cite{kiselyov_lambda_2018}. A lower value on the Y-axis is better.}
+    \label{fig:bracket-vs-kiselyov}
+\end{figure*}
 
-Evaluate performance of:
+All of the algorithms presented in~\cite{kiselyov_lambda_2018} produce more efficient code than the reference bracket abstraction algorithm.
+The $\lazyeta$ compilation scheme produces the best code for the Fibonacci program.
+For many of the test programs in~\cite{kiselyov_lambda_2018} it was shown to generate the shortest programs of all the algorithms presented there.
+This is also true for the Fibonacci program, as shown in figure~\ref{fig:fib-sizes}.
 
-\begin{itemize}
-    \item Bracket abstraction
-    \item Kiselyov
-    \item GHC
-\end{itemize}
+\begin{figure}
+    \begin{tikzpicture}
+        \begin{axis}[
+                xlabel={Compiled expression size (nodes)},
+                xmin=0,
+                ylabel={Translation algorithm},
+                symbolic y coords={Bracket,$\strict$,$\lazy$,$\lazyeta$,$\linear$},
+            ]
+            \addplot[xbar] coordinates {
+                    (73,Bracket)
+                    (59,$\strict$)
+                    (59,$\lazy$)
+                    (47,$\lazyeta$)
+                    (53,$\linear$)
+                };
+        \end{axis}
+    \end{tikzpicture}
+    \centering
+    \caption{Size of combinator expression generated for the Fibonacci program by different translation algorithm}
+    \label{fig:fib-sizes}
+\end{figure}
 
-Suggestion from \textit{Common Benchmarking Pitfalls}: Always look at your distribution.
-When you have run benchmarks, take a look at the times for each sample.
+It comes as no surprise that the shortest translation is also the fastest to execution, but it is worth noting that while the Bracket abstraction translation is only 55\% larger than the $\lazyeta$ one, it runs 94\% slower.
+We see a similar pattern for the Ackermann function, for which results may be found in appendix~\ref{app:ack}.
+This suggests that while the semantic translation algorithms from~\cite{kiselyov_lambda_2018} reduce the size of the combinator expression, they reduce runtime even further.
+
+\section{Related work}
+supercombinators
+
+GHC
+
+Kiselyov's later work.
+
+Edward Kmett's SIMD combinator stuff.
 
@@
 
+\newpage
+
+\subsection{Ackermann benchmark results}
+\label{app:ack}
+
+\begin{figure*}
+    \begin{lstlisting}
+(defun ack (x z) (if (= x 0)
+                     (+ z 1)
+                     (if (= z 0)
+                         (ack (- x 1) 1)
+                         (ack (- x 1) (ack x (- z 1))))))
+    \end{lstlisting}
+    \centering
+    \caption{
+        Ackermann's function.
+    }
+    \label{fig:ack}
+\end{figure*}
+
+\begin{figure*}
+    \includesvg[width=\textwidth]{ack_bracket_vs_kiselyov}
+    \centering
+    \caption{Time to execute the Ackermann benchmark program compiled with the algorithms from~\cite{kiselyov_lambda_2018}. A lower value on the Y-axis is better.}
+    \label{fig:ack-bracket-vs-kiselyov}
+\end{figure*}
+
+\begin{figure}
+    \begin{tikzpicture}
+        \begin{axis}[
+                xlabel={Compiled expression size (nodes)},
+                xmin=0,
+                ylabel={Translation algorithm},
+                symbolic y coords={Bracket,$\strict$,$\lazy$,$\lazyeta$,$\linear$},
+            ]
+            \addplot[xbar] coordinates {
+                    (415,Bracket)
+                    (175,$\strict$)
+                    (137,$\lazy$)
+                    (109,$\lazyeta$)
+                    (123,$\linear$)
+                };
+        \end{axis}
+    \end{tikzpicture}
+    \centering
+    \caption{Size of combinator expression generated for the Ackermann program by different translation algorithm}
+    \label{fig:ack-sizes}
+\end{figure}
+
 \end{document}

### 2023-01-24 experiments/superg/doc/paper/paper.tex
Write conclusion.

@@
 
+While the $\linear$ strategy produces slightly more compact code than $\strict$ and $\lazy$, there is no clear performance improvement.
+Our small benchmark programs only require bulk combinators for up to $2$ arguments, so the $\linear$ strategy cannot win much in terms of code size.
+We expect that for certain worst-case inputs, as given in~\cite{kiselyov_lambda_2018}, $\linear$ may still prove useful.
+
 It comes as no surprise that the shortest translation is also the fastest to execution, but it is worth noting that while the Bracket abstraction translation is only 55\% larger than the $\lazyeta$ one, it runs 94\% slower.
@@
 % Repeat the main question / objective of your research
+Our objective with this work was to revive research around Combinator Graph Reduction by improving the performance.
+We have attempted this by using a novel semantic translation algorithm to compile lambda calculus programs into combinator expressions~\cite{kiselyov_lambda_2018}, and by implementing combinator graph reduction engines based on the designs of Miranda~\cite{turner_new_1979} and TIGRE~\cite{koopman_fresh_1989}.
 
 % Give the answer to the question
-Source code for our Superg evaluator is available at \url{https://github.com/wildarch/mono/tree/main/experiments/superg}.
+We have developed a reference implementation of a combinator graph reduction engine based on the now open source Miranda~\cite{turner_new_1979}.
+% Why is this the answer? Present the main arguments
+Our implementation comprises under 1000 lines of Rust code and outperforms the original Miranda engine on our benchmarks.
 
+% Give the answer to the question
+Our port of the TIGRE~\cite{koopman_fresh_1989} engine runs natively on stock x86-64 hardware and executes our benchmark program with fewer instructions and data accesses than our Miranda-style engine.
 % Why is this the answer? Present the main arguments
+Unfortunately, its performance still lags behind the reference implementation because its inherent frequent writes to executable code sections trigger expensive invalidations of the instruction cache.
+We conclude that the TIGRE design is a poor fit for architectures with split instruction and data caches, though some concepts like \texttt{LIT} combinators and tag elimination may still prove effective.
 
-% What are the implications of this answer?
+% Give the answer to the question
+Applying Kiselyov's translation algorithms~\cite{kiselyov_lambda_2018} to our benchmarks was a resounding success.
+% Why is this the answer? Present the main arguments
+All proposed variations of the semantic translation produced a combinator expression that was smaller than the one generated by the reference bracket abstraction algorithm.
+Moreover, the effect of the smaller code size was amplified at execution time, with the smallest program running nearly twice as fast as the reference program.
 
 % Suggestions for future research
+There is ample room for improvement both on the compilation and the runtime side.
+Our $\linear$ translation, like Kiselyov's, is based on the $\strict$ translation rather than the $\lazy$ or $\lazyeta$ translation schemes.
+Combining bulk combinators with lazy weakening could further reduce the size of compiled programs, and therefore also improve performance.
+
+On the runtime side, ideas from TIGRE could be selectively integrated into the Miranda-style engine to improve its performance, such as \texttt{LIT} combinators or eliminating tags.
+
+Our current runtime is single-threaded: it always reduces one combinator at a time.
+However, reducing multiple combinators in parallel still leads to a correct execution, so it may be possible to run programs faster by evaluating separate parts of the graph in parallel on different cores.
+As suggested by Kmett~\cite{kmett_combinators_2018}, it may even be possible to use SIMD instructions to reduce multiple combinators in parallel on a \emph{single} core.
+
+Our garbage collector and the one used by Miranda is a simple mark-and-sweep collector.
+While our benchmark programs do not run long enough to exhaust the available memory, we expect that for real programs a faster garbage collector could improve performance significantly.
+In particular, a concurrent garbage collector~\cite{jones_garbage_2011} could prevent execution stalls during collection time.
+We hypothesize that a limited form of reference counting would also be beneficial: the S combinator (B and C are very similar) currently requires the allocation of two new nodes when it executes, and it also drops a reference to two other nodes.
+With reference counting we could determine if those two nodes are not needed elsewhere in the graph, and reuse them, eliminating the creation of garbage.
+
+Source code for our translation algorithms and runtimes, as well as associated tests and benchmarks, are available at \url{https://github.com/wildarch/mono/tree/main/experiments/superg}.
 

### 2023-01-24 experiments/superg/benches/bench.rs
Remove unused argument in benchmarks.

@@ -144,46 +144,30 @@ fn bench_compilers_ack(c: &mut Criterion) {
     }
 }
 
-fn add_all_compilers<P: Display + Clone>(
+fn add_all_compilers<P: Display + Clone + Copy>(
     group: &mut BenchmarkGroup<WallTime>,
     n: P,
     program: &str,
     expected_res: i32,
 ) {
     // Bracket
-    group.bench_with_input(BenchmarkId::new("Bracket", n.clone()), &n, |b, n| {
+    group.bench_function(BenchmarkId::new("Bracket", n), |b| {
         bench_instance::<_, TurnerEngine>(b, program, expected_res, BracketCompiler);
     });
     // Strict
-    group.bench_with_input(
-        BenchmarkId::new("Kiselyov - Strict", n.clone()),
-        &n,
-        |b, n| {
-            bench_instance::<_, TurnerEngine>(b, program, expected_res, StrictCompiler);
-        },
-    );
+    group.bench_function(BenchmarkId::new("Kiselyov - Strict", n), |b| {
+        bench_instance::<_, TurnerEngine>(b, program, expected_res, StrictCompiler);
+    });
     // Lazy
-    group.bench_with_input(
-        BenchmarkId::new("Kiselyov - Lazy", n.clone()),
-        &n,
-        |b, n| {
-            bench_instance::<_, TurnerEngine>(b, program, expected_res, LazyCompiler);
-        },
-    );
+    group.bench_function(BenchmarkId::new("Kiselyov - Lazy", n), |b| {
+        bench_instance::<_, TurnerEngine>(b, program, expected_res, LazyCompiler);
+    });
     // LazyOpt
-    group.bench_with_input(
-        BenchmarkId::new("Kiselyov - LazyOpt", n.clone()),
-        &n,
-        |b, n| {
-            bench_instance::<_, TurnerEngine>(b, program, expected_res, LazyOptCompiler);
-        },
-    );
+    group.bench_function(BenchmarkId::new("Kiselyov - LazyOpt", n), |b| {
+        bench_instance::<_, TurnerEngine>(b, program, expected_res, LazyOptCompiler);
+    });
     // Linear
-    group.bench_with_input(
-        BenchmarkId::new("Kiselyov - Linear", n.clone()),
-        &n,
-        |b, n| {
-            bench_instance::<_, TurnerEngine>(b, program, expected_res, LinearCompiler);
-        },
-    );
[…]

### 2023-01-24 experiments/phd/papers.py
Make list of papers.

@@ -1,0 +1,68 @@
+#!/usr/bin/env python3
+import os
+from lxml import etree
+from dataclasses import dataclass
+
+PEOPLE_DIR='people/'
+
+@dataclass
+class Paper:
+    key: str
+    title: str
+    authors: list[str]
+    year: int
+    ee: str
+
+papers = {}
+
+for xml_filename in os.listdir(PEOPLE_DIR):
+    xml_path = PEOPLE_DIR + xml_filename
+    tree = etree.parse(xml_path)
+
+    for r in tree.iterfind('//r/*'):
+        title = etree.tostring(r.find('.//title')).decode('utf-8')
+        title = title.strip().replace('<title>', '').replace('</title>', '')
+        paper = Paper(
+            key = r.get('key'),
+            title = title,
+            authors = [a.text for a in r.iterfind('.//author')],
+            year = int(r.findtext('.//year')),
+            ee = r.findtext('.//ee'),
+        )
+
+        papers[paper.key] = paper
+
+print(len(papers))
+
+papers = list(papers.values())
+papers.sort(key=lambda p: p.year, reverse=True)
+
+with open('papers.html', 'w') as f:
+    f.write('''
+<!DOCTYPE html>
+<html lang="en">
+<meta charset="UTF-8">
+<title>Papers</title>
+<meta name="viewport" content="width=device-width,initial-scale=1">
+<body>
+<h1>Papers</h1>
+
+    ''')
+
+    year = papers[0].year
+    f.write(f'<h2>{year}</h2>')
+    for paper in papers:
+        if paper.year != year:
+            year = paper.year
+            f.write(f'<h2>{year}</h2>')
+        
+        f.write('<div>')
+        f.write(f'<a href="{paper.ee}">{paper.title}</a> {", ".join(paper.authors)}')
+        f.write('<br />')
+        f.write('<br />')
+        f.write('</div>')
+
+    f.write('''
+</body>
+</html>
+    ''')

### 2023-01-24 experiments/phd/papers.py
Mark important authors.

@@ -13,11 +13,15 @@ class Paper:
     year: int
     ee: str
 
+author_names = set()
 papers = {}
 
 for xml_filename in os.listdir(PEOPLE_DIR):
     xml_path = PEOPLE_DIR + xml_filename
     tree = etree.parse(xml_path)
+
+    for name in tree.iterfind('./person/author'):
+        author_names.add(name.text)
 
     for r in tree.iterfind('//r/*'):
         title = etree.tostring(r.find('.//title')).decode('utf-8')
@@ -56,8 +60,10 @@ with open('papers.html', 'w') as f:
             year = paper.year
             f.write(f'<h2>{year}</h2>')
         
+        authors = [f'<b>{n}</b>' if n in author_names else n for n in paper.authors]
+        
         f.write('<div>')
-        f.write(f'<a href="{paper.ee}">{paper.title}</a> {", ".join(paper.authors)}')
+        f.write(f'<a href="{paper.ee}">{paper.title}</a> {", ".join(authors)}')
         f.write('<br />')
         f.write('<br />')
         f.write('</div>')

### 2023-01-24 experiments/superg/doc/paper/paper.tex
Write abstract.

@@
     urlcolor=cyan,
-    pdftitle={Reviving the Lost Art of Combinator Graph Reduction},
-    pdfpagemode=FullScreen,
+    pdftitle={Reviving the Lost Art of Combinator Graph Reduction}
     }
@@
 \usepackage{pgfplots}
+\pgfplotsset{compat=1.17}
 
@@
     % - 1 sentence background
+    Combinator Graph Reduction was once the dominant execution strategy for lazy functional languages, but inefficient translation algorithms and lack of an open-source implementation caused the technique to fall into obscurity in the late 1980s.
     % - 1 sentence motive/problem
+    Recently, a new family of translation algorithms has been proposed~\cite{kiselyov_lambda_2018}.
+    The new translations produce smaller combinator expressions, and for some their output size is linear in the size of the input in the worst case, rather than quadratic.
+    While the effect on code size is known, the difference in runtime performance has remained unquantified.
     % - 1 sentence objective
+    Our goal is to find out if the gains in code size hold up during execution time on a modern, efficient Combinator Graph Reduction engine.
     % - 1 sentence approach/method
+    Since no such engine existed at the time, we have developed our own based on the design of Miranda~\cite{turner_miranda_1985} and TIGRE~\cite{koopman_fresh_1989}.
+    We compile a set of test programs with the proposed translation algorithms~\cite{kiselyov_lambda_2018} and measure their execution time on our best engine.
     % - 2 sentences support
+    Our benchmarks show that our Miranda-style engine outperforms the original Miranda, and is therefore a good reference point to measure performance improvements against.
+    We find that the threaded code approach proposed in TIGRE reduces the number of instructions and data accesses needed to execute programs, but frequent cache misses make it unattractive on architectures with split instruction and data caches.
+    Kiselyov's new translation algorithms are shown to significantly boost performance in our benchmarks, the effect on execution time is even greater than we observe on code size.
     % - 1 sentence conclusion
+    We conclude that the new translation algorithms significantly improve performance, and there may be further gains to be had by combining ideas from the different translation algorithms.
+    Many opportunities remain to improve the basic combinator graph reduction engine, such as the integration of other ideas from TIGRE, a concurrent garbage collector, and parallel reduction.
     % - 1 sentence implication
-    % - 1 sentence limitation
-    This is the abstract.
+    Our results show that there are significant improvements to be made in this area of research that has been largely static in recent years, and we hope there will be renewed interest from the community to do research in this field.
 \end{abstract}
@@
 Eventually, the interest in Miranda and combinator graph reduction with it waned.
-Part of the reason may be that Miranda was until very recently a closed-source, licensed software (its source code was released only in 2020~\cite{noauthor_open_2021}), leading to the creation of the open-source Haskell language~\cite{hudak_history_2007}.
+Part of the reason may be that Miranda was until very recently closed-source, licensed software (its source code was released only in 2020~\cite{noauthor_open_2021}), leading to the creation of the open-source Haskell language~\cite{hudak_history_2007}.
 But combinator reduction also deserves part of the blame.

### 2023-01-24 experiments/superg/doc/paper/paper.tex
Related work.

@@
 \section{Related work}
-supercombinators
-
-GHC
-
-Kiselyov's later work.
-
-Edward Kmett's SIMD combinator stuff.
+\subsection{Super-combinators}
+Super-combinators~\cite{hughes_super-combinators_1982} are the basis of many lazy functional runtimes~\cite{fairbairn_tim_1987,kieburtz_g-machine_1985,burn_spineless_1988,jones_implementing_nodate} developed after Miranda.
+A supercombinator is a lambda expression of the form $\lambda x_1.
+\lambda x_2.
+...
+\lambda x_n.
+E$, where $E$ has no free variables, and any lambda abstraction in $E$ is also a supercombinator~\cite{spj_impl}.
+Any program in the lambda calculus may be transformed into supercombinators by \emph{lifting} free variables.
+For example, we may consider $\lambda f.f\ (\lambda x.f\ x\ 2)$, which is not a super-combinator, because $f$ is free in the inner lambda.
+To make this expression a super-combinator, we can pass $f$ explicitly:
+
+\[
+    \lambda f. f\ ((\lambda f.\lambda x. f\ x\ 2)\ f)
+\]
+
+This transformation is useful because the output of a supercombinator depends entirely on the value of its inputs, and thus its implementation may be described as a fixed set of (machine) instructions.
+
+Programs translated into super-combinators have a code size that is linear in the size of the input in the worst case, an improvement over the worst-case quadratic size of bracket abstraction.
+
+Perhaps more importantly, the reduction steps for a program based on super-combinators are larger.
+A downside is that this makes the execution less lazy, but an important advantage is that the larger step size makes the overhead of the reduction loop smaller.
+The small step size is an often noted issue with combinator reduction engines~\cite{spj_impl}, and a major reason why they were largely abandoned.
+
+\subsection{Haskell}
+Haskell is without a doubt the most popular lazy functional programming language today.
+It too has its roots in the super-combinator approach.
+The popular Glasgow Haskell Compiler is based on the Spineless Tagless G-Machine~\cite{jones_implementing_nodate}, which in turn is based on the Spineless G-Machine~\cite{burn_spineless_1988} and G-Machine~\cite{kieburtz_g-machine_1985}.
+GHC compiles super-combinators into machine code ahead of time, which makes the resulting binaries self-contained and highly efficient.
+In future work, we will compare the performance of our runtime against that of GHC.
+
+\subsection{SIMD Combinator Reduction}
+In his talk \emph{Combinator Revisited}~\cite{kmett_combinators_2018}, Kmett hypothesizes the use of SIMD instructions to perform combinator reductions.
+The rationale is that combinators have a relatively simple implementation, and SIMD instructions are getting more and more powerful, so potentially they can be used to implement the reduction of multiple combinators in a single instructions.
+Assuming the reduction of a single combinator uses 64-bit wide values, the same routine implemented with AVX-512 instructions could be used to perform as many as 8 reductions in parallel on a single core.
+An eightfold improvement in performance is certainly promising, although it is unclear what level of parallelism a typical program may allow for.
+A work-in-progress version of the compiler is available at \url{https://github.com/ekmett/coda}.
 

### 2023-01-24 experiments/superg/doc/paper/paper.tex
Stub out sections of preliminaries.

@@
 \textbf{TODO: write section. Currently contains only a few small fragments}
+\subsection{Lambda Calculus}
+
 \subsection{Combinatory Logic}
@@
 
-\subsection{Graph Reduction}
-% In the reduction graph, vertices are combinators, constants or function applications, and edges connect the function and argument for application (combinators and constants have no outgoing edges, they are leaf nodes).
+\subsection{Bracket abstraction}
 
@@
     \item Miranda supports (exclusively) arbitrary precision integers.
-          They are implemented by the high-performance GNU MP~\cite{gnump} library, but are still slower than the native 32-bit machine integers that our engine uses.
+          They are implemented by the high-performance GNU MP~\cite{gnump} library but are still slower than the native 32-bit machine integers that our engine uses.
     \item As mentioned in~\ref{sec:miranda}, the original Miranda must be compiled with optimizations disabled.
@@
 In his talk \emph{Combinator Revisited}~\cite{kmett_combinators_2018}, Kmett hypothesizes the use of SIMD instructions to perform combinator reductions.
-The rationale is that combinators have a relatively simple implementation, and SIMD instructions are getting more and more powerful, so potentially they can be used to implement the reduction of multiple combinators in a single instructions.
+The rationale is that combinators have a relatively simple implementation, and SIMD instructions are getting more and more powerful, so potentially they can be used to implement the reduction of multiple combinators in a single instruction.
 Assuming the reduction of a single combinator uses 64-bit wide values, the same routine implemented with AVX-512 instructions could be used to perform as many as 8 reductions in parallel on a single core.

### 2023-01-25 experiments/superg/doc/paper/paper.tex
Resolve comments up to Miranda section.

@@
 
-\title{Reviving the Lost Art of Combinator Graph Reduction}
+\title{Reviving the Lost Art of \\Combinator Graph Reduction}
 
@@
 
+\section{Submission}
+Submitted in partial fulfillment of the requirements for completion of the TU Eindhoven course Advanced Functional Programming (Capita Selecta SET, 2IMP05), under the guidance of prof. Tom Verhoeff.
+
 \section{Introduction}
@@
 Eventually, the interest in Miranda and combinator graph reduction with it waned.
-Part of the reason may be that Miranda was until very recently closed-source, licensed software (its source code was released only in 2020~\cite{noauthor_open_2021}), leading to the creation of the open-source Haskell language~\cite{hudak_history_2007}.
+Part of the reason may be that Miranda was until very recently closed-source, licensed software (its source code was released only in 2020~\cite{turner_open_2021}), leading to the creation of the open-source Haskell language~\cite{hudak_history_2007}.
 But combinator reduction also deserves part of the blame.
@@
 For a lambda calculus expression of size $N$, the traditional bracket abstraction algorithm produces an output of $O(N^2)$ in the worst case, where other approaches such as supercombinators are linear\cite{spj_impl}.
-Larger programs run slower, and along with other disadvantages, this led to combinator graph reduction being all but abandoned.
+Along with other disadvantages, this led to combinator graph reduction being all but abandoned.
 
@@
 % - Sub-questions
-We use GHC as the reference for a contemporary evaluator, and then measure the execution time of a set of benchmark programs.
-We also need a fast reduction engine that can run programs compiled to combinator expressions.
+We compare Kiselyov's translation algorithm to bracket abstraction by compiling a set of benchmark programs and measuring their execution time.
+Future work will add GHC as a reference point for a contemporary evaluator, along with additional benchmark programs.
+
+We also need a fast Combinator Graph Reduction engine that can run programs compiled to combinator expressions.
 Alas, there is no performant reduction engine tailored to executing such programs.
-The now open-sourced Miranda is a good starting point, but it is showing its age.
-The 2020 port compiles cleanly on a recent GCC, but enabling optimizations is known to break the garbage collector.
-Miranda's architecture is based on cells with a tag, head and tail as described in\cite{turner_new_1979}, but later work has shown that a tagless representation with direct pointers to code leads to better performance\cite{koopman_fresh_1989}.
+The now open-sourced Miranda software is a good starting point, but it is showing its age.
+The 2020 port compiles cleanly on a recent GCC, but enabling optimizations is known to break the garbage collector (section~\ref{sec:miranda} covers this issue in more detail).
+Miranda's architecture is based on cells with a tag, head, and tail as described in\cite{turner_new_1979}, but later work has shown that a tagless representation with direct pointers to code leads to better performance\cite{koopman_fresh_1989}.
 In absence of an up-to-date reference on building a practical combinator reduction engine, we formulate the second research question:
 
-\textbf{RQ2:} How is combinator graph reduction implemented efficiently on conventional hardware?
+\textbf{RQ2:} How is combinator graph reduction implemented efficiently on general-purpose hardware?
 
@@
 \begin{itemize}
-    \item Kiselyov develops not one but two algorithms. The first does not have linear complexity, but tends to produce very compact code.
+    \item Kiselyov develops not one but two algorithms. The first does not have linear complexity but tends to produce very compact code.
           The second is guaranteed to be of linear complexity, at the cost of being less compact in certain cases.
-          Kiselyov's description is somewhat theoretical, so in section~\ref{sec:kiselyov} we describe both algorithms in a way that is straightforward to implement.
+          Kiselyov's description requires that the reader be familiar with OCaml and the tagless-final style of programming.
+          In section~\ref{sec:kiselyov} we describe both algorithms in a more conventional style that is directly implementable in a general-purpose language.
     \item We implement a fast combinator graph reduction engine based on the ideas in Miranda~\cite{turner_miranda_1985} and later improvements~\cite{koopman_architecture_1992}.
           We elaborate on its design in sections~\ref{sec:miranda} and~\ref{sec:tigre}.
-    \item We evaluate the performance of Miranda, Haskell and our engine on a set of benchmark programs in section~\ref{sec:eval}.
+    \item We evaluate the performance of Miranda, our engines, and Kiselyov's translation algorithms on a set of benchmark programs in section~\ref{sec:eval}.
 \end{itemize}
 
-For readers unfamiliar with combinatory logic and/or graph reduction, we provide a short introduction to these topics in section~\ref{sec:prelim}.
+For readers unfamiliar with lambda calculus, combinatory logic or bracket abstraction, we provide a short introduction to these topics in section~\ref{sec:prelim}.
 
@@
 Kiselyov's paper~\cite{kiselyov_lambda_2018} describes three iterations of his non-linear semantic translation, and finally the linear algorithm.
-They are presented as a type system accompanied with operational semantics, with the exception of the $\eta$-optimization given as an OCaml code patch.
-OCaml's type system can represent these algorithms very concisely in tagless-final style, but how should we represent them in a more conventional style, possibly in a language that does not have such a strong type system?
+They are presented as a type system accompanied by an operational semantics, except for the $\eta$-optimization, which is given as an OCaml code patch.
+OCaml's type system can represent these algorithms very concisely in a tagless-final style.
+Below we describe the same algorithms in a more conventional style.
+Our description places no requirements on the type system and can be directly implemented in any general-purpose programming language.
 
@@
 
-In our implementation, we unroll the recursion into a while loop, and repeatedly pop elements from the two arrays until both are empty.
-This builds the new context backwards, so as a final step the array is reversed.
+In our implementation, we unroll the recursion into a while loop and repeatedly pop elements from the two arrays until both are empty.
+This builds the new context back to front, so as a final step, the array is reversed.
 
@@
 While there is published work on Miranda~\cite{turner_miranda_1985}, it does not provide much detail on the reduction engine.
-Fortunately, the source code for Miranda is now freely available~\cite{noauthor_open_2021}, and from the code, we can deduce that its design is very similar to SASL~\cite{turner_new_1979}, an earlier language developed by Turner.
+Fortunately, the source code for Miranda is now freely available~\cite{turner_open_2021}, and from the code, we can deduce that its design is very similar to SASL~\cite{turner_new_1979}, an earlier language developed by Turner.
 

### 2023-01-25 experiments/superg/doc/paper/paper.tex
Capitalize 'section'.

@@
 The now open-sourced Miranda software is a good starting point, but it is showing its age.
-The 2020 port compiles cleanly on a recent GCC, but enabling optimizations is known to break the garbage collector (section~\ref{sec:miranda} covers this issue in more detail).
+The 2020 port compiles cleanly on a recent GCC, but enabling optimizations is known to break the garbage collector (Section~\ref{sec:miranda} covers this issue in more detail).
 Miranda's architecture is based on cells with a tag, head, and tail as described in\cite{turner_new_1979}, but later work has shown that a tagless representation with direct pointers to code leads to better performance\cite{koopman_fresh_1989}.
@@
           Kiselyov's description requires that the reader be familiar with OCaml and the tagless-final style of programming.
-          In section~\ref{sec:kiselyov} we describe both algorithms in a more conventional style that is directly implementable in a general-purpose language.
+          In Section~\ref{sec:kiselyov} we describe both algorithms in a more conventional style that is directly implementable in a general-purpose language.
     \item We implement a fast combinator graph reduction engine based on the ideas in Miranda~\cite{turner_miranda_1985} and later improvements~\cite{koopman_architecture_1992}.
-          We elaborate on its design in sections~\ref{sec:miranda} and~\ref{sec:tigre}.
-    \item We evaluate the performance of Miranda, our engines, and Kiselyov's translation algorithms on a set of benchmark programs in section~\ref{sec:eval}.
+          We elaborate on its design in Sections~\ref{sec:miranda} and~\ref{sec:tigre}.
+    \item We evaluate the performance of Miranda, our engines, and Kiselyov's translation algorithms on a set of benchmark programs in Section~\ref{sec:eval}.
 \end{itemize}
 
-For readers unfamiliar with lambda calculus, combinatory logic or bracket abstraction, we provide a short introduction to these topics in section~\ref{sec:prelim}.
+For readers unfamiliar with lambda calculus, combinatory logic or bracket abstraction, we provide a short introduction to these topics in Section~\ref{sec:prelim}.
 
@@
 Cells without this bit set are considered free and may be overwritten to store new nodes in the graph.
-More on this in the section on Garbage collection.
+More on this in Section~\ref{sec:gc}.
     \item \texttt{RHS\_INT}: Set if the right pointer is not a pointer but an integer value.
@@
 \subsection{Garbage collection}
+\label{sec:gc}
 The graph reduction process may lead to some nodes becoming unreachable.

### 2023-01-25 experiments/superg/doc/paper/paper.tex
Comments by Tom.

@@
     Combinator Graph Reduction was once the dominant execution strategy for lazy functional languages, but inefficient translation algorithms and lack of an open-source implementation caused the technique to fall into obscurity in the late 1980s.
+    The traditional translation algorithm produces a combinator expression that is in the worse case quadratic in the size of the input.
     % - 1 sentence motive/problem
-    Recently, a new family of translation algorithms has been proposed~\cite{kiselyov_lambda_2018}.
-    The new translations produce smaller combinator expressions, and for some their output size is linear in the size of the input in the worst case, rather than quadratic.
+    Recently, Kiselyov proposed a new family of translation algorithms that generate smaller combinator expressions, with some variants being linear in the size of the input.
     While the effect on code size is known, the difference in runtime performance has remained unquantified.
     % - 1 sentence objective
-    Our goal is to find out if the gains in code size hold up during execution time on a modern, efficient Combinator Graph Reduction engine.
+    Our goal is to find out how the reduction in code size affects execution time on a modern, efficient Combinator Graph Reduction engine.
     % - 1 sentence approach/method
-    Since no such engine existed at the time, we have developed our own based on the design of Miranda~\cite{turner_miranda_1985} and TIGRE~\cite{koopman_fresh_1989}.
-    We compile a set of test programs with the proposed translation algorithms~\cite{kiselyov_lambda_2018} and measure their execution time on our best engine.
+    In absence of a suitable engine, we have developed our own based on the design of Miranda and TIGRE.
+    We compile a set of test programs with Kiselyov's translation algorithms and measure their execution time on our best engine.
     % - 2 sentences support
+
     Our benchmarks show that our Miranda-style engine outperforms the original Miranda, and is therefore a good reference point to measure performance improvements against.
     We find that the threaded code approach proposed in TIGRE reduces the number of instructions and data accesses needed to execute programs, but frequent cache misses make it unattractive on architectures with split instruction and data caches.
-    Kiselyov's new translation algorithms are shown to significantly boost performance in our benchmarks, the effect on execution time is even greater than we observe on code size.
+    Kiselyov's new translation algorithms are shown to significantly boost performance in our benchmarks.
+    The effect on execution time is even greater than we observe on code size.
     % - 1 sentence conclusion
     We conclude that the new translation algorithms significantly improve performance, and there may be further gains to be had by combining ideas from the different translation algorithms.
-    Many opportunities remain to improve the basic combinator graph reduction engine, such as the integration of other ideas from TIGRE, a concurrent garbage collector, and parallel reduction.
+
+    %Many opportunities remain to improve the basic combinator graph reduction engine, such as the integration of other ideas from TIGRE, a concurrent garbage collector, and parallel reduction.
     % - 1 sentence implication
-    Our results show that there are significant improvements to be made in this area of research that has been largely static in recent years, and we hope there will be renewed interest from the community to do research in this field.
+    %Our results show that there are significant improvements to be made in this area of research that has been largely static in recent years, and we hope there will be renewed interest from the community to do research in this field.
 \end{abstract}
@@
     \item We compare our TIGRE-style engine to our Miranda-style engine, to see if the techniques described in~\cite{koopman_fresh_1989} are viable on the X86 architecture.
-    \item We compare the traditional bracket abstraction and the different semantic translation algorithms from~\cite{kiselyov_lambda_2018} by compiling the same source programs with each translation algorithm and measuring the runtime on our Miranda-style engine, to see if the semantic translation algorithms produce compiled programs that are faster to execute.
+    \item We compare the traditional bracket abstraction and the various semantic translation algorithms from~\cite{kiselyov_lambda_2018} by compiling the same source programs with each translation algorithm and measuring the runtime on our Miranda-style engine, to see if the semantic translation algorithms produce compiled programs that execute faster.
 \end{itemize}
@@
 \subsection{Comparison with original Miranda}
-Figure~\ref{fig:turner-vs-miranda} shows the performance of our Miranda-style engine with the traditional bracket abstraction compared to the original Miranda~\cite{noauthor_open_2021}.
+Figure~\ref{fig:turner-vs-miranda} shows the performance of our Miranda-style engine with the traditional bracket abstraction compared to the original Miranda~\cite{turner_open_2021}.
 The test program used for the benchmark is the function given in figure~\ref{fig:fib}.
@@
 We observe that our Miranda-style engine is more than twice as fast as the original Miranda.
-While we have not attempted an in-depth evaluation of the performance differences between these two systems, we can point to two likely reasons that our implementation performs better on this benchmark:
+While we have not attempted an in-depth evaluation of the performance differences, we can point to two likely reasons why our implementation performs better on this benchmark:
 \begin{itemize}
@@
           They are implemented by the high-performance GNU MP~\cite{gnump} library but are still slower than the native 32-bit machine integers that our engine uses.
-    \item As mentioned in~\ref{sec:miranda}, the original Miranda must be compiled with optimizations disabled.
+    \item As mentioned in Section~\ref{sec:miranda}, the original Miranda must be compiled with optimizations disabled.
           Without these optimizations, the compiler is likely to generate a slower implementation of key functions.

### 2023-01-25 experiments/cast/youtubedl/clean_subs.sh
Script for cleaning up YouTube subs.

@@ -1,0 +1,12 @@
+#!/bin/bash
+# Run this script within a directory of subtitle files (.en.vtt extension).
+
+set -x
+
+for file in *.en.vtt; do
+	echo "Removing align and position from '$file'"
+	sed -i 's/align:start position:0%//' "$file"
+
+	echo "Removing word highlight from '$file'"
+	sed -i 's/<[^>]*>//g' "$file"
+done

### 2023-01-28 experiments/superg/doc/paper/paper.tex
Final general comments by Tom.

@@
 
-\subsection{Bracket abstraction}
+\subsection{Bracket Abstraction}
+
+\subsection{Graph Reduction}
+Must cover: outermost reduction.
 
@@
 \subsection{Strict algorithm}
-This is based on the rules in figure 6 of Kiselyov's paper.
+This is based on the rules in Figure 6 of Kiselyov's paper.
 We call it \textit{strict} it does not have lazy weakening: most notably it lacks the K-optimization required for full laziness.
@@
 \label{sec:miranda}
-For our base implementation, we try to stick to Miranda's design as much as possible.
+For our base implementation of a Combinator Graph Reduction engine, we try to stick to Miranda's design as much as possible.
 While there is published work on Miranda~\cite{turner_miranda_1985}, it does not provide much detail on the reduction engine.
@@
 
-Miranda uses a graph reduction engine, so after a program has been converted into combinatory logic using Turner's abstraction algorithm~\cite{turner_another_1979}, it is loaded into a graph structure.
-The graph is binary: every node has exactly two children, and the node itself represents the application of the right node to the left one.
+The approach we describe here is the one taken for our implementation, which may not fully correspond to Miranda's.
+Not all implementation issues are fully described in the literature, and in places where Miranda's implementation was not clear to us either, we developed our own solution.
+In particular, Section~\ref{sec:strict} covers \emph{our} approach to the resumption of the main reduction after arguments to strict combinators have been reduced, but this may be implemented differently in Miranda.
+
+Miranda converts source programs into combinatory logic using Turner's modified bracket abstraction algorithm~\cite{turner_another_1979}.
+The converted program, composed of combinators, constants and function application, is then loaded into a graph structure.
+The graph is binary: every node has exactly two children, and the node itself represents the application of the left node to the right one.
 Let us start with an example program $SKK \ 3$.
-The $S$ combinator takes 3 arguments, but currying lets us write it as $((SK) \; K)\ 3$.
-In this form, it is straightforward to construct the associated binary graph, shown in figure~\ref{fig:skk3}.
+The $S$ combinator takes three arguments, but a node can only store the application to a single argument.
+The left associativity of function application lets us rewrite the example program as $((SK) \; K)\ 3$.
+In this form, it is straightforward to construct the associated binary graph, shown in Figure~\ref{fig:skk3}.
 
 \begin{figure}
-    \includegraphics[width=.23\columnwidth]{skk3}
+    \includegraphics[width=.15\columnwidth]{skk3}
     \centering
@@
 
-In this example graph, after following two pointers we find the $S$ combinator, and we apply the definition $S f g x \Rightarrow (f x) (g x)$ to rewrite the graph, as shown in figure~\ref{fig:skk3_sred}.
+In this example graph, after following two pointers we find the $S$ combinator, and we apply the definition $S f g x \Rightarrow (f x) (g x)$ to rewrite the graph, as shown in Figure~\ref{fig:skk3_sred}.
 
 \begin{figure}
-    \includegraphics[width=.8\columnwidth]{skk3_sred}
+    \includegraphics[width=.51\columnwidth]{skk3_sred}
     \centering
@@
 The reduction process continues until the program has been reduced to a constant value, or there are insufficient arguments for the next combinator to reduce.
-The expression is said to be in \textit{weak head normal form}: some inner part of the expression may be reducible, but we only perform outermost reduction, so this is as close as we can get to normal form.
+The expression is said to be in \textit{weak head normal form}: some inner part of the expression may be reducible, but we only perform outermost reduction, so this is as close as we can get to the normal form.
 
 \subsection{Sharing}
-In the reduction in figure~\ref{fig:skk3_sred}, we have copied the value $3$ to two nodes.
+In the reduction in Figure~\ref{fig:skk3_sred}, we have copied the value $3$ to two nodes.
 For another reduction where the $x$ value is a pointer to a subgraph, performing the reduction may result in multiple pointers to the same expression.
@@
 It turns out that this is not too difficult to implement: when we make a reduction step, we update the outermost node with the contents of the new outermost node.
-In figure~\ref{fig:skk3_sred}, this would mean we overwrite node 1 with the contents of node 4.
+In Figure~\ref{fig:skk3_sred}, this would mean we overwrite node 1 with the contents of node 4.
 Now any other pointers to node 1 automatically point to the reduced equivalent.
@@
 
-It is unclear why a representation with three separate arrays was chosen over a single array where each element contains a 'cell' struct with a head, tail and tag.
+It is unclear why a representation with three separate arrays was chosen over a single array where each element contains a `cell' struct with a head, tail, and tag.
 We suspect decomposing the graph into separate arrays leads to poor data locality, as almost any operation will access all three parts of a cell.
@@
 \subsection{Strict combinators}
+\label{sec:strict}
 Miranda has special built-in combinators to perform arithmetic operations and compare numbers.
 Where most combinators place no restrictions on their arguments, these combinators require their arguments to be fully reduced to integers (hence call them \textit{strict}), and so require special treatment.
-We found Miranda's implementation difficult to follow, the approach we describe here is the one taken for our implementation, and may not fully correspond to Miranda's.
 
@@
 The graph reduction process may lead to some nodes becoming unreachable.
-Coming back to the example in figure~\ref{fig:skk3_sred}, after the reduction step nodes 2 and 3 are no longer used: they are \textit{garbage}.
+Coming back to the example in Figure~\ref{fig:skk3_sred}, after the reduction step nodes 2 and 3 are no longer used: they are \textit{garbage}.
 In this example, it is easy to conclude which nodes are garbage, but in larger programs, there are often multiple references to the same node, so we cannot mark nodes as garbage during the reduction step.
@@
 An apparent downside of this is that simple expressions require more indirection, and therefore a larger graph to reduce.
-Take for example the expression $1+2$, which in the Miranda-style engine can be represented with just two nodes, shown in figure~\ref{fig:1_plus_2_miranda}.
+Take for example the expression $1+2$, which in the Miranda-style engine can be represented with just two nodes, shown in Figure~\ref{fig:1_plus_2_miranda}.
 
@@
 
-The TIGRE version of this program, shown in figure~\ref{fig:1_plus_2_tigre}, is larger, occupying four nodes.
+The TIGRE version of this program, shown in Figure~\ref{fig:1_plus_2_tigre}, is larger, occupying four nodes.
 
@@
 We exploit this by encoding nodes as a \texttt{CALL} instruction to the left pointer, followed by the right pointer as a raw value.
-The memory layout of a node is illustrated in figure~\ref{fig:tigre_layout}.
+The memory layout of a node is illustrated in Figure~\ref{fig:tigre_layout}.
 
@@
 Combinators that return a value (\texttt{LIT} or strict combinators) can be implemented by computing the value and returning it with \texttt{RET}.
-See for example the implementation of \texttt{LIT} in figure~\ref{fig:combLIT}.
+See for example the implementation of \texttt{LIT} in Figure~\ref{fig:combLIT}.
 
@@
 Figure~\ref{fig:turner-vs-miranda} shows the performance of our Miranda-style engine with the traditional bracket abstraction compared to the original Miranda~\cite{turner_open_2021}.
-The test program used for the benchmark is the function given in figure~\ref{fig:fib}.
+The test program used for the benchmark is the function given in Figure~\ref{fig:fib}.
 
@@
 Both engines are configured to use the same compiler, so they will execute identical combinator expressions.
-The results can be seen in figure~\ref{fig:turner-vs-tigre}.
+The results can be seen in Figure~\ref{fig:turner-vs-tigre}.
 
@@
 The bracket abstraction algorithm has been added as a reference point.
-The results can be seen in figure~\ref{fig:bracket-vs-kiselyov}.
+The results can be seen in Figure~\ref{fig:bracket-vs-kiselyov}.
 
@@
 For many of the test programs in~\cite{kiselyov_lambda_2018} it was shown to generate the shortest programs of all the algorithms presented there.
-This is also true for the Fibonacci program, as shown in figure~\ref{fig:fib-sizes}.
+This is also true for the Fibonacci program, as shown in Figure~\ref{fig:fib-sizes}.
 

### 2023-01-28 experiments/superg/doc/paper/paper.tex
Add intro to perf. eval.

@@
 
-\subsection{Comparison with original Miranda}
-Figure~\ref{fig:turner-vs-miranda} shows the performance of our Miranda-style engine with the traditional bracket abstraction compared to the original Miranda~\cite{turner_open_2021}.
-The test program used for the benchmark is the function given in Figure~\ref{fig:fib}.
-
-\begin{figure*}
-    \includesvg[width=\textwidth]{turner_vs_miranda}
-    \centering
-    \caption{
-        Time to execute the Fibonacci benchmark program on the original Miranda and our Miranda-style engine.
-        A lower value on the Y-axis is better.}
-    \label{fig:turner-vs-miranda}
-\end{figure*}
+The program used in these benchmarks is a recursive definition of a function to compute elements of the Fibonacci sequence.
+Figure~\ref{fig:fib} shows the definition of the function in the Miranda language (for use with the original Miranda), as well as a definition in the LISP-like language used by our tools.
+The two definitions are identical, all differences are syntactical.
+In our benchmarks, we vary the value of parameter $n$ between $10$ and $20$ to show how the execution time changes as we increase the number of recursive calls (and therefore the number of operations performed).
 
@@
 \end{figure}
+
+We use the Criterion~\cite{heisler_criterionrs_2023} benchmarking library to measure the performance of our engines and Miranda.
+The original Miranda does not capture performance metrics, so we apply a small patch to the source code to output the time elapsed during a reduction operation.
+The 20-line patch and instructions on how to apply it are included with the source code for our implementation.
+
+Benchmarks were run on an otherwise quiescent laptop equipped with an x86-64 AMD CPU (Ryzen 7 4800U), running Debian 11 (Bullseye).
+The original Miranda was compiled using GCC 10.2.1 without any optimization flags, to avoid breaking the garbage collector as discussed in Section~\ref{sec:miranda}.
+All other code is written in Rust and compiled using the \texttt{release} profile (\texttt{rustc} version 1.65.0).
+
+\subsection{Comparison with original Miranda}
+Figure~\ref{fig:turner-vs-miranda} shows the performance of our Miranda-style engine with the traditional bracket abstraction compared to the original Miranda~\cite{turner_open_2021}.
+
+\begin{figure*}
+    \includesvg[width=\textwidth]{turner_vs_miranda}
+    \centering
+    \caption{
+        Time to execute the Fibonacci benchmark program on the original Miranda and our Miranda-style engine.
+        A lower value on the Y-axis is better.}
+    \label{fig:turner-vs-miranda}
+\end{figure*}
 

### 2023-01-28 experiments/superg/doc/paper/paper.tex
Plot for turner v. Miranda.

@@
 The 20-line patch and instructions on how to apply it are included with the source code for our implementation.
+We use Criterion's default measurement strategy: each benchmark runs repeatedly for 5 seconds as a warmup, after which 100 samples of the execution time are collected.
+Finally, an estimate of the average execution time is computed using linear regression over the samples.
 
@@
 
-\begin{figure*}
-    \includesvg[width=\textwidth]{turner_vs_miranda}
+\begin{figure}
+    %\includesvg[width=\textwidth]{turner_vs_miranda}
+    \begin{tikzpicture}
+        \begin{axis}[
+                title={Time to execute \texttt{fib} $n$},
+                xlabel={Parameter $n$},
+                xtick={10, 15, 20},
+                ylabel={Execution time (in milliseconds)},
+                ymin = 0,
+                scaled y ticks = false,
+                legend pos = north west,
+            ]
+            \addplot[
+                color = blue,
+                mark=square,
+            ]
+            coordinates {
+                    (10.0, 0.18171432823529418)
+                    (15.0, 1.7536414818181825)
+                    (20.0, 17.489519720000008)
+                };
+
+            \addplot[
+                color = red,
+                mark=o,
+            ]
+            coordinates {
+                    (10.0, 0.05903952613565834)
+                    (15.0, 0.6745943900990098)
+                    (20.0, 7.2472888442857135)
+                };
+
+            \legend{Miranda(original), Miranda-style Engine}
+
+        \end{axis}
+    \end{tikzpicture}
     \centering
@@
     \label{fig:turner-vs-miranda}
-\end{figure*}
+\end{figure}
 

### 2023-01-28 experiments/superg/doc/paper/benches/typeset.py

@@ -1,0 +1,79 @@
+#!/usr/bin/env python3
+"""
+Parses Criterion benchmark results for superg, and outputs a Latex/pgfplots-friendy format.
+"""
+
+import sys
+import json
+
+ENGINES = {
+    "Miranda": "Miranda (original)",
+    "TurnerEngine": "Miranda-style Engine",
+    "TigreEngine": "TIGRE-style Engine",
+}
+
+BENCHMARKS = {
+    "Fibonacci (Turner vs. Miranda)",
+    "Fibonacci (Turner vs. TIGRE)",
+    "Fibonacci n (All compilers)",
+    "Ackermann 3 n (All compilers)",
+}
+
+COMPILERS = {
+    "Bracket",
+    "Kiselyov - Strict",
+    "Kiselyov - Lazy",
+    "Kiselyov - LazyOpt",
+    "Kiselyov - Linear",
+}
+
+miranda_coords = []
+turner_coords = []
+
+def handle_message(msg):
+    if msg["reason"] != "benchmark-complete":
+        return
+    benchmark, config, val = tuple(msg["id"].split("/"))
+    val = float(val)
+    score = msg["typical"]["estimate"]
+
+    # Turner v. Miranda
+    if benchmark != "Fibonacci (Turner vs. Miranda)":
+        return
+    if config == "Miranda":
+        miranda_coords.append((val, score))
+    elif config == "TurnerEngine (Bracket)":
+        turner_coords.append((val, score))
+    
+
+with open(sys.argv[1]) as f:
+    for line in f:
+        msg = json.loads(line)
+        handle_message(msg)
+
+print("=== Original Miranda vs. Turner ===")
+print("""
+\\addplot[
+    color = blue,
+    mark=square,
+    ]
+    coordinates {""")
+for (val, ns) in miranda_coords:
+    ms = ns / 1000.0 / 1000.0
+    print(f"        ({val}, {ms})")
+print("""    };
+""")
+
+print("""
+\\addplot[
+    color = red,
+    mark=o,
+    ]
+    coordinates {""")
+for (val, ns) in turner_coords:
+    ms = ns / 1000.0 / 1000.0
+    print(f"        ({val}, {ms})")
+print("""    };
+""")
+
+print("\\legend{Miranda(original), Miranda-style Engine}")

### 2023-01-28 experiments/superg/doc/paper/benches/typeset.py
Better plots.

@@ -6,29 +6,21 @@ Parses Criterion benchmark results for superg, and outputs a Latex/pgfplots-frie
 import sys
 import json
 
-ENGINES = {
-    "Miranda": "Miranda (original)",
-    "TurnerEngine": "Miranda-style Engine",
-    "TigreEngine": "TIGRE-style Engine",
-}
-
-BENCHMARKS = {
-    "Fibonacci (Turner vs. Miranda)",
-    "Fibonacci (Turner vs. TIGRE)",
-    "Fibonacci n (All compilers)",
-    "Ackermann 3 n (All compilers)",
-}
+BASE_DIR="experiments/superg/doc/paper/benches"
 
 COMPILERS = {
-    "Bracket",
-    "Kiselyov - Strict",
-    "Kiselyov - Lazy",
-    "Kiselyov - LazyOpt",
-    "Kiselyov - Linear",
+    "Bracket": "Bracket Abstraction",
+    "Kiselyov - Strict": "Kiselyov - $\\strict$",
+    "Kiselyov - Lazy": "Kiselyov - $\\lazy$",
+    "Kiselyov - LazyOpt": "Kiselyov - $\\lazyeta$",
+    "Kiselyov - Linear": "Kiselyov - $\\linear$",
 }
 
 miranda_coords = []
 turner_coords = []
+tigre_coords = []
+
+compiler_coords = {c: [] for c in COMPILERS.values()}
 
 def handle_message(msg):
     if msg["reason"] != "benchmark-complete":
@@ -38,42 +30,89 @@ def handle_message(msg):
     score = msg["typical"]["estimate"]
 
     # Turner v. Miranda
-    if benchmark != "Fibonacci (Turner vs. Miranda)":
-        return
-    if config == "Miranda":
-        miranda_coords.append((val, score))
-    elif config == "TurnerEngine (Bracket)":
-        turner_coords.append((val, score))
+    if benchmark == "Fibonacci (Turner vs. Miranda)":
+        if config == "Miranda":
+            miranda_coords.append((val, score))
+        elif config == "TurnerEngine (Bracket)":
+            turner_coords.append((val, score))
+    # Turner v. TIGRE
+    if benchmark == "Fibonacci (Turner vs. TIGRE)":
+        if config == "TigreEngine (Bracket)":
+            tigre_coords.append((val, score))
+        # Turner results are the same as Turner v. Miranda
+    
+    if benchmark == "Fibonacci n (All compilers)":
+        compiler = COMPILERS[config]
[…]

### 2023-01-29 experiments/superg/doc/paper/paper.tex
Better plots.
Start to move around kiselyov algos.

@@
 \usepackage{svg}
+\usepackage{placeins}
 
@@
 Our description places no requirements on the type system and can be directly implemented in any general-purpose programming language.
+For the first and most basic algorithm, we give full definitions of all functions inline.
+Definitions for the more advanced algorithms are given in Appendix~\ref{app:kiselyov}.
 
@@
 As Kiselyov already notes in his treatment of the linear algorithm further on in the paper, the translation does not actually care what elements are in the context, just that we keep track of the size of the context.
-This lets us express compilation as a recursive function:
-
+This lets us express compilation as a recursive function, shown in Figure~\ref{fig:strict}.
+
+\begin{figure}
     \begin{equation*}
@@
     \end{equation*}
+    \centering
+    \caption{Compilation function for the $\strict$ algorithm.}
+    \label{fig:strict}
+\end{figure}
 
@@
 if the context is empty, the first rule applies, otherwise, we use the second one.
-The semantic function may be defined as:
-
+The semantic function may be defined as in Figure~\ref{fig:strict-sem}.
+
+\begin{figure}
     \begin{equation*}
@@
     \end{equation*}
+    \centering
+    \caption{Semantic function for the $\strict$ algorithm.}
+    \label{fig:strict-sem}
+\end{figure}
 
@@
             ]
-            \addplot[
-                color = blue,
-                mark=square,
-            ]
-            coordinates {
-                    (10.0, 0.18171432823529418)
-                    (15.0, 1.7536414818181825)
-                    (20.0, 17.489519720000008)
-                };
-
-            \addplot[
-                color = red,
-                mark=o,
-            ]
-            coordinates {
-                    (10.0, 0.05903952613565834)
-                    (15.0, 0.6745943900990098)
-                    (20.0, 7.2472888442857135)
-                };
-
-            \legend{Miranda(original), Miranda-style Engine}
+            \input{benches/miranda_vs_turner.tex}
 
@@
 
-\begin{figure*}
-    \includesvg[width=\textwidth]{turner_vs_tigre}
+\begin{figure}
+    %\includesvg[width=\textwidth]{turner_vs_tigre}
+    \begin{tikzpicture}
+        \begin{axis}[
+                title={Time to execute \texttt{fib} $n$},
+                xlabel={Parameter $n$},
+                xtick={10, 15, 20},
+                ylabel={Execution time (in milliseconds)},
+                ymin = 0,
+                scaled y ticks = false,
+                legend pos = north west,
+            ]
+            \input{benches/turner_vs_tigre.tex}
+
+        \end{axis}
+    \end{tikzpicture}
     \centering
@@
     \label{fig:turner-vs-tigre}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure*}
-    \includesvg[width=\textwidth]{bracket_vs_kiselyov}
+\begin{figure}
+    %\includesvg[width=\textwidth]{bracket_vs_kiselyov}
+    \begin{tikzpicture}
+        \begin{axis}[
+                title={Time to execute \texttt{fib} $n$},
+                width=\columnwidth,
+                height=1.5\columnwidth,
+                xlabel={Parameter $n$},
+                xtick={10, 15, 20},
+                ylabel={Execution time (in milliseconds)},
+                ymin = 0,
+                scaled y ticks = false,
+                legend pos = north west,
+            ]
+            \input{benches/bracket_vs_kiselyov.tex}
+        \end{axis}
+    \end{tikzpicture}
     \centering
@@
     \label{fig:bracket-vs-kiselyov}
-\end{figure*}
+\end{figure}
 
@@
 
+\newpage
 \appendix
@@
 
-\newpage
-
+\FloatBarrier
+\subsection{Semantic translation definitions}
+\label{app:kiselyov}.
+
+\begin{figure*}
+    \begin{equation*}
+        \begin{array}{l l l l}
+            \lazy & z         & = & t \models I                                                             \\
+            \lazy & s \ e     & = & \Gamma,f \models c                                                      \\
+                  &           &   & \text{where } (\Gamma \models c) := \lazy e                             \\
+            \\
+            \lazy & \lambda e & = & \emptyset \models K \; c                                                \\
+                  &           &   & \text{where } (\emptyset \models c) := \lazy e                          \\
+            \lazy & \lambda e & = & (\emptyset \models K) \coprod (\Gamma \models c)                        \\
+                  &           &   & \text{where } (\Gamma,f \models c) := \lazy e                           \\
+            \lazy & \lambda e & = & \Gamma \models K \; c                                                   \\
+                  &           &   & \text{where } (\Gamma,t \models c) := \lazy e                           \\
+            \\
+            \lazy & e_1 \ e_2 & = & (\Gamma \models ((\Gamma_1 \models c_1) \coprod (\Gamma_2 \models c_2)) \\
+                  &           &   & \text{where } (\Gamma_1 \models c_1) := \lazy e_1                       \\
+                  &           &   & \text{where } (\Gamma_2 \models c_2) := \lazy e_2                       \\
+                  &           &   & \text{where } \Gamma := \Gamma_1 \sqcup \Gamma_2                        \\
+        \end{array}
+    \end{equation*}
+    \centering
+    \caption{Compilation function with Lazy Weakening.}
+    \label{fig:lazy}
+\end{figure*}
+
+\begin{figure*}
+    \begin{equation*}
+        \begin{array}{l l l l l}
+            \Gamma_1    & \sqcup & \emptyset   & = & \Gamma_1                     \\
+            \emptyset   & \sqcup & \Gamma_2    & = & \Gamma_2                     \\
+            \Gamma_1,t  & \sqcup & \Gamma_2,\_ & = & (\Gamma_1 \sqcup \Gamma_2),t \\
+            \Gamma_1,\_ & \sqcup & \Gamma_2,t  & = & (\Gamma_1 \sqcup \Gamma_2),t \\
+            \Gamma_1,f  & \sqcup & \Gamma_2,f  & = & (\Gamma_1 \sqcup \Gamma_2),f \\
+        \end{array}
+    \end{equation*}
+    \centering
+    \caption{Semantic function with Lazy Weakening.}
+    \label{fig:lazy-sem}
+\end{figure*}
+
+\begin{figure*}
+    \begin{equation*}
+        \begin{array}{l l l l}
+            \lazyeta & z         & = & \emptyset \models V                                                          \\
+            \lazyeta & s \ e     & = & \Gamma,f \models c                                                           \\
+                     &           &   & \text{where } (\Gamma \models c) := \lazyeta e                               \\
+            \\
+            \lazyeta & \lambda e & = & \emptyset \models I                                                          \\
+                     &           &   & \text{where } (\emptyset \models V) := \lazyeta e                            \\
+            \lazyeta & \lambda e & = & \emptyset \models K \; c                                                     \\
+                     &           &   & \text{where } (\emptyset \models c) := \lazyeta e                            \\
+            \lazyeta & \lambda e & = & (\emptyset \models K) \coprod_\eta (\Gamma \models c)                        \\
+                     &           &   & \text{where } (\Gamma,f \models c) := \lazyeta e                             \\
+            \lazyeta & \lambda e & = & \Gamma \models K \; c                                                        \\
+                     &           &   & \text{where } (\Gamma,t \models c) := \lazyeta e                             \\
+            \\
+            \lazyeta & e_1 \ e_2 & = & (\Gamma \models ((\Gamma_1 \models c_1) \coprod_\eta (\Gamma_2 \models c_2)) \\
+                     &           &   & \text{where } (\Gamma_1 \models c_1) := \lazyeta e_1                         \\
+                     &           &   & \text{where } (\Gamma_2 \models c_2) := \lazyeta e_2                         \\
+                     &           &   & \text{where } \Gamma := \Gamma_1 \sqcup_\eta \Gamma_2                        \\
+        \end{array}
+    \end{equation*}
+    \centering
+    \caption{Compilation function with Eta optimization.}
+    \label{fig:lazyeta}
+\end{figure*}
+
+\begin{figure*}
+
+    \begin{equation*}
+        \begin{array}{l l l l l}
+            (\Gamma_1,\_ \models c_1) & \sqcup_\eta & (\emptyset \models V)     & = & \Gamma_1,t                        \\
+            (\emptyset \models V)     & \sqcup_\eta & (\Gamma_2,\_ \models c_2) & = & \Gamma_2,t                        \\
+            (\emptyset \models c_1)   & \sqcup_\eta & (\emptyset \models V)     & = & t                                 \\
+            (\emptyset \models V)     & \sqcup_\eta & (\emptyset \models c_2)   & = & t                                 \\
+            (\emptyset \models V)     & \sqcup_\eta & (\emptyset \models V)     & = & \mathbf{impossible}               \\
+            \\
+            % Original ones
+            (\Gamma_1 \models c_1)    & \sqcup_\eta & (\emptyset \models c_2)   & = & \Gamma_1                          \\
+            (\emptyset \models c_1)   & \sqcup_\eta & (\Gamma_2 \models c_2)    & = & \Gamma_2                          \\
+            (\Gamma_1,t \models c_1)  & \sqcup_\eta & (\Gamma_2,\_ \models c_2) & = & (\Gamma_1 \sqcup_\eta \Gamma_2),t \\
+            (\Gamma_1,\_ \models c_1) & \sqcup_\eta & (\Gamma_2,t \models c_2)  & = & (\Gamma_1 \sqcup_\eta \Gamma_2),t \\
+            (\Gamma_1,f \models c_1)  & \sqcup_\eta & (\Gamma_2,f \models c_2)  & = & (\Gamma_1 \sqcup_\eta \Gamma_2),f \\
+        \end{array}
+    \end{equation*}
+    \centering
+    \caption{Context merge function for Eta optimization.}
+    \label{fig:lazyeta-ctx}
+\end{figure*}
+
+\begin{figure*}
+    \begin{equation*}
+        \arraycolsep=1.4pt
+        \begin{array}{l l l l l}
+            % Added for eta optimization
+            (\Gamma_1,f \models c_1) & \coprod_\eta & (\emptyset \models V)    & = & c_1                                                        \\
+            (\emptyset \models V)    & \coprod_\eta & (\Gamma_2,f \models c_2) & = & (\emptyset \models CI) \coprod_\eta (\Gamma_2 \models c_2) \\
+
+            (\Gamma_1,t \models c_1) & \coprod_\eta & (\emptyset \models V)    & = & (\Gamma_1 \models (s \coprod_\eta l)) \coprod_\eta i       \\
+                                     &              &                          &   & \text{where } l := (\Gamma_1 \models c_1 )                 \\
+                                     &              &                          &   & \text{where } s := (\emptyset \models S )                  \\
+                                     &              &                          &   & \text{where } i := (\emptyset \models I )                  \\
+            (\emptyset \models V)    & \coprod_\eta & (\Gamma_2,t \models c_2) & = & (\emptyset \models SI) \coprod_\eta (\Gamma_2 \models c_2) \\
+            (\emptyset \models c_1)  & \coprod_\eta & (\emptyset \models V)    & = & c_1                                                        \\
+            (\emptyset \models V)    & \coprod_\eta & (\emptyset \models c_2)  & = & CI\ c_2                                                    \\
+            (\emptyset \models V)    & \coprod_\eta & (\emptyset \models V)    & = & \mathbf{impossible}                                        \\
+            \\
+
+            (\Gamma_1 \models c_1)   & \coprod_\eta & (\Gamma_2 \models)       & = & (\Gamma_1 \models c_1) \coprod (\Gamma_2 \models)          \\
+        \end{array}
+    \end{equation*}
+    \centering
+    \caption{Semantic function with Eta optimization.}
+    \label{fig:lazyeta-sem}
+\end{figure*}
+
+\begin{figure*}
+    \begin{equation*}
+        \begin{array}{l l l l}
+            \linear & z         & = & 1 \models I                                                              \\
+            \linear & s \ e     & = & n+1 \models (0 \models K) \coprod (n \models c)                          \\
+                    &           &   & \text{where } (n \models c) := \linear e                                 \\
+            \\
+            \linear & \lambda e & = & 0 \models K \; c                                                         \\
+                    &           &   & \text{where } (0 \models c) := \linear e                                 \\
+            \linear & \lambda e & = & (n-1) \models K \; c                                                     \\
+                    &           &   & \text{where } (n \models c) := \linear e                                 \\
+            \\
+            \linear & e_1 \ e_2 & = & (n \models ((n_1 \models c_1) \coprod_{\text{linear}} (n_2 \models c_2)) \\
+                    &           &   & \text{where } (n_1 \models c_1) := \linear e_1                           \\
+                    &           &   & \text{where } (n_2 \models c_2) := \linear e_2                           \\
+                    &           &   & \text{where } n := \max n_1 \, n_2                                       \\
+        \end{array}
+    \end{equation*}
+    \centering
+    \caption{Compilation function for Linear algorithm}
+    \label{fig:linear}
+\end{figure*}
+
+TODO
+
+\begin{figure*}
+    \begin{equation*}
+        \arraycolsep=1.4pt
+        \begin{array}{l l l l l}
+            (0 \models c_1)   & \coprod & (0 \models c_2)   & = & c_1 \ c_2                               \\
+            \\
+            (0 \models c_1)   & \coprod & (n \models c_2)   & = & B_n \ c_1 \ c_2                         \\
+            \\
+            (n \models c_1)   & \coprod & (0 \models c_2)   & = & C_n \ c_1 \ c_2                         \\
+            \\
+            (n \models c_1)   & \coprod & (n \models c_2)   & = & S_n \ c_1 \ c_2                         \\
+            \\
+            (n_1 \models c_1) & \coprod & (n_2 \models c_2) & = & B_{n_2-n_1} \ (S_{n_1} \ c_1) \ c_2     \\
+                              &         &                   &   & \text{where } n_1 < n_2                 \\
+            (n_1 \models c_1) & \coprod & (n_2 \models c_2) & = & C_{n_1-n_2} \ (B_{n_1-n_2} \ c_1) \ c_2 \\
+                              &         &                   &   & \text{where } n_1 > n_2                 \\
+        \end{array}
+    \end{equation*}
+    \centering
+    \caption{Semantic function for Linear algorithm}
+    \label{fig:linear-sem}
+\end{figure*}
+
+\FloatBarrier
 \subsection{Ackermann benchmark results}
@@
 
+\FloatBarrier
+
 \end{document}

### 2023-01-29 experiments/superg/doc/paper/paper.tex
Insert refs for Kiselyov.

@@
 We always match the last element of lists and represent the remainder of the list with $\Gamma$.
-
-\begin{equation*}
-    \begin{array}{l l l l}
-        \lazy & z         & = & t \models I                                                             \\
-        \lazy & s \ e     & = & \Gamma,f \models c                                                      \\
-              &           &   & \text{where } (\Gamma \models c) := \lazy e                             \\
-        \\
-        \lazy & \lambda e & = & \emptyset \models K \; c                                                \\
-              &           &   & \text{where } (\emptyset \models c) := \lazy e                          \\
-        \lazy & \lambda e & = & (\emptyset \models K) \coprod (\Gamma \models c)                        \\
-              &           &   & \text{where } (\Gamma,f \models c) := \lazy e                           \\
-        \lazy & \lambda e & = & \Gamma \models K \; c                                                   \\
-              &           &   & \text{where } (\Gamma,t \models c) := \lazy e                           \\
-        \\
-        \lazy & e_1 \ e_2 & = & (\Gamma \models ((\Gamma_1 \models c_1) \coprod (\Gamma_2 \models c_2)) \\
-              &           &   & \text{where } (\Gamma_1 \models c_1) := \lazy e_1                       \\
-              &           &   & \text{where } (\Gamma_2 \models c_2) := \lazy e_2                       \\
-              &           &   & \text{where } \Gamma := \Gamma_1 \sqcup \Gamma_2                        \\
-    \end{array}
-\end{equation*}
+Appendix~\ref{app:kiselyov} contains the new compilation function $\lazy$ (Figure~\ref{fig:lazy}).
 
 We have introduced yet another case for $\lambda e$, this time to distinguish between $t$ or $f$ as the last element in the context of the inner expression.
-There is also the new function $\sqcup$, which we need to implement:
-
-\begin{equation*}
-    \begin{array}{l l l l l}
-        \Gamma_1    & \sqcup & \emptyset   & = & \Gamma_1                     \\
-        \emptyset   & \sqcup & \Gamma_2    & = & \Gamma_2                     \\
-        \Gamma_1,t  & \sqcup & \Gamma_2,\_ & = & (\Gamma_1 \sqcup \Gamma_2),t \\
-        \Gamma_1,\_ & \sqcup & \Gamma_2,t  & = & (\Gamma_1 \sqcup \Gamma_2),t \\
-        \Gamma_1,f  & \sqcup & \Gamma_2,f  & = & (\Gamma_1 \sqcup \Gamma_2),f \\
-    \end{array}
-\end{equation*}
+There is also the new function $\sqcup$, whose definition is given in Figure~\ref{fig:lazy-ctx}.
 
@@
 
-Our semantic function now has to handle more cases, and thus is somewhat larger:
-
-\begin{equation*}
-    \arraycolsep=1.4pt
-    \begin{array}{l l l l l}
-        (\emptyset \models c_1)  & \coprod & (\emptyset \models c_2)  & = & c_1 \ c_2                                                \\
-        \\
-        (\emptyset \models c_1)  & \coprod & (\Gamma_2,t \models c_2) & = & (\emptyset \models B c_1) \coprod (\Gamma_2 \models c_2) \\
-        \\
-        (\Gamma_1,t \models c_1) & \coprod & (\emptyset \models c_2)  & = & (\emptyset \models C C c_2) \coprod (\Gamma \models c_1) \\
-        \\
-        (\Gamma_1,t \models c_1) & \coprod & (\Gamma_2,t \models c_2) & = & (\Gamma_1 \models (s \coprod l)) \coprod r)              \\
-                                 &         &                          &   & \text{where } s := (\emptyset \models S )                \\
-                                 &         &                          &   & \text{where } l := (\Gamma_1 \models c_1 )               \\
-                                 &         &                          &   & \text{where } r := (\Gamma_2 \models c_2 )               \\
-        \\
-        % The new additions
-        (\Gamma_1,f \models c_1) & \coprod & (\Gamma_2,f \models c_2) & = & (\Gamma_1 \models c_1) \coprod (\Gamma_2 \models c_2)    \\
-        \\
-        (\Gamma_1,f \models c_1) & \coprod & (\Gamma_2,t \models c_2) & = & (\Gamma_1 \models (b \coprod l)) \coprod r)              \\
-                                 &         &                          &   & \text{where } b := (\emptyset \models B )                \\
-                                 &         &                          &   & \text{where } l := (\Gamma_1 \models c_1 )               \\
-                                 &         &                          &   & \text{where } r := (\Gamma_2 \models c_2 )               \\
-        \\
-        (\Gamma_1,t \models c_1) & \coprod & (\Gamma_2,f \models c_2) & = & (\Gamma_1 \models (c \coprod l)) \coprod r)              \\
-                                 &         &                          &   & \text{where } c := (\emptyset \models C )                \\
-                                 &         &                          &   & \text{where } l := (\Gamma_1 \models c_1 )               \\
-                                 &         &                          &   & \text{where } r := (\Gamma_2 \models c_2 )               \\
-        \\
-        (\emptyset \models c_1)  & \coprod & (\Gamma_2,f \models c_2) & = & (\emptyset \models c_1) \coprod (\Gamma_2 \models c_2)   \\
-        \\
-        (\Gamma_1,f \models c_1) & \coprod & (\emptyset \models c_2)  & = & (\Gamma_1 \models c_1) \coprod (\emptyset \models c_2)   \\
-    \end{array}
-\end{equation*}
+Our semantic function now has to handle more cases, and thus is somewhat larger.
+See Figure~\ref{fig:lazy-sem} for the full definition.
 
@@
 Our implementation uses a tagged union to define a type that can be either an expression or the top variable.
-For the compilation function, we only need to change the definition of $\lazyeta \ z$ and add an extra case for $\lazyeta \ \lambda e$:
-
-\begin{equation*}
-    \begin{array}{l l l l}
-        \lazyeta & z         & = & \emptyset \models V                                                          \\
-        \lazyeta & s \ e     & = & \Gamma,f \models c                                                           \\
-                 &           &   & \text{where } (\Gamma \models c) := \lazyeta e                               \\
-        \\
-        \lazyeta & \lambda e & = & \emptyset \models I                                                          \\
-                 &           &   & \text{where } (\emptyset \models V) := \lazyeta e                            \\
-        \lazyeta & \lambda e & = & \emptyset \models K \; c                                                     \\
-                 &           &   & \text{where } (\emptyset \models c) := \lazyeta e                            \\
-        \lazyeta & \lambda e & = & (\emptyset \models K) \coprod_\eta (\Gamma \models c)                        \\
-                 &           &   & \text{where } (\Gamma,f \models c) := \lazyeta e                             \\
-        \lazyeta & \lambda e & = & \Gamma \models K \; c                                                        \\
-                 &           &   & \text{where } (\Gamma,t \models c) := \lazyeta e                             \\
-        \\
-        \lazyeta & e_1 \ e_2 & = & (\Gamma \models ((\Gamma_1 \models c_1) \coprod_\eta (\Gamma_2 \models c_2)) \\
-                 &           &   & \text{where } (\Gamma_1 \models c_1) := \lazyeta e_1                         \\
-                 &           &   & \text{where } (\Gamma_2 \models c_2) := \lazyeta e_2                         \\
-                 &           &   & \text{where } \Gamma := \Gamma_1 \sqcup_\eta \Gamma_2                        \\
-    \end{array}
-\end{equation*}
-
-Our context merge function $\sqcup$ now also depends on the compiled expression, since we have a special case for $V$:
-
-\begin{equation*}
-    \begin{array}{l l l l l}
-        (\Gamma_1,\_ \models c_1) & \sqcup_\eta & (\emptyset \models V)     & = & \Gamma_1,t                        \\
-        (\emptyset \models V)     & \sqcup_\eta & (\Gamma_2,\_ \models c_2) & = & \Gamma_2,t                        \\
-        (\emptyset \models c_1)   & \sqcup_\eta & (\emptyset \models V)     & = & t                                 \\
-        (\emptyset \models V)     & \sqcup_\eta & (\emptyset \models c_2)   & = & t                                 \\
-        (\emptyset \models V)     & \sqcup_\eta & (\emptyset \models V)     & = & \mathbf{impossible}               \\
-        \\
-        % Original ones
-        (\Gamma_1 \models c_1)    & \sqcup_\eta & (\emptyset \models c_2)   & = & \Gamma_1                          \\
-        (\emptyset \models c_1)   & \sqcup_\eta & (\Gamma_2 \models c_2)    & = & \Gamma_2                          \\
-        (\Gamma_1,t \models c_1)  & \sqcup_\eta & (\Gamma_2,\_ \models c_2) & = & (\Gamma_1 \sqcup_\eta \Gamma_2),t \\
-        (\Gamma_1,\_ \models c_1) & \sqcup_\eta & (\Gamma_2,t \models c_2)  & = & (\Gamma_1 \sqcup_\eta \Gamma_2),t \\
-        (\Gamma_1,f \models c_1)  & \sqcup_\eta & (\Gamma_2,f \models c_2)  & = & (\Gamma_1 \sqcup_\eta \Gamma_2),f \\
-    \end{array}
-\end{equation*}
+For the compilation function, we only need to change the definition of $\lazyeta \ z$ and add an extra case for $\lazyeta \ \lambda e$.
+Figure~\ref{fig:lazyeta} shows the new compilation function.
+
+Our context merge function $\sqcup$ now also depends on the compiled expression, since we have a special case for $V$.
+The updated definition is given in Figure~\ref{fig:lazyeta-ctx}.
 
 The $V$ value effectively encodes an additional implicit used element in the context.
-We also have some new cases for the semantic function:
-
-\begin{equation*}
-    \arraycolsep=1.4pt
-    \begin{array}{l l l l l}
-        % Added for eta optimization
-        (\Gamma_1,f \models c_1) & \coprod_\eta & (\emptyset \models V)    & = & c_1                                                        \\
-        (\emptyset \models V)    & \coprod_\eta & (\Gamma_2,f \models c_2) & = & (\emptyset \models CI) \coprod_\eta (\Gamma_2 \models c_2) \\
-
-        (\Gamma_1,t \models c_1) & \coprod_\eta & (\emptyset \models V)    & = & (\Gamma_1 \models (s \coprod_\eta l)) \coprod_\eta i       \\
-                                 &              &                          &   & \text{where } l := (\Gamma_1 \models c_1 )                 \\
-                                 &              &                          &   & \text{where } s := (\emptyset \models S )                  \\
-                                 &              &                          &   & \text{where } i := (\emptyset \models I )                  \\
-        (\emptyset \models V)    & \coprod_\eta & (\Gamma_2,t \models c_2) & = & (\emptyset \models SI) \coprod_\eta (\Gamma_2 \models c_2) \\
-        (\emptyset \models c_1)  & \coprod_\eta & (\emptyset \models V)    & = & c_1                                                        \\
-        (\emptyset \models V)    & \coprod_\eta & (\emptyset \models c_2)  & = & CI c_2                                                     \\
-        (\emptyset \models V)    & \coprod_\eta & (\emptyset \models V)    & = & \mathbf{impossible}                                        \\
-        \\
-
-        (\Gamma_1 \models c_1)   & \coprod_\eta & (\Gamma_2 \models)       & = & (\Gamma_1 \models c_1) \coprod (\Gamma_2 \models)          \\
-    \end{array}
-\end{equation*}
+We also have some new cases for the semantic function, available in Figure~\ref{fig:lazyeta-sem}.
 
@@
 The implementation of the linear algorithm is much simpler because it is based on the $\strict$ algorithm.
-The compilation function is identical save for a different semantic function $\coprod_{\text{linear}}$:
-
-\begin{equation*}
-    \begin{array}{l l l l}
-        \linear & z         & = & 1 \models I                                                              \\
-        \linear & s \ e     & = & n+1 \models (0 \models K) \coprod (n \models c)                          \\
-                &           &   & \text{where } (n \models c) := \linear e                                 \\
-        \\
-        \linear & \lambda e & = & 0 \models K \; c                                                         \\
-                &           &   & \text{where } (0 \models c) := \linear e                                 \\
-        \linear & \lambda e & = & (n-1) \models K \; c                                                     \\
-                &           &   & \text{where } (n \models c) := \linear e                                 \\
-        \\
-        \linear & e_1 \ e_2 & = & (n \models ((n_1 \models c_1) \coprod_{\text{linear}} (n_2 \models c_2)) \\
-                &           &   & \text{where } (n_1 \models c_1) := \linear e_1                           \\
-                &           &   & \text{where } (n_2 \models c_2) := \linear e_2                           \\
-                &           &   & \text{where } n := \max n_1 \, n_2                                       \\
-    \end{array}
-\end{equation*}
-
-below we give the corresponding semantic function, which is no longer recursive, making this algorithm linear:
-
-\begin{equation*}
-    \arraycolsep=1.4pt
-    \begin{array}{l l l l l}
-        (0 \models c_1)   & \coprod & (0 \models c_2)   & = & c_1 \ c_2                               \\
-        \\
-        (0 \models c_1)   & \coprod & (n \models c_2)   & = & B_n \ c_1 \ c_2                         \\
-        \\
-        (n \models c_1)   & \coprod & (0 \models c_2)   & = & C_n \ c_1 \ c_2                         \\
-        \\
-        (n \models c_1)   & \coprod & (n \models c_2)   & = & S_n \ c_1 \ c_2                         \\
-        \\
-        (n_1 \models c_1) & \coprod & (n_2 \models c_2) & = & B_{n_2-n_1} \ (S_{n_1} \ c_1) \ c_2     \\
-                          &         &                   &   & \text{where } n_1 < n_2                 \\
-        (n_1 \models c_1) & \coprod & (n_2 \models c_2) & = & C_{n_1-n_2} \ (B_{n_1-n_2} \ c_1) \ c_2 \\
-                          &         &                   &   & \text{where } n_1 > n_2                 \\
-    \end{array}
-\end{equation*}
+The compilation function is identical save for a different semantic function $\coprod_{\text{linear}}$.
+For the full definition, see Figure~\ref{fig:linear}.
+
+In figure~\ref{fig:linear-sem} we give the corresponding semantic function, which is no longer recursive, making this algorithm linear.
 
@@
 
-\begin{figure*}
+\begin{figure}
     \begin{equation*}
@@
     \label{fig:lazy}
-\end{figure*}
-
-\begin{figure*}
+\end{figure}
+
+\begin{figure}
     \begin{equation*}
@@
     \centering
+    \caption{Context merge function for Lazy Weakening.}
+    \label{fig:lazy-ctx}
+\end{figure}
+
+\begin{figure}
+    \begin{equation*}
+        \arraycolsep=1.4pt
+        \begin{array}{l l l l l}
+            (\emptyset \models c_1)  & \coprod & (\emptyset \models c_2)  & = & c_1 \ c_2                                                \\
+            \\
+            (\emptyset \models c_1)  & \coprod & (\Gamma_2,t \models c_2) & = & (\emptyset \models B c_1) \coprod (\Gamma_2 \models c_2) \\
+            \\
+            (\Gamma_1,t \models c_1) & \coprod & (\emptyset \models c_2)  & = & (\emptyset \models C C c_2) \coprod (\Gamma \models c_1) \\
+            \\
+            (\Gamma_1,t \models c_1) & \coprod & (\Gamma_2,t \models c_2) & = & (\Gamma_1 \models (s \coprod l)) \coprod r)              \\
+                                     &         &                          &   & \text{where } s := (\emptyset \models S )                \\
+                                     &         &                          &   & \text{where } l := (\Gamma_1 \models c_1 )               \\
+                                     &         &                          &   & \text{where } r := (\Gamma_2 \models c_2 )               \\
+            \\
+            % The new additions
+            (\Gamma_1,f \models c_1) & \coprod & (\Gamma_2,f \models c_2) & = & (\Gamma_1 \models c_1) \coprod (\Gamma_2 \models c_2)    \\
+            \\
+            (\Gamma_1,f \models c_1) & \coprod & (\Gamma_2,t \models c_2) & = & (\Gamma_1 \models (b \coprod l)) \coprod r)              \\
+                                     &         &                          &   & \text{where } b := (\emptyset \models B )                \\
+                                     &         &                          &   & \text{where } l := (\Gamma_1 \models c_1 )               \\
+                                     &         &                          &   & \text{where } r := (\Gamma_2 \models c_2 )               \\
+            \\
+            (\Gamma_1,t \models c_1) & \coprod & (\Gamma_2,f \models c_2) & = & (\Gamma_1 \models (c \coprod l)) \coprod r)              \\
+                                     &         &                          &   & \text{where } c := (\emptyset \models C )                \\
+                                     &         &                          &   & \text{where } l := (\Gamma_1 \models c_1 )               \\
+                                     &         &                          &   & \text{where } r := (\Gamma_2 \models c_2 )               \\
+            \\
+            (\emptyset \models c_1)  & \coprod & (\Gamma_2,f \models c_2) & = & (\emptyset \models c_1) \coprod (\Gamma_2 \models c_2)   \\
+            \\
+            (\Gamma_1,f \models c_1) & \coprod & (\emptyset \models c_2)  & = & (\Gamma_1 \models c_1) \coprod (\emptyset \models c_2)   \\
+        \end{array}
+    \end{equation*}
+    \centering
     \caption{Semantic function with Lazy Weakening.}
     \label{fig:lazy-sem}
-\end{figure*}
-
-\begin{figure*}
+\end{figure}
+
+\begin{figure}
     \begin{equation*}
@@
     \label{fig:lazyeta}
-\end{figure*}
-
-\begin{figure*}
+\end{figure}
+
+\begin{figure}
 
@@
     \label{fig:lazyeta-ctx}
-\end{figure*}
-
-\begin{figure*}
+\end{figure}
+
+\begin{figure}
     \begin{equation*}
@@
     \label{fig:lazyeta-sem}
-\end{figure*}
-
-\begin{figure*}
+\end{figure}
+
+\begin{figure}
     \begin{equation*}
@@
     \label{fig:linear}
-\end{figure*}
-
-TODO
-
-\begin{figure*}
+\end{figure}
+
+\begin{figure}
     \begin{equation*}
@@
     \label{fig:linear-sem}
-\end{figure*}
+\end{figure}
 

### 2023-01-30 experiments/superg/doc/paper/paper.tex
Resolve more TODOs.

@@
 The runtime need not concern itself with passing arguments, this is all handled by the combinators.
-The compiled program is executed using graph reduction, a technique that is also common in other reduction engines\cite{kieburtz_g-machine_1985,fairbairn_tim_1987}.
+
+\textbf{TODO: } Combinators $B$ and $C$.
 
@@
 \subsection{Strict algorithm}
-This is based on the rules in Figure 6 of Kiselyov's paper.
-We call it \textit{strict} it does not have lazy weakening: most notably it lacks the K-optimization required for full laziness.
-
-The most naive implementation is to perform the type checking, then use the built context to apply the elimination rules and the translated program.
-As Kiselyov already notes in his treatment of the linear algorithm further on in the paper, the translation does not actually care what elements are in the context, just that we keep track of the size of the context.
+The compilation function below, which we denote $\strict$, is based on the rules in Figure 6 of Kiselyov's paper~\cite{kiselyov_lambda_2018}.
+We call it \textit{strict} it does not have lazy weakening: most notably it lacks the K-optimization required for full laziness~\cite{spj_impl}.
+
+Kiselyov describes the algorithm as a type system with an operational semantics.
+The type system describes how to assign a type, or context, to a lambda calculus term.
+The operational semantics define how a typed term may be converted into combinatory logic.
+A naive implementation thus requires two phases: an initial typing step, followed by a translation step.
+As Kiselyov already notes in his treatment of the linear algorithm further on in the paper, the translation does not care what elements are in the context, just that we keep track of the size of the context.
 This lets us express compilation as a recursive function, shown in Figure~\ref{fig:strict}.
@@
 if the context is empty, the first rule applies, otherwise, we use the second one.
-The semantic function may be defined as in Figure~\ref{fig:strict-sem}.
+Figure~\ref{fig:strict-sem} defines the semantic function.
 
@@
 
-These simplified functions can be implemented directly and efficiently in a conventional programming language.
+These functions are equivalent to Kiselyov's type system and can be implemented directly and efficiently in a conventional programming language.
 
 \subsection{Lazy weakening}
-Adding lazy weakening requires that we track which items in the context are ignored.
-We change our context representation from an integer to a list of boolean values.
-In the new compilation functions below, we use a shorthand $t$ for the value \textit{true} and $f$ for \textit{false}.
-We always match the last element of lists and represent the remainder of the list with $\Gamma$.
-Appendix~\ref{app:kiselyov} contains the new compilation function $\lazy$ (Figure~\ref{fig:lazy}).
+The addition of Kiselyov's \textit{Lazy Weakening} requires that we track which items in the context are ignored.
+The context representation changes from an integer to a list of boolean values.
+In the new compilation function $\lazy$, shown in Figure~\ref{fig:lazy}, we use the shorthand $t$ for the value \textit{true} and $f$ for \textit{false}.
+We always match the last element of lists and denote the remainder of the list by $\Gamma$.
 
 We have introduced yet another case for $\lambda e$, this time to distinguish between $t$ or $f$ as the last element in the context of the inner expression.
-There is also the new function $\sqcup$, whose definition is given in Figure~\ref{fig:lazy-ctx}.
+There is also the new function $\sqcup$ to merge the left and right contexts of a function application term, whose definition is given in Figure~\ref{fig:lazy-ctx}.
+Merging follows the natural rule that if either of the two contexts requires a particular item, the merged context also requires it.
 
@@
 This builds the new context back to front, so as a final step, the array is reversed.
-
 Alternatively, an implementation can pad the shorter list (at the start!) with $f$ and zip them.
@@
 
-Our semantic function now has to handle more cases, and thus is somewhat larger.
+Our definition of the semantic function now has to handle more cases, and thus is somewhat larger.
 See Figure~\ref{fig:lazy-sem} for the full definition.
-
-The first 4 cases are direct translations from the previous definitions, the others are new additions to handle ignored elements.
-The last 2 cases are not included in the original paper, but they are necessary to make the pattern matching exhaustive, and are present in the reference OCaml implementation.
+The first four cases are direct translations from the previous definitions, the others are new additions to handle ignored elements.
+The last two cases are not included in the original paper, but they are necessary to make the pattern matching exhaustive, and are present in the reference OCaml implementation\footnote{From~\cite{kiselyov_lambda_2018}: \url{https://okmij.org/ftp/tagless-final/skconv.ml}}.
 
@@
 
-The case $(\emptyset \models V) \coprod (\emptyset \models V) $ is impossible in this translation, implementations may throw an error or leave it as undefined behavior.
-\textbf{TODO:} Why is this an impossible case?
+The case $(\emptyset \models V) \coprod (\emptyset \models V) $ is impossible in this translation:
+it represents the application of a value to itself, e.g. $\lambda x. x\ x$.
+For such an $x$, we cannot assign a proper type.
+Implementations may throw an error or leave it as undefined behavior.
 
@@
 The implementation of the linear algorithm is much simpler because it is based on the $\strict$ algorithm.
-The compilation function is identical save for a different semantic function $\coprod_{\text{linear}}$.
+The compilation function is identical save for a different semantic function $\coprod_{\text{linear}}$, shown in Figure~\ref{fig:linear-sem}.
 For the full definition, see Figure~\ref{fig:linear}.
 
-In figure~\ref{fig:linear-sem} we give the corresponding semantic function, which is no longer recursive, making this algorithm linear.
-
-\textbf{TODO: } Can we add lazy weakening and/or eta optimization?
+The main achievement is that the semantic function is no longer recursive.
+The $\linear$ algorithm visits each term exactly once, so arguing its linearity becomes an almost trivial exercise.
+
+We leave the addition of full laziness and $\eta$-optimization to the $\linear$ algorithm as future work.
 
@@
 \begin{figure}
-    \includegraphics[width=.8\columnwidth]{add_step1}
+    \includegraphics[width=.6\columnwidth]{add_step1}
     \centering
@@
 \begin{figure}
-    \includegraphics[width=.8\columnwidth]{add_step2}
+    \includegraphics[width=.6\columnwidth]{add_step2}
     \centering
@@
 Originally designed for the VAX instruction set, our implementation targets the x86-64 instruction set but mirrors the TIGRE design otherwise.
-The TIGRE paper focuses on key differences from Miranda, the elimination of tags and directly executable graph nodes, so where the paper omits implementation details we take the same approach as our reference implementation based on Miranda.
+The TIGRE paper focuses on key differences from Miranda, the elimination of tags and directly executable graph nodes, so where the TIGRE paper omits implementation details we take the same approach as our reference implementation based on Miranda.
 
@@
 \begin{figure}
-    \includegraphics[width=.2\columnwidth]{1_plus_2_miranda}
+    \includegraphics[width=.17\columnwidth]{1_plus_2_miranda}
     \centering
@@
 \begin{figure}
-    \includegraphics[width=.46\columnwidth]{1_plus_2_tigre}
+    \includegraphics[width=.38\columnwidth]{1_plus_2_tigre}
     \centering
@@
 
-\textbf{TODO: Maybe add a step-by-step reduction example?}
+%\textbf{TODO: Maybe add a step-by-step reduction example?}
 
@@
 
-It comes as no surprise that the shortest translation is also the fastest to execution, but it is worth noting that while the Bracket abstraction translation is only 55\% larger than the $\lazyeta$ one, it runs 94\% slower.
+It comes as no surprise that the shortest translation is also the fastest to execute, but it is worth noting that while the Bracket abstraction translation is only 55\% larger than the $\lazyeta$ one, it runs 94\% slower.
 We see a similar pattern for the Ackermann function, for which results may be found in appendix~\ref{app:ack}.
@@
 Our objective with this work was to revive research around Combinator Graph Reduction by improving the performance.
-We have attempted this by using a novel semantic translation algorithm to compile lambda calculus programs into combinator expressions~\cite{kiselyov_lambda_2018}, and by implementing combinator graph reduction engines based on the designs of Miranda~\cite{turner_new_1979} and TIGRE~\cite{koopman_fresh_1989}.
+We used a novel semantic translation algorithm to compile lambda calculus programs into smaller combinator expressions~\cite{kiselyov_lambda_2018} and have implemented modern combinator graph reduction engines based on the designs of Miranda~\cite{turner_new_1979} and TIGRE~\cite{koopman_fresh_1989}.
 
@@
 % Why is this the answer? Present the main arguments
-Unfortunately, its performance still lags behind the reference implementation because its inherent frequent writes to executable code sections trigger expensive invalidations of the instruction cache.
+Unfortunately, its performance still lags behind the Miranda-style engine because its inherent frequent writes to executable code sections trigger expensive invalidations of the instruction cache.
 We conclude that the TIGRE design is a poor fit for architectures with split instruction and data caches, though some concepts like \texttt{LIT} combinators and tag elimination may still prove effective.

### 2023-01-30 experiments/superg/src/kiselyov/linear.rs

@@ -1,4 +1,4 @@
-/// The strict compilation algorithm, as given in fig.10
+/// The linear compilation algorithm, as given in fig.10
 use super::BExpr;
 
 use crate::compiled_expr::*;

### 2023-01-30 experiments/superg/doc/paper/paper.tex
Graph reduction.

@@
 \label{sec:prelim}
-\textbf{TODO: write section. Currently contains only a few small fragments}
 \subsection{Lambda Calculus}
-
+For an accessible introduction to the Lambda Calculus, we can recommend~\cite{rojas_tutorial_2015}.
+Most introductory resources, including~\cite{rojas_tutorial_2015}, present Lambda Calculus with named bindings.
+Kiselyov's algorithms~\cite{kiselyov_lambda_2018} use De Bruijn indices~\cite{de_bruijn_lambda_1972} instead.
+
+%The lambda ($\lambda$) calculus is perhaps the smallest programming language.
+%At its core there is only the definition of functions (\emph{abstraction}) and the \emph{application} of a function to an argument.
+%We use the $\lambda$ symbol to denote abstraction, e.g.
+$\lambda x.x$ defines a function (unnamed) that takes an argument $x$ (the $x$ before the dot), and when applied to an $y$ returns that same value $y$.
+%This application may be written as $(\lambda x.x) y$.
+%$\lambda x.x$ is in fact the definition of the \emph{identity} function: a function that returns its argument unchanged.
+%we can nest abstractions to create functions that accept multiple arguments, for example $\lambda x.\lambda y. x$, a function that takes two arguments and returns the first (the second is discarded).
+%
+%Functions may be arguments to other functions: $(\lambda x.x) (\lambda x.x)$ is a valid expression in the lambda calculus (it evaluates to $\lambda x.x$).
+%
+%TODO: Outermost reduction
+%\url{https://personal.utdallas.edu/~gupta/courses/apl/lambda.pdf}
+%
 \subsection{Combinatory Logic}
-Starting from a lambda calculus expression, we first translate it into an expression consisting exclusively of the application of a small set of predefined functions, or \textit{combinators}.
-It turns out that any lambda calculus expression can be expressed using just three combinators:
-\begin{itemize}
-    \item $S \ f \ g \ x = f \ x \ (g \ x)$
-    \item $K \ x \ y = x$
-    \item $I \ x = x$\footnote{It turns out that $SKK = I$, so even this combinator is not strictly necessary, but all reasonable implementations do include $I$, and it simplifies the example.}
-\end{itemize}
-
-For example, $\lambda x.x+1$ can be converted into the combinator expression:
-
-\[
-    S \ (S \  (K \  +) \  I) \ (K \ 1)
-\]
-
-We invite the reader to verify this by applying $x$ to this expression and reducing it to $x+1$ with the usual $\beta$-reduction rule.
-While the generated expression is not as easy to read, the simplicity of its components is appealing: to execute a program compiled to combinators, an evaluator only needs to support four operations: the built-in $S$, $K$ and $I$ combinators, and function application\footnote{And any builtins of the language, like the plus operator in the example.}.
-Better still, conversion to combinators removes all bound variables.
-The runtime need not concern itself with passing arguments, this is all handled by the combinators.
-
-\textbf{TODO: } Combinators $B$ and $C$.
-
+Chapter 16 of~\cite{spj_impl} starts with a good introduction to Combinatory Logic and the Bracket Abstraction algorithm.
+
+%Starting from a lambda calculus expression, we first translate it into an expression consisting exclusively of the application of a small set of predefined functions, or \textit{combinators}.
+%It turns out that any lambda calculus expression can be expressed using just three combinators:
+%\begin{itemize}
+%    \item $S \ f \ g \ x = f \ x \ (g \ x)$
+%    \item $K \ x \ y = x$
+%    \item $I \ x = x$\footnote{It turns out that $SKK = I$, so even this combinator is not strictly necessary, but all reasonable implementations do include $I$, and it simplifies the example.}
+%\end{itemize}
+%
+%For example, $\lambda x.x+1$ can be converted into the combinator expression:
+%
+%\[
+%    S \ (S \  (K \  +) \  I) \ (K \ 1)
+%\]
+%
+%We invite the reader to verify this by applying $x$ to this expression and reducing it to $x+1$ with the usual $\beta$-reduction rule.
+%While the generated expression is not as easy to read, the simplicity of its components is appealing: to execute a program compiled to combinators, an evaluator only needs to support four operations: the built-in $S$, $K$ and $I$ combinators, and function application\footnote{And any builtins of the language, like the plus operator in the example.}.
+%Better still, conversion to combinators removes all bound variables.
+%The runtime need not concern itself with passing arguments, this is all handled by the combinators.
+%
+%\textbf{TODO: } Combinators $B$ and $C$.
+%
 \subsection{Bracket Abstraction}
@@
 \subsection{Graph Reduction}
-Must cover: outermost reduction.
+A good tutorial-style introduction to Graph Reduction may be found at~\cite{peyton_jones_implementing_2000}.
+For a more formal and detailed reference, see~\cite{spj_impl}.
+
+Graph Reduction is a common technique for implementing non-strict functional programming languages, i.e. languages that evaluate expressions only when their value is needed.
+Nodes in the graph are either constant values (leaf nodes) or application nodes (representing the application of a function to an argument).
+Edges are used to connect application nodes with their function and argument nodes.
+The usual convention is to denote application nodes with \emph{@}.
+Here is how we might represent the expression \texttt{square 3}:
+
+\begin{verbatim}
+       @
+      / \
+square   3
+\end{verbatim}
+
+This graph has three nodes: a reference to the function \texttt{square}, an argument 3, and an application node.
+We can \emph{reduce} this graph by applying rewrite rules to it. For example, we know that an expression \texttt{square x} produces a value $x*x$.
+For the above graph, we can therefore state that:
+
+\begin{verbatim}
+       @
+      / \    reduces to    9
+square   3
+\end{verbatim}
+
+The example graph happens to be a tree, but this is not always the case: A node may be referenced by multiple application nodes.
+Consider the graph for \texttt{square (square 3)}:
+
+\begin{verbatim}
+       @
+      / \
+     |   @
+     |  / \
+  square   3
+\end{verbatim}
+
+The two application nodes \emph{share} the \texttt{square} node.
+When we perform the reduction of \texttt{square 3}, the upper application node maintains its (now exclusive) reference to \texttt{square}, and the other application node is replaced with the constant value 9.
+Sharing nodes is important for efficiency because it prevents unnecessary re-evaluation of expressions.
+The graph representation of \texttt{(square 3) + (square 3)} highlights this:
+
+\begin{verbatim}
+      @ 
+     / \
+    @  |
+   / \ |
+  +   \|
+       @
+      / \
+square   3
+\end{verbatim}
+
+We would like to evaluate \texttt{square 3} only once.
+One way to do this is to replace the top application node of an evaluated expression with the new value.
+For the example graph above, we get:
+
+\begin{verbatim}
+      @ 
+     / \
+    @  |
+   / \ |
+  +   \|
+       9
+\end{verbatim}
+
+Now both the left and right sides of the + see the updated value (notice that we are passing the two arguments to + using a double application node).
+
+A graph could have multiple possible rewrites (reduction steps) that we could perform, in which case we need to decide what rewrite rules should be applied first.
+There are many possible ways to make this decision~\cite{spj_impl}, but most practical implementations choose a \emph{left-outermost} reduction strategy.
+In the graph representation, this amounts to traversing the graph from the root, following \emph{left} edges of application nodes.
+This chain of edges from the root down is called the \emph{spine} of the reduction.
+Eventually, we find a leaf node and apply the rewrite rule that matches that particular node (if any apply).
+The reduction then repeats from the root again, until eventually, no more rewrites are possible.
+
+At this point, it could be that the graph has been reduced to a constant value (e.g. 9).
+Another possibility is parts of the graph are still reduceable, but our reduction strategy does not `reach' it.
+For example, we can have the following graph (including a $K$ combinator) that our strategy will not reduce:
+
+\begin{verbatim}
+     @
+    / \
+   K   @
+      / \
+square   3
+\end{verbatim}
+
+$K$ requires two arguments, but we have given it only one, so the usual rewrite rule $K\ x\ y \Rightarrow x$ does not apply.
+Because the graph contains a subexpression that \emph{could} be reduced, we say the reduction has reached \emph{Weak Head Normal Form}, rather than \emph{Normal Form}.
 
@@
 \begin{figure*}
-    \includegraphics[width=2\columnwidth]{tigre_layout}
+    \includegraphics[width=1.4\columnwidth]{tigre_layout}
     \centering

### 2023-01-30 experiments/superg/doc/paper/paper.tex
Start bracket abstraction.

@@
 \subsection{Bracket Abstraction}
+Bracket Abstraction is the traditional algorithm for compiling lambda calculus expressions into combinatory logic.
+The basic algorithm requires only the $S$, $K$, and $I$ combinators.
+Its compilation function can be defined recursively as:
+
+\begin{equation*}
+    \begin{array}{l l l}
+        C [[\ e_1\ e_2\ ]]     & = & C [[\ e_1\ ]]\ C [[\ e_2\ ]]                  \\
+        C [[\ \lambda x.e\ ]]  & = & A\ x\ [[\ C [[\ e\ ]]\ ]]                     \\
+        C [[\ cv\ ]]           & = & cv                                            \\
+        \\
+        A\ x\ [[\ f_1\ f_2\ ]] & = & \ S\ (A\ x\ [[\ f_1\ ]])\ (A\ x\ [[\ f_2\ ]]) \\
+        A\ x\ [[\ x\ ]]        & = & I                                             \\
+        A\ x\ [[\ cv\ ]]       & = & K\ cv                                         \\
+    \end{array}
+\end{equation*}
+
+Any lambda calculus expression can be compiled using this algorithm.
+
+TODO more!
 

### 2023-01-31 experiments/superg/doc/paper/paper.tex
Finish up preliminaries.

@@
 For an accessible introduction to the Lambda Calculus, we can recommend~\cite{rojas_tutorial_2015}.
+
+The Lambda Calculus comes in many flavors.
+The most basic version has only function abstraction and application, and is untyped.
+Kiselyov's description of his new algorithms~\cite{kiselyov_lambda_2018} use a typed variant.
+Our description of the translation is not defined as a type system, hence the untyped Lambda Calculus suffices for our treatment.
+We do, however, enrich the Lambda Calculus with a few primitives that are convenient in a practical implementation. They are:
+\begin{itemize}
+    \item Machine integers (signed 32-bit). These can be emulated with e.g. Church Encoding, but arithmetic over a native integer type is faster.
+    \item Arithmetic operations. To manipulate machine integers, we introduce the usual $+$, $-$, $*$ and $/$ operations.
+    \item Comparison operations. To compare machine integers, we also add $==$, $<$ etc. These operations return $1$ if true, and $0$ if false (machine integers do double duty as a boolean type).
+\end{itemize}
+
 Most introductory resources, including~\cite{rojas_tutorial_2015}, present Lambda Calculus with named bindings.
 Kiselyov's algorithms~\cite{kiselyov_lambda_2018} use De Bruijn indices~\cite{de_bruijn_lambda_1972} instead.
+As this is core to the algorithm, we adopt the same approach.
+However, for example expressions, we continue to use the easier-to-read named bindings.
 
@@
 Chapter 16 of~\cite{spj_impl} starts with a good introduction to Combinatory Logic and the Bracket Abstraction algorithm.
+
+Combinatory Logic is a notation designed to remove variables (named or otherwise) from Lambda Calculus expressions.
+The core idea is to define a set of simple basic functions, or \emph{combinators}, that can be combined to pass variables to the proper subexpressions.
+We only need three combinators to convert any Lambda Calculus expression:
+\begin{itemize}
+    \item $S \ f \ g \ x = f \ x \ (g \ x)$: Passes an argument to both subexpressions of an application.
+    \item $K \ x \ y = x$: Drops the second argument.
+    \item $I \ x = x$\footnote{$SKK = I$, so even this combinator is not strictly necessary, but all reasonable implementations do include $I$.}: The identity function that accepts an argument and returns it.
+\end{itemize}
+
+The traditional algorithm to compile Lambda Calculus expressions into Combinator Logic with these three combinators is Bracket Abstraction, which we present in the next section.
+
+The set of combinators used in the translation may be extended to generate smaller programs.
+Kiselyov's algorithms~\cite{kiselyov_lambda_2018} for example use the following additional combinators as special cases of $S$:
+
+\begin{itemize}
+    \item $B\ f\ g\ x = f\ (g\ x)$: Passes an argument to the \emph{right} subexpression.
+    \item $C\ f\ g\ x = (f\ x)\ g$: Passes an argument to the \emph{left} subexpression.
+\end{itemize}
+
+Another addition, specifically for the $\linear$ algorithm, is the addition of \emph{bulk} combinators, variants of $S$, $B$ and $C$ that pass multiple arguments:
+
+\begin{itemize}
+    \item $S_n\ f\ g\ x_1\ ...\ x_n = (f\ x_1\ ...\ x_n)\ (g\ x_1\ ...\ x_n)$
+    \item $B_n\ f\ g\ x_1\ ...\ x_n = f\ (g\ x_1\ ...\ x_n)$
+    \item $C_n\ f\ g\ x_1\ ...\ x_n = (f\ x_1\ ...\ x_n)\ g$
+\end{itemize}
 
@@
 Any lambda calculus expression can be compiled using this algorithm.
-
-TODO more!
+As an example, we can compile the expression $\lambda x.\lambda y. x$, and obtain:
+
+\begin{equation*}
+    \begin{array}{l l}
+        C\ [[\ \lambda x.\lambda y. x\ ]]        & \Rightarrow \\
+        A\ x\ [[\ C\ [[\ \lambda y. x\ ]]\ ]]    & \Rightarrow \\
+        A\ x\ [[\ A\ y\ [[\ C\ [[\ x\ ]]\ ]]\ ]] & \Rightarrow \\
+        A\ x\ [[\ A\ y\ [[\ x\ ]]\ ]]            & \Rightarrow \\
+        A\ x\ [[\ K\ x\ ]]                       & \Rightarrow \\
+        S\ (A\ x\ [[\ K\ ]])\ (A\ x\ [[\ x\ ]])  & \Rightarrow \\
+        S\ (K\ K)\ I                             & \Rightarrow \\
+    \end{array}
+\end{equation*}
+
+The generated expression consists exclusively of combinators, as expected.
+We can show that it produces the same result as the original lambda expression by applying it to two abstract variables $x$ and $y$:
+
+\begin{equation*}
+    \begin{array}{l l}
+        S\ (K\ K)\ I\ x\ y   & \Rightarrow \\
+        (K\ K\ x)\ (I\ x)\ y & \Rightarrow \\
+        K\ x\ y              & \Rightarrow \\
+        x                    &             \\
+    \end{array}
+\end{equation*}
+
+While correct, the Bracket Abstraction algorithm is not optimal.
+Looking at the derivation above, $S\ (K\ K)\ I$ is identical to the $K$ combinator.
+In Section~\ref{sec:kiselyov}, we cover the better compilation algorithms proposed in~\cite{kiselyov_lambda_2018}.
 

### 2023-01-31 experiments/superg/doc/paper/paper.tex
Spelling.

@@
 The Lambda Calculus comes in many flavors.
-The most basic version has only function abstraction and application, and is untyped.
-Kiselyov's description of his new algorithms~\cite{kiselyov_lambda_2018} use a typed variant.
+The most basic version has only function abstraction and application and is untyped.
+Kiselyov's description of his new algorithms~\cite{kiselyov_lambda_2018} uses a typed variant.
 Our description of the translation is not defined as a type system, hence the untyped Lambda Calculus suffices for our treatment.

### 2023-02-01 experiments/superg/doc/paper/paper.tex
Final revision.

@@
 Kiselyov's algorithms~\cite{kiselyov_lambda_2018} use De Bruijn indices~\cite{de_bruijn_lambda_1972} instead.
-As this is core to the algorithm, we adopt the same approach.
-However, for example expressions, we continue to use the easier-to-read named bindings.
+As this is core to the algorithm, we adopt the same approach for our algorithm descriptions in Section~\ref{sec:kiselyov}.
+However, in the example expressions we give in this paper, we continue to use the easier-to-read named bindings.
 
@@
 Combinatory Logic is a notation designed to remove variables (named or otherwise) from Lambda Calculus expressions.
-The core idea is to define a set of simple basic functions, or \emph{combinators}, that can be combined to pass variables to the proper subexpressions.
+The core idea is to define a set of basic functions, or \emph{combinators}, that can be combined to pass variables to the proper subexpressions.
 We only need three combinators to convert any Lambda Calculus expression:
@@
         S\ (A\ x\ [[\ K\ ]])\ (A\ x\ [[\ x\ ]])  & \Rightarrow \\
-        S\ (K\ K)\ I                             & \Rightarrow \\
+        S\ (K\ K)\ I                             &             \\
     \end{array}
@@
 At this point, it could be that the graph has been reduced to a constant value (e.g. 9).
-Another possibility is parts of the graph are still reduceable, but our reduction strategy does not `reach' it.
+Another possibility is that parts of the graph are still reduceable, but our reduction strategy does not `reach' it.
 For example, we can have the following graph (including a $K$ combinator) that our strategy will not reduce:
@@
 The original Miranda was compiled using GCC 10.2.1 without any optimization flags, to avoid breaking the garbage collector as discussed in Section~\ref{sec:miranda}.
-All other code is written in Rust and compiled using the \texttt{release} profile (\texttt{rustc} version 1.65.0).
+All other code is written in Rust and compiled using the optimizing \texttt{release} profile (\texttt{rustc} version 1.65.0).
 
@@
 Our objective with this work was to revive research around Combinator Graph Reduction by improving the performance.
-We used a novel semantic translation algorithm to compile lambda calculus programs into smaller combinator expressions~\cite{kiselyov_lambda_2018} and have implemented modern combinator graph reduction engines based on the designs of Miranda~\cite{turner_new_1979} and TIGRE~\cite{koopman_fresh_1989}.
+We used a recently proposed semantic translation algorithm to compile lambda calculus programs into smaller combinator expressions~\cite{kiselyov_lambda_2018} and have implemented modern combinator graph reduction engines based on the designs of Miranda~\cite{turner_new_1979} and TIGRE~\cite{koopman_fresh_1989}.
 
